### 2018-02-05 report/paper.tex
Adding slides and report

\documentclass[sigplan,preprint,10pt]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\settopmatter{printacmref=false} 
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage[shortlabels]{enumerate}
\setlist{leftmargin=6mm}

\newcounter{thc}
\newcounter{dfc}

\theoremstyle{plain}
\newtheorem{lem}[thc]{Lemma}
\newtheorem{theorem}[thc]{Theorem}

\theoremstyle{definition}
\newtheorem{definition}[dfc]{Definition}

%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
%\acmYear{2017}
%\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}

\title{Wrattler: \textnormal{A platform for AI-assisted data science}}

\author{Authors}
%\affiliation{
%  \institution{The Alan Turing Institute}
%  \country{London, United Kingdom}
%}
%\email{tomas@tomasp.net}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}

\begin{document}
\maketitle

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can run a machine learning algorithm to provide advice 
based on statistical analysis of the data and discover hidden structures or corner cases, but only
a human can decide what those mean.
A data science environment of the future thus needs to support an interaction model that 
keeps the human in the loop, allows an efficient interaction between the human and the computer
and can be extended with new AI asistants that provide advice about data.

In this report, we present Wrattler, a notebook system that is interactive, reproducible, 
polyglot and smart.

\paragraph{Interactive.}
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state.
A data store tracks the provenance and semantics of data, supports versioning and keeps 
the history, making the data analyses fully reproducible.

\paragraph{Polyglot.}
The use of data store makes it possible to combine multiple programming languages in 
a single notebook. Data scientists can mix interactive languages for quick data exploration
that run in the web browser and provide live previews with robust systems such as R and Python.

\paragraph{Smart.}
Last, but not least, Wrattler serves as a platform for AI assistants that use novel
machine learning algorithms to provide suggestions about data. Such AI assistants connect to the
data store to infer types and meaning of data, provide help with data cleaning and joining, but also help data scientists
make sense of data by finding typical and atypical data points and automatically visualizing data.

\section{Wrattler and notebooks}

Notebook systems such as Jupyter became a popular programming environment for data science, because 
they support gradual data exploration and provide a convenient way of interleaving code with 
comments and visualizations. However, notebooks suffer from a number of issues that hanper 
reproducibility, make it difficult to combine multiple languages and limit the possible interaction
model.

Notebooks can be used in a way that breaks reproducibility. The state is maintained by a \emph{kernel} 
and running a code in a cell overwrites the current state. There is no record of how the current 
state was obtained and no way to rollback to a previous state. The fact that the state is 
maintained by the kernel means that it is hard to combine multiple programming languages and
other components such as AI assistants. Finally, notebooks provide a very limited interaction 
model. To see the effect of a code change, an entire cell and all subsequent cells need to be
manually reevalutated.

\begin{figure}
\includegraphics[scale=0.6]{diagram.pdf}
\caption{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are separated and enriched with AI assistants.}
\label{fig:arch}
\end{figure}

The architecture of Wrattler allows us to address these issues, as well as to provide a platform
for building novel AI assistants and interactive programming. The architecture is illustrated
in Figure~\ref{fig:arch}. The components of Wrattler are:
%
\begin{itemize}
\item[--] \textbf{Data store}. Imported external data, results of running scripts and of 
  applying AI assistants are stored in the data store. It stores versioned data frames with 
  other semantic information about data such as types, semantics, data format or provenance.
\vspace{0.25em}
\item[--] \textbf{Language runtimes}. Scripts are evaluated by one or more language runtimes.
  The runtimes read input data from and write results back to the data store.
\vspace{0.25em}
\item[--] \textbf{AI assistants}. When invoked from the notebook, AI assistants read data
  from data store and provide hints to the data analyst. They help to write data cleaning
  scripts or annotate data in the data store with additional metadata such as inferred types.
\vspace{0.25em}
\item[--] \textbf{Notebook}. The notebook is displayed in a web browser, which orchestrates 
  all other components. The browser builds a dependency graph between cells or individual 
  expressions in the cells. It calls language runtimes to evaluate code that has changed
  and reads data from the data store to display results.  
\end{itemize}

\section{Wrattler components}

b

\subsection{Notebook user interface}

\subsection{Dependency graph}

\subsection{Data store}

\subsection{Language runtimes}

\subsection{AI assistants}

\newpage
~
\newpage

platform for AI-assisted data science that  
Wrattler - an experimental notebook system that addresses the above issues. Wrattler is
 based on a simple idea of moving the state from a single language-specific stateful kernel
  to a datastore that can communicate with multiple workers that evaluate code or provide insights about data.

  I will discuss the far reaching consequences that the Wrattler architecture has. First, we can 
  provide live previews by recomputing only parts of notebooks. Second, our notebooks track 
  history and are fully reproducible. Third, it becomes possible to mix multiple languages and 
  also integrate AI assistants that communicate directly with the datastore to provide 
  suggestions about data extraction, cleaning and integration.

% context
% problem
% how we solve it
% what follows

Data science properties
What follows from this
Notebook issues
Wrattler architecture
What this enables

\end{document}

### 2018-02-08 report/paper.tex
Finish first report draft

@@
 Data science is an iterative, exploratory process that requires a collaboration between a 
-computer system and a human. A computer can run a machine learning algorithm to provide advice 
+computer system and a human. A computer can provide advice 
 based on statistical analysis of the data and discover hidden structures or corner cases, but only
@@
 A data science environment of the future thus needs to support an interaction model that 
-keeps the human in the loop, allows an efficient interaction between the human and the computer
+keeps the human in the loop, allows an efficient human-computer interaction 
 and can be extended with new AI asistants that provide advice about data.
-
-In this report, we present Wrattler, a notebook system that is interactive, reproducible, 
-polyglot and smart.
+In this report, we present Wrattler, a notebook system that is:
 
@@
 \paragraph{Polyglot.}
-The use of data store makes it possible to combine multiple programming languages in 
-a single notebook. Data scientists can mix interactive languages for quick data exploration
-that run in the web browser and provide live previews with robust systems such as R and Python.
+Multiple languages can be used in a single notebook and share data via the data store.
+Analysts can use R and Python, but also interactive languages for data exploration
+that run in the browser and provide live previews.
 
 \paragraph{Smart.}
-Last, but not least, Wrattler serves as a platform for AI assistants that use novel
-machine learning algorithms to provide suggestions about data. Such AI assistants connect to the
-data store to infer types and meaning of data, provide help with data cleaning and joining, but also help data scientists
-make sense of data by finding typical and atypical data points and automatically visualizing data.
+Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
+about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
+help with data cleaning and joining, but also help data exploration by finding typical and atypical 
+data points and automatically visualizing data.
 
@@
 comments and visualizations. However, notebooks suffer from a number of issues that hanper 
-reproducibility, make it difficult to combine multiple languages and limit the possible interaction
-model.
+reproducibility and limit the possible interaction model.
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.6]{diagram.pdf}
-\caption{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
-  Wrattler, those functions are separated and enriched with AI assistants.}
-\label{fig:arch}
-\end{figure}
-
 The architecture of Wrattler allows us to address these issues, as well as to provide a platform
@@
 %
-\begin{itemize}
-\item[--] \textbf{Data store}. Imported external data, results of running scripts and of 
+\paragraph{Data store.} Imported external data, results of running scripts and of 
 applying AI assistants are stored in the data store. It stores versioned data frames with 
-  other semantic information about data such as types, semantics, data format or provenance.
-\vspace{0.25em}
-\item[--] \textbf{Language runtimes}. Scripts are evaluated by one or more language runtimes.
+metadata such as types, inferred semantics, data formats or provenance.
+
+\paragraph{Language runtimes.} Scripts are evaluated by one or more language runtimes.
 The runtimes read input data from and write results back to the data store.
-\vspace{0.25em}
-\item[--] \textbf{AI assistants}. When invoked from the notebook, AI assistants read data
+
+\paragraph{AI assistants.} When invoked from the notebook, AI assistants read data
 from data store and provide hints to the data analyst. They help to write data cleaning
 scripts or annotate data in the data store with additional metadata such as inferred types.
-\vspace{0.25em}
-\item[--] \textbf{Notebook}. The notebook is displayed in a web browser, which orchestrates 
+
+\paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates 
 all other components. The browser builds a dependency graph between cells or individual 
-  expressions in the cells. It calls language runtimes to evaluate code that has changed
-  and reads data from the data store to display results.  
-\end{itemize}
+expressions in the cells. It calls language runtimes to evaluate code that has changed,
+AI assistants to provide hints and reads data from the data store to display results.
+
+\begin{figure}
+\vspace{-1em}
+\includegraphics[scale=0.6]{diagram.pdf}
+\vspace{-1em}
+\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
+  Wrattler, those functions are separated and enriched with AI assistants.}}
+\label{fig:arch}
+\vspace{-1em}
+\end{figure}
 
@@
 
-b
-
-\subsection{Notebook user interface}
+Wrattler consists of a notebook user interface running in the web browser, which communicates with 
+a number of server-side components including language runtimes, the data store and AI assistants. 
+In this section, we discuss the components in more detail, starting with the dependency graph
+that is maintained on the client-side, by the web browser.
 
@@
 
+When opening a notebook, Wrattler parses the source of the notebook (consisting of text cells and 
+code cells) and constructs a dependency graph, which serves as a runtime representation of the
+notebook.
+
+The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
+individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
+a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
+depend on a data frame (hexagons) exported by an earlier cell. 
+When code in a cell changes, Wrattler updates the dependency graph, keeping existing
+nodes for entities that were unaffected by the change.
+
+The dependency graph enables several features that are difficult for most notebook systems:
+%
+\begin{itemize}
+\item[--] When code changes, Wrattler only needs to recompute small part of the graph.
+  This makes it possible to provide live previews, especially for simple data analytical DSLs.
+\vspace{-0.85em}
+\item[--] Refactoring can extract parts of the notebook that
+  are needed to compute a specified output.
+\vspace{0.25em}
+\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
+\vspace{0.25em}
+\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
+  or code cleanup.
+\end{itemize}
+
 \subsection{Data store}
+
+The data store provides a way for persistently stroing data and enables communication between 
+individual Wrattler components. Data store keeps external data files imported into Wrattler,
+as well as computed data frames (and, potentially, other data structures). It is immutable 
+and keeps past versions of data to allow efficient state rollback. For persistency and versioning, 
+data store also serializes and stores multiple versions of the dependency graph.
+
+The data store supports a mechanism for annotating data frames with additional semantic information, such as:
+%
+\begin{itemize}
+\item[--] Columns can be annotated with (and stored as) primitive data type such as date or floating-point number.
+\vspace{0.25em}
+\item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or and address.
+\vspace{-0.85em}
+\item[--] Columns, rows and cells of the data frame can be annotated with other metadata such as provenance.
+\end{itemize}
 
@@
 
+Language runtimes are responsible for evaluating code and assisting with code analysis during
+the construction of dependency graph. They can run as server-side components (e.g.~for R and Python) 
+or as client-side components (for JavaScript and small analytical DSLs used by AI assistants). 
+Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
+the data store. (Using a cache for efficiency.)
+
+\begin{figure}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+\vspace{-0.5em}
+\caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
+opaque R/Python code cells. Data cleaning is written in a domain-specific langauge understood by
+Wrattler and represented by multiple cells. Latter two cells depend on data frames exported by
+earlier cells.}}
+\label{fig:deps}
+\vspace{-0.5em}
+\end{figure}
+
 \subsection{AI assistants}
 
-\newpage
-~
-\newpage
+The Wrattler architecture enables integration of AI assistants in a number of ways. First, AI
+assistants have access to the data store (and the dependency graph) and can use those to provide
+guidance about both data and code. Second, Wrattler includes an extensible framework for defining
+domain-specific languages (DSLs) that can describe, for example, data extraction, cleaning, transofrmation or 
+visualization. An AI assistant can work in two ways:
+%
+\begin{itemize}
+\item[--] When invoked, the assistant creates a new frame in the data store with additional
+  information, e.g. probabilistic type inference (ptype) will annotate columns (and possibly
+  also cells) with their probabilistic types.
+\vspace{0.25em}
+\item[--] The assistant defines a DSL for the task at hand. When invoked from a notebook,
+  it guides the user in creating a script in the DSL that performs the required operation
+  (such as extracting data or reformatting a data frame).
+\end{itemize}
+%
+The second case is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
+store and returns suggestions in a small DSL to the notebook, which the user can review 
+(using a live preview mechanism) and accept. For example, an AI
+assistant based on datadiff can suggest the following script (written using the Wrattler DSL
+framework):
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{datadiff(broadband2014, broadband2015)}\\
+\quad.\ident{drop\_column}(\str{WT\_national})\\
+\quad.\ident{drop\_column}(\str{WT\_ISP})\\
+\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
+\end{array}  
+\end{equation*}
+%
+The script specifies that two columns should be dropped from the badly structured data frame and
+one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
+\strf{Urban} and \strf{Rural}).
 
-platform for AI-assisted data science that  
-Wrattler - an experimental notebook system that addresses the above issues. Wrattler is
- based on a simple idea of moving the state from a single language-specific stateful kernel
-  to a datastore that can communicate with multiple workers that evaluate code or provide insights about data.
-
-  I will discuss the far reaching consequences that the Wrattler architecture has. First, we can 
-  provide live previews by recomputing only parts of notebooks. Second, our notebooks track 
-  history and are fully reproducible. Third, it becomes possible to mix multiple languages and 
-  also integrate AI assistants that communicate directly with the datastore to provide 
-  suggestions about data extraction, cleaning and integration.
-
-% context
-% problem
-% how we solve it
-% what follows
-
-Data science properties
-What follows from this
-Notebook issues
-Wrattler architecture
-What this enables
-
+The above script is constructed interactively. When the user types the first line and types
+``.'' the datadiff assistant offers the most likely patches and the user can choose. While
+choosing, a preview of the result is computed on-the-fly, giving the user an immediate feedback.
+Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
+Once it is interactively constructed, it can be translated to multiple supported languages
+such as R, Python or JavaScript.
 \end{document}

### 2018-02-20 report/paper.tex
Extend, copy paste AI assistant text, add Data Store thoughts

@@
 Data science is an iterative, exploratory process that requires a collaboration between a 
-computer system and a human. A computer can provide advice 
-based on statistical analysis of the data and discover hidden structures or corner cases, but only
-a human can decide what those mean.
-A data science environment of the future thus needs to support an interaction model that 
-keeps the human in the loop, allows an efficient human-computer interaction 
-and can be extended with new AI asistants that provide advice about data.
-In this report, we present Wrattler, a notebook system that is:
+computer system and a human. A computer can provide advice based on statistical analysis of the 
+data and discover hidden structures or corner cases, but only a human can decide what those mean
+and decide how to handle them in data science scripts.
+
+A data science environment of the future needs to support an interaction model that keeps the 
+human in the loop, allows an efficient human-computer interaction and can be extended with new 
+AI asistants that provide advice about data. Such advice needs to be explainable to the human user,
+as well as machine-readable by other components of the system. In this report, we present Wrattler, 
+a system that is:
 
@@
 Wrattler separates the task of running scripts from the task of managing state. 
-A data store tracks the provenance and semantics of data, supports versioning and keeps 
-the history, making the data analyses fully reproducible.
+This is handled by a data store, which tracks the provenance and semantics of data, supports 
+versioning and keeps the history, making the data analyses fully reproducible.
 
@@
 
+\paragraph{Explainable.}
+The hints provided by AI assistants are explainable. Rather than working as black boxes that 
+transform one dataset into another, AI assistants generate scripts in simple domain-specific
+languages, that specify how the data should be transformed. Those scripts can be reviewed 
+and modified by a human.
+
+\vspace{0.5em}
+\noindent
+In the rest of this document, we discuss limitations of current notebook systems and how Wrattler
+resolves them (Section~\ref{sec:overview}). We discuss how the individual components of Wrattler
+work together (Section~\ref{sec:wrattler}) and then focus on two of them in detail --
+we look at how Wrattler can be extended with AI assistants (Section~\ref{sec:ai}) and
+how semantic information about data is managed by the data store (Section~\ref{sec:datastore}).
+
 \section{Wrattler and notebooks}
+\label{sec:overview}
 
@@
 manually reevalutated.
+
+\begin{figure}
+\vspace{-0.5em}
+\includegraphics[scale=0.6]{diagram.pdf}
+\vspace{-0.5em}
+\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
+  Wrattler, those functions are separated and enriched with AI assistants.}}
+\label{fig:arch}
+\vspace{-0.5em}
+\end{figure}
 
@@
 
-\begin{figure}
-\vspace{-1em}
-\includegraphics[scale=0.6]{diagram.pdf}
-\vspace{-1em}
-\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
-  Wrattler, those functions are separated and enriched with AI assistants.}}
-\label{fig:arch}
-\vspace{-1em}
-\end{figure}
-
 \section{Wrattler components}
+\label{sec:wrattler}
 
@@
 \includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
-\vspace{-0.5em}
+
 \caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
@@
 \subsection{AI assistants}
+\label{sec:wrattler-ai}
 
@@
 such as R, Python or JavaScript.
+
+\section{AI assistants}
+\label{sec:ai}
+
+{\color{blue}
+An AI assistant guides data analysts through a single data collection, integration, preparation, 
+analytics, or reporting task. Typically analysts will interact with many separate assistants over 
+the course of an analysis, each of which specialize in a single kind of task.
+}
+\subsection{Architecture of AI assistants}
+
+{\color{blue}
+AI assistants have access to information in the Wrattler data store and can use two separate
+aspects for their guidance:
+%
+\begin{itemize}
+\item[--] AI assistants can access all data stored in the data
+store. This includes imported external data files, such as raw text or raw HTML content from which
+the analyst wants to extract data, as well as structured data frames imported from CSV files or
+produced as a result of previous computations. Moreover, an AI assistant can also access multiple
+versions of a file.
+\vspace{0.25em}
+
+\item[--] AI assistants can access the dependency graph created for a 
+notebook. This allows the them to access the transformation history, which is the code 
+that has already been run during the analysis. The idea is that assistants may need to know the 
+context of transformations that have already been performed in order to decide what to suggest next.
+\end{itemize}
+
+The assistants should be self-contained in the sense that they interact only through the 
+data store. This restriction is designed to encourage the Unix "do one thing well" design for 
+assistants.
+
+There are two important design aspects of AI assistants. First, they always produce code, rather
+than being a black-box that transforms data. Second, they are designed to be used interactively.
+
+\paragraph{Domain specific languages.}
+The output of an assistant is code in a small domain-specific language designed for the task that
+the AI assistant is solving. The code can be used in two ways -- during the interaction with the
+AI assistant, the code is exposed through a Wrattler DSL framework as discussed in 
+Section~\ref{sec:wrattler-ai}. For production use, the code is translated to a language like R or 
+Python. A typical code produced by an AI assistant will implement a data transformation or
+produce a visualization.
+
+\paragraph{Interactivity.}
+Assistants should usually contain user interaction in their design, and adapt rapidly to feedback 
+from the analyst. For example, a data cleaning assistant would probably want to ask the user to 
+confirm potential changes, and if the user says no, to reconfigure the cleaning module 
+significantly so that the user does not need to keep rejecting similar changes.
+
+However, at the same time, our data preparation tools should be usable in "batch mode" by analysts 
+who already know what they want to do and don't want to mess with a wizard. We should aim to 
+produce the equivalent of scikit-learn for data preparation.
+
+The output snippet should be readable, and something that a data analyst can look at to verify 
+that the transformation accomplishes what they had wanted. As far as possible, the snippet 
+should be "reusable" in a sense that it could be copy-pasted into another script that was 
+processing a similar dataset and would still work. You can think of a snippet as something 
+that a person might write in a single cell of a Jupyter notebook.
+}
+
+\subsection{Types of AI assistants}
+{\color{blue}
+AI assistants can be imagined for every step of the data analytics process. We attempt a fairly c
+omprehensive list of the data analytics tasks for which can imagine targetting assistants. Then, 
+for each task, we list a few examples of specific agents. Even though the list of agents is 
+incomplete, it is still too long to implement fully in the current project.
+
+\paragraph{Data collection.} Assistants can search the data store, as well as external repositories
+to recommend data sets that are relevant to the task at hand and offer to import these. For example,
+when analysing data about different countries, more information about countries can be offered.
+
+\paragraph{Data integration.} Assistants can identify pairs of variables across multiple data 
+sources that are likely to be coding the same variable.
+
+\paragraph{Schema inference.} Inferring types of variables and doing this probabilistically.
+
+\paragraph{Data parsing.}
+Quick tool for specifying a scraper for web pages (related to wrapper induction, Sumit Gulwani 
+work, possibly this paper Robust and Noise Resistant Wrapper Induction
+Assistant to automatically infer format of csv files
+Assistants to extract tables from pdf's, html
+Identify which values are likely to be "special", like missing data markers
+
+\paragraph{Data cleaning}
+Identifying values that seem to be out of range, e.g. missing value indicators
+Identifying possible inconsistent coding of strings
+Identifying potential functional dependencies (1:1 correspondences) and returning violations
+Manual cleaning of subset of data (a la Sample Clean)
+Indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns
+Identifying distinct records that may refer to the same entity (record linkage)
+
+\paragraph{Data monitoring.}
+Data diff: Returning a report of the differences between two data sets
+Monitoring classifier to detect shifts in its input distribution
+Detecting change points and covariate shift
+Montioring classifier to suggest when/how often new data points shoud be labelled for monitoring
+
+\paragraph{Exploratory visualization.}
+Summarizing most representative rows of a data table
+Automatically styling scatterplots and bar charts
+
+\paragraph{Performing analytics.}
+Which classification methods should I use? e.g., by search through similar problems on Kaggle and Open ML (a la AutoML)
+Simiarly for feature selection methods and so on
+Hyperparameter exploration
+
+\paragraph{Debugging and interpreting predictive analytics.}
+Debugging predictive analytics, e.g., deciding whether and how to add more features
+Error analysis assistant: Summarize the output on validation data
+Error analysis based on visualization of classifier score (see work of Saleema Amershi)
+Tools for explaining the predictions of a model on validation data (a la LIME) * Research directions: Improved model families for the interpretable explanation.
+Decision lists on marginal distribution rather than joint
+"Training set blame": Given an error in the validation data, what examples from the training set caused me to make that error?
+Explaining differences between predictions: why was instance 24601 treated differently than 24675?
+}
+\subsection{Possible AI assistants}
+
+{\color{blue}
+
+\paragraph{Data Parsing.}
+
+Simple scraper assistant: Simple information extraction wizard. Create a data table based
+on HTML pages that have been automatically generated, like product review pages,
+for which a simple pattern on the HTML tree will be sufficient to extract the desired information.
+This is called wrapper induction in the information extraction literature.
+
+* *Input*: Web page that contains information in unstructured form, such as a list of search resuts or an HTML list.
+
+* *Output*: A data frame (i.e. a relational table) that contains the information that has
+been extracted from the web page.
+
+* *User interaction*: User needs to initially highlight and specify which items to be scraped.
+Need to summarize the results of the scraping and highlight cases
+which might require correction from the user. Update scraper accordingly.
+
+* *Potential method*: A wrapper is a path through the HTML DOM. Use a language (like XPath) for navigating through the DOM. Search for the shortest path that retrieves
+the labels provided by the user.
+
+* *Research issues*: Not clear, given the large amount of related work in this area.
+We would need to read around carefully before we can identify opportunities.
+Possible opportunities include:
+    (a) Using a language like [Scrapy](https://scrapy.org/) to induce the wrappers;
+    (b) Using a language model over the scraped values to fine-tune the wrapper (e.g. if a wrapper returns both dates and cities, it's probably wrong), or
+	(c) using a probabilistic model over the program representation of the wrapper to focus attention in the search
+for good wrappers. Perhaps there are enough scrapers on line (e.g. Github projects that use scrapy) that you can get a training set for this.
+* *Related work*: Gulwani's work on data wrangling, lots of work on wrapper induction.
+There is a lot of work on this. Scrapy is a popular open source framework that several
+of my BSc/MSc students have found useful.
+
+\paragraph{Data cleaning.}
+*Inconsistent string assistant*: Identify potential typos in string data. The idea is to look for strings that do not occur commonly in the data set, but are very close in edit distance
+to strings that do occur commonly.
+
+* *Input*: A column of a data table whose values are strings, like addresses, cities, states/provinces, occupations, and so on.
+
+* *Output*: A list of suggestions for strings that should be renamed: e.g. `[ ("Stocland", "Scotland"), ("Untied Kringdom", "United Kingdom") ]`
+
+* *Potential method*: Simplest cut: Define a threshold $\delta$ on Levenshtein distance,
+and $0 < \alpha < 1$ on the number of occurrences. Return the set of all strings $y$ that
+occur less than $\alpha N$ times in the data set but where there is a corrected string $x$ which is less than $\delta$ in string edit distance from $y$.
+More ambitious: Noisy channel model for correction.
+Learn a probability distribution $p(x)$ over the strings in the column,
+and  define an error model $p(x'|x)$. Then search for ways to transform the  dirty column $x'$  to a clean version $x$ in such a way to maximize $p(x|x')$.
+This formulation deserves some more thought,
+though, before we try to implement it.
+
+* *Research questions*: Need to connect this to existing concepts in the databases and
+data mining literature, such as U-repair.
+
+* *User Interaction*: The user should be able to reject suggestions in the output list,
+and this should rapidly update the models used in such a way that similar suggestions
+to the bad ones are also removed from the list. A noisy channel model would be nice
+for this, not sure about the edit distance one.
+
+\paragraph{Exploratory visualization.}
+
+*Data table summarization assistant*: Displays a short subtable of the most "representative"
+columns of a given data frame. The number of rows in the summary should be small enough
+to fit on a given screen. Essentially an alternative to the `head()` function in pandas/R.
+
+*Input*:  A data frame, along with type information (continuous/categorical) for each of the columns. Desired number $k$ of rows in output.
+
+*Output*: A list of $k$ indices of rows in the dataframe
+
+*Potential method*:  Simplest possible implementation:
+Run $k$-medoids and return the medoid values.
+
+*Research questions:* First, are there research questions here, or is this so simple
+that it  is part of the infrastructure?  If $k$-medoids is so simple for this,
+why doesn't everyone also do it? How to evaluate whether the summary is good?
+Maybe there are examples of data sets where medoids is not good.
+For example, in the Karpathy ICML example, medoids would not make sense,
+listing the institutions that have the most papers is most informative, because
+it's a "long tail" type of column. How do you tell "long tail" data from "just show
+the clusters" type data? Perhaps a model-based framework could distinguish?
+Maybe you want to summarize the ways in which two data sets differ?
+Is the best summary the cluster centroids or the "top k" along some value? 
+* *User interaction:* I'm not sure what interactions are enecessary.
+Users could ask for more rows, mark two items as similar, or mark an item as uninteresting.
+Or perhaps better would be to allow drill down, i.e., to make it easy to explore
+the clusters represented by each example.
+}
+
+\section{Data store}
+\label{sec:datastore}
+
+{\color{red}
+\textbf{TP: My rough notes from discussion with JG}
+
+~
+
+\noindent
+Data store stores data frames with metadata. This is a good start for data science work,
+because most languages can provide mappings to/from data frames.
+}
+
+\subsection{Semantics annotations}
+{\color{red}
+Data store provides a way of annotating cells, columns, combinations of columns and rows.
+For example, a column may be annotated with a type that indicates that its values are a mix of
+postcodes and city names. A cell can be annotated with an information saying that this value is 
+more likely a postcode than a city name. We also need to support annotaitons on a combination of
+columns, because three columns can jointly represent an address. A row can be annotated with a 
+provenance -- for example, when merging rows from two data sets with different sources.
+}
+
+\subsection{Annotation format}
+{\color{red}
+We want a lightweight annotation format so that implementing an AI assistant or a language
+runtime that needs to communicate with the data store is not too hard.
+Our annotation format will follow something like \url{http://schema.org}, which defines
+a hierarchy of entity types and has a simple way for adding annotations to things.
+We will need to define our own hierarchy for schemas for data science.
+}
+
+\subsection{Schemas for data science}
+{\color{red}
+This hierarchy will include some basic things that the data store understands and can 
+convert between (integers, floats, unbounded size integers, ISO dates, etc.)
+
+In addition, we will define some purely semantic annotations that are commonly useful and 
+we will define a hierarchy for those. This represents things like addresses, postcodes, countries,
+cities, etc.
+
+Everyone can define their own schemas (like schema.org) by defining their own namespace (URL)
+and providing entities with their custom namespace. The data store will simply save and return
+this information - without doing any checking.}
+
+\subsection{Content negotiation}
+{\color{red}
+The only bit where the data store does something clever is that it will be able to return data
+in a format that the client asked for (a bit like HTTP negotiation). The client can say, ``I only
+understand semantic information defined by these schema namespaces`` and the data store will do
+its best to return data in the required format.
+
+For simple types (e.g. dates) that the data store understands, it will convert data accordingly.
+For other semantic information that the data store does not understand (e.g. information that
+a certain cell is missing with a probability $p$), the data store will filter out the extra
+information and just return the raw number.
+}
+
 \end{document}

### 2018-02-27 report/paper.tex
Clarify dependency graph, polish AI assistant intro

@@
 \paragraph{AI assistants.} When invoked from the notebook, AI assistants read data
-from data store and provide hints to the data analyst. They help to write data cleaning
-scripts or annotate data in the data store with additional metadata such as inferred types.
+from data store and provide hints to the analyst. They help to write data cleaning
+scripts or annotate data with additional metadata such as inferred types.
 
@@
 In this section, we discuss the components in more detail, starting with the dependency graph
-that is maintained on the client-side, by the web browser.
+that is maintained on the client-side, by the web browser. AI assistants are discussed later
+in Section~\ref{sec:ai}
 
@@
 
-The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
-individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
-a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
-depend on a data frame (hexagons) exported by an earlier cell. 
-When code in a cell changes, Wrattler updates the dependency graph, keeping existing
+As an example, the graph in Figure~\ref{fig:deps} represents the notebook in Figure~\ref{fig:notebook}.
+Top-level nodes (squares) represent individual notebook cells. Code (circles) is represented either as 
+a single node (Python, JavaScript) or as a sub-graph with node for each operation (cleaning DSL 
+understood by Wrattler). Any node of a cell can depend on a data frame (hexagons) exported by an 
+earlier cell. When code in a cell changes, Wrattler updates the dependency graph, keeping existing
 nodes for entities that were unaffected by the change.
@@
 \begin{itemize}
-\item[--] When code changes, Wrattler only needs to recompute small part of the graph.
+\item[--] When code changes, Wrattler only needs to recompute a small part of the graph.
   This makes it possible to provide live previews, especially for simple data analytical DSLs.
@@
 \subsection{Data store}
+\label{sec:wrattler-ds}
 
@@
 
+\noindent
+The format of the annotations as well as other features of the data store are discussed in 
+more detail in Section~\ref{sec:datastore}.
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\textit{\small{We retrieve data from a government web site using Python:}}\\[0.3em]
+\quad\ident{url}=\ident{open}(\str{http://data.gov.uk/\ldots/bb2014.csv})\\
+\quad\ident{bbRaw}=\ident{csv.reader}(\ident{url})\\[0.2em]
+\textit{\small{In the next block, we used an AI assistant to assist with cleaning}}\\[-0.1em]
+\textit{\small{and created a script that marks \#!NULL values as missing:}}\\[0.3em]
+\quad\kvd{let}~\ident{bbClean} = \ident{bbRaw}.\ident{missing}.\ident{as\_missing}(\str{\#!NULL})\\[0.3em]
+\textit{\small{Finally, we use JavaScript to create a histogram of upload speeds:}}\\[0.3em]
+\quad\ident{vega}.\ident{hsit}(\ident{bbClean}, \str{Upload speed})
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{\small{A small notebook consisting of three cells. The first Python cell exports 
+\ident{bbRaw}; the second Wrattler cell uses it and exports \ident{bbClean} and the 
+JavaScript cell visualizes \ident{bbClean}.}}
+\label{fig:notebook}
+\end{figure}
+\begin{figure}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+
+\caption{\small{Dependency graph for the notebook in Figure~\ref{fig:notebook}.
+Data acquisition is represented as a single Python cell. Data cleaning is understood
+by Wrattler and is the graph has a node for each call and constant. Hexagons represent
+data frames shared by cell and stored in the data store.}}
+\label{fig:deps}
+\vspace{-0.5em}
+\end{figure}
+
 \subsection{Language runtimes}
@@
 
-\begin{figure}
-\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+\section{AI assistants}
+\label{sec:ai}
 
-\caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
-opaque R/Python code cells. Data cleaning is written in a domain-specific langauge understood by
-Wrattler and represented by multiple cells. Latter two cells depend on data frames exported by
-earlier cells.}}
-\label{fig:deps}
-\vspace{-0.5em}
-\end{figure}
+An AI assistant guides data analysts through a single data collection, integration, preparation, 
+analytics, or reporting task. Typically analysts will interact with many separate assistants over 
+the course of an analysis, each of which specialize in a single kind of task.
 
-\subsection{AI assistants}
-\label{sec:wrattler-ai}
+\subsection{Architecture of AI assistants}
 
-The Wrattler architecture enables integration of AI assistants in a number of ways. First, AI
-assistants have access to the data store (and the dependency graph) and can use those to provide
-guidance about both data and code. Second, Wrattler includes an extensible framework for defining
-domain-specific languages (DSLs) that can describe, for example, data extraction, cleaning, transofrmation or 
-visualization. An AI assistant can work in two ways:
+The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
+assistants have access to the data store and can use it to provide guidance about both data and 
+code. Second, Wrattler includes an extensible framework for defining domain-specific languages 
+(DSLs) that can describe, for example, data extraction, cleaning, transofrmation or visualization.
+
+An assistant is self-contained and it interacts only through the data store. This encourages a 
+design where AI assistant ``does one thing well''.  The data store provides AI assistants with 
+two information that can be used for providing analyst with guidance:
 %
 \begin{itemize}
-\item[--] When invoked, the assistant creates a new frame in the data store with additional
-  information, e.g. probabilistic type inference (ptype) will annotate columns (and possibly
-  also cells) with their probabilistic types.
+\item[--] AI assistants can access all data stored in the data
+store. This includes imported external data files, such as raw text or raw HTML content from which
+the analyst wants to extract data, as well as structured data frames imported from CSV files or
+produced as a result of previous computations. Moreover, an AI assistant can also access multiple
+versions of a file.
 \vspace{0.25em}
-\item[--] The assistant defines a DSL for the task at hand. When invoked from a notebook,
-  it guides the user in creating a script in the DSL that performs the required operation
-  (such as extracting data or reformatting a data frame).
+
+\item[--] AI assistants can access the dependency graph created for a 
+notebook. This allows the them to access the transformation history, which is the code 
+that has already been run during the analysis. The gives assistants information about the
+context such as transformations that have already been performed.
 \end{itemize}
-%
-The second case is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
-store and returns suggestions in a small DSL to the notebook, which the user can review 
-(using a live preview mechanism) and accept. For example, an AI
-assistant based on datadiff can suggest the following script (written using the Wrattler DSL
-framework):
+
+\subsection{Design of AI assistants}
+There are two important design aspects of AI assistants. First, they always produce code, rather
+than being a black-box that transforms data. Second, they are designed to be used interactively.
+
+An AI assistant defines a domain specific language (DSL) for the task at hand. When invoked from 
+a notebook, it guides the user in creating a script in the DSL. Running the script performs
+the required operation (such as data cleaning). The assistant defines the semantics
+of those operations and, when the script is executed, stores the result into a data
+store. This new data frame can contain new data or additional information (such as their
+inferred probabilistic type).
+
+The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
+store and returns suggestions to the notebook, which the user can review and accept.
+
+~
+
+\paragraph{Domain specific languages.}
+
+Wrattler includes a scripting language with simple syntax that is used by individual AI assistants
+to define domain specific languages. The syntax is shared across multiple AI assistnts, but the
+operations are provided by individual assistants.
+
+For example, an AI assistant based on datadiff
+guides the user in writing scripts that transform a malformed data frame into a correct format
+specified by a correct data frame. The resulting script might look as follows:
 %
@@
 
-The above script is constructed interactively. When the user types the first line and types
-``.'' the datadiff assistant offers the most likely patches and the user can choose. While
-choosing, a preview of the result is computed on-the-fly, giving the user an immediate feedback.
-Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
-Once it is interactively constructed, it can be translated to multiple supported languages
-such as R, Python or JavaScript.
+The code produced by an AI assistant can be used in two ways. First, the analyst writes the code
+interactively during an interaction with the AI assistant. During this phase, Wrattler 
+keeps a fine-grained representation of the code in the dependency graph (second cell in 
+Figure~\ref{fig:deps}), which allows Wrattler to provie live previews of the result, giving the
+user an immediate feedback.
 
-\section{AI assistants}
-\label{sec:ai}
-
-{\color{blue}
-An AI assistant guides data analysts through a single data collection, integration, preparation, 
-analytics, or reporting task. Typically analysts will interact with many separate assistants over 
-the course of an analysis, each of which specialize in a single kind of task.
-}
-\subsection{Architecture of AI assistants}
-
-{\color{blue}
-AI assistants have access to information in the Wrattler data store and can use two separate
-aspects for their guidance:
-%
-\begin{itemize}
-\item[--] AI assistants can access all data stored in the data
-store. This includes imported external data files, such as raw text or raw HTML content from which
-the analyst wants to extract data, as well as structured data frames imported from CSV files or
-produced as a result of previous computations. Moreover, an AI assistant can also access multiple
-versions of a file.
-\vspace{0.25em}
-
-\item[--] AI assistants can access the dependency graph created for a 
-notebook. This allows the them to access the transformation history, which is the code 
-that has already been run during the analysis. The idea is that assistants may need to know the 
-context of transformations that have already been performed in order to decide what to suggest next.
-\end{itemize}
-
-The assistants should be self-contained in the sense that they interact only through the 
-data store. This restriction is designed to encourage the Unix "do one thing well" design for 
-assistants.
-
-There are two important design aspects of AI assistants. First, they always produce code, rather
-than being a black-box that transforms data. Second, they are designed to be used interactively.
-
-\paragraph{Domain specific languages.}
-The output of an assistant is code in a small domain-specific language designed for the task that
-the AI assistant is solving. The code can be used in two ways -- during the interaction with the
-AI assistant, the code is exposed through a Wrattler DSL framework as discussed in 
-Section~\ref{sec:wrattler-ai}. For production use, the code is translated to a language like R or 
-Python. A typical code produced by an AI assistant will implement a data transformation or
-produce a visualization.
+At a later stage, when the code is used in production, the code can be translated to a language
+like R or Python. This translation uses the fine-grained representation of the code in the 
+dependency graph and maps individual nodes to corresponding Python or R functions.
 
 \paragraph{Interactivity.}
+
 Assistants should usually contain user interaction in their design, and adapt rapidly to feedback 
@@
 
-However, at the same time, our data preparation tools should be usable in "batch mode" by analysts 
-who already know what they want to do and don't want to mess with a wizard. We should aim to 
-produce the equivalent of scikit-learn for data preparation.
+In the above example, the interaction is provied through the ``.'' operator of the scripting
+language. When the analyst types the first line to invoke the AI assistant and types ``.''
+the datadiff assistant offers the most likely patches and the user can choose which ones
+to accept. Scripts constructed in such way are readable, allowin the data analyst to look at them
+and verify that the transformations acomplish the intended task (such verification further
+benefits from live previews that Wrattler provides when the analyst navigates through the
+individual transformations).
 
-The output snippet should be readable, and something that a data analyst can look at to verify 
-that the transformation accomplishes what they had wanted. As far as possible, the snippet 
-should be "reusable" in a sense that it could be copy-pasted into another script that was 
-processing a similar dataset and would still work. You can think of a snippet as something 
-that a person might write in a single cell of a Jupyter notebook.
-}
+At the same time, AI assistants can be used in a fully automatic ``batch mode'' by analysts 
+who already know what they want to achieve and know that the assistant produces the result they
+require. In this case, Wrattler will simply automatically accept the default recommendations made
+by an AI assistant. This allows packaging the assistant as an equivalent of scikit-learn for 
+data preparation.
 
@@
 
+\newpage
+~
+\newpage
+
 \section{Data store}
@@
 
-{\color{red}
-\textbf{TP: My rough notes from discussion with JG}
+The data store links together individual components of Wrattler. It stores the raw input 
+data for data analysis such as downloaded web pages, raw data frames and also data frames
+produced as the result of running a script or invoking an AI assistant. In addition to data,
+the data store also stores multiple versions of the dependency graph created by Wrattler.
 
-~
+In the initial version of Wrattler, we restrict the data storage to raw input files and
+data frames, however we note that it would be desirable to include other data structures
+such graph or network data.
+
+In this section, we focus on the initial functionality that is required from the data store.
+We focus on the public interface exposed by the data store, but We do not discuss the technolgy 
+used for data storage -- this can be decided during the development.
+
+Data frames stored in the data store can be annotated with metadata. Such metadata can 
+communicate information between multiple AI assistants. For example, first AI assistant can
+infer that a type of a pair of columns represents GPS coordinate and produce a data frame
+with annotations specifying this information. A second AI assistant can then use this information
+and automatically provide a data visualization in the form of map. In the rest of this section,
+we discuss the structure of those annotations.
+
+\subsection{Semantic annotations}
+As noted in Section~\ref{sec:wrattler-ds}, annotations can be attached to individual cells, 
+entire columns and rows, but also combinations of multiple columns or rows. As an example,
+consider the following small table that records data on broadband speed (and illustrates a number
+of real-world problems):
+
+\vspace{1em}
+\noindent
+\begin{tabular}{llc}
+\toprule
+\textbf{ISP}\qquad\qquad\qquad & \textbf{Postcode}\qquad\qquad & \textbf{Download} \\
+\midrule
+Virgin & CB43BE & 46.34 \\
+BT & CB43PT & N/A \\
+Virgin & NW53HL & 41.2 \\
+BT & Tottenham & 16.4 \\
+\bottomrule
+\end{tabular}
+\vspace{1em}
 
 \noindent
-Data store stores data frames with metadata. This is a good start for data science work,
-because most languages can provide mappings to/from data frames.
-}
+If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
+it will annotate the table with a number of information:
+\begin{itemize}
+\item[--] aa
+\end{itemize}
 
-\subsection{Semantics annotations}
 {\color{red}

### 2018-02-28 report/paper.tex
assistants section; add lots to intro

--- 2018-02-20 report/paper.tex
@@
 data and discover hidden structures or corner cases, but only a human can decide what those mean
-and decide how to handle them in data science scripts.
+and decide how to handle them in data science scripts. Data science is often cited as an expensive
+and time consuming task, especially due the costs of data cleaning and data wrangling.
+We propose four fundamental reasons why
+practical data science is expensive:
 
-A data science environment of the future needs to support an interaction model that keeps the 
-human in the loop, allows an efficient human-computer interaction and can be extended with new 
-AI asistants that provide advice about data. Such advice needs to be explainable to the human user,
-as well as machine-readable by other components of the system. In this report, we present Wrattler, 
-a system that is:
+\paragraph{Big data is big,} 
+so the analyst doesn't understand it all.*
+Even if a data set is small enough fits on one computer, it's still too large to fit in 
+an analyst's working memory.
+But this means every analysis is huanted
+by the spectre of lurking, potentially
+unknown data quality  issues.
+This also makes it more difficult to do data fusion, because there may be corner cases that make it more difficult to join two disparate data sources than expected.
 
-\paragraph{Interactive.}
+\paragraph{The double Anna Karenina principle.}
+Not only is every dirty data set  dirty in its own way, but \emph{pace} Tolstoy, every clean data set is clean in its own way as well. 
+"Data" is such an abstract concept that specific integrity conditions to characterize whether data is dirty, and potentially even the most appropriate formalism for integrity conditions,  differ dramatically across
+the vast array of disparate use cases of data science,
+ranging form relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.
+
+\paragraph{Death by a thousand cuts.} 
+Often data transformation and processing steps are individually  very simple. But  there may need to be lots of them, and
+because big data is big,
+an analyst never knows if she has found them all.
+
+\paragraph{Feedback cycles everywhere.} Data science is 
+not a pipeline but a connected mess of epicycles. This is because every step 
+in a data analysis actually teaches 
+the analyst more about the data and the problem, which might require rehtinking the earlier steps. 
+For example, there might be data quality issues that 
+are not uncovered until the analyst
+investigates the output of a regression model.
+
+\vspace{1em}
+To meet these challenges, we present Wrattler,
+a new type of system for data science
+that aims to transform the process
+of data analysis.
+Wrattler combines the interactive and literal programming paradigms
+of notebook systems such as Jupyter
+with new advances in AI systems for data wrangling and in provenance.
+The main design principles in Wrattler are:
+
+\paragraph{Interactive.} Interactivity
+is a necessary because ``big data is big'', so the analyst learns about 
+the data set and the problem as she explores it.
 Wrattler enables an efficient interaction by bringing computation closer to the human.
@@
 \paragraph{Reproducible.} 
+Data analyses must be reproducible
+because of feedback cycles. As the
+analyst learns more about the problem,
+this may uncover data cleaning or preparation issues that require
+redesigning and rerunning the analysis.
 Wrattler separates the task of running scripts from the task of managing state. 
@@
 \paragraph{Polyglot.}
+Modern data science naturally draws
+on many competing languages and libraries, such as R and Scipy. As a side effect
+of our interactive, reproducible design,
+we obtain nearly for free the ability
+to support polyglot data analyses.
 Multiple languages can be used in a single notebook and share data via the data store.
@@
 \paragraph{Smart.}
+AI can examine and find patterns
+in big data where a human cannot,
+again aiming at the problem that
+big data is big.
+AI can be used to direct the analyst's attention and to generalize  decisions about data transformation
+to new data that that the analyst
+hasn't seen.
 Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
@@
 
-{\color{blue}
-An AI assistant guides data analysts through a single data collection, integration, preparation, 
-analytics, or reporting task. Typically analysts will interact with many separate assistants over 
+{
+An AI assistant is a component
+of the system that 
+guides data analysts through a single data collection, integration, preparation, 
+analytics, or reporting task.
+For example, an assistant might
+specialize in identifying suspicious
+values, such as "-1" in a column labelled
+"heart rate", which might indicate
+that certain data was not collected.
+Assistants are interactive, in that
+they propose a transformation
+of the data to the analyst, based on combinations
+of statistical and symbolic AI analysis,
+and then refine the transformation
+based on feedback from the analyst.
+Typically analysts will interact with many separate assistants over 
 the course of an analysis, each of which specialize in a single kind of task.
@@
 
-{\color{blue}
-AI assistants have access to information in the Wrattler data store and can use two separate
-aspects for their guidance:
-%
-\begin{itemize}
-\item[--] AI assistants can access all data stored in the data
-store. This includes imported external data files, such as raw text or raw HTML content from which
-the analyst wants to extract data, as well as structured data frames imported from CSV files or
-produced as a result of previous computations. Moreover, an AI assistant can also access multiple
-versions of a file.
-\vspace{0.25em}
+There are two important design aspects of AI assistants, which may seem 
+contradictory.
+ First, they must be interactive, following
+ the first principle behind Wrattler, but
+ they must also be reproducible.
+We square this circle by using the notebook
+system as a reproducible interlingua. 
+The output of an AI assistant is always code, rather
+than being a black-box that transforms data,
+and it is this code that is agreed between the AI and the analyst during the interaction.
+Thus the interaction is viewed as a ``design phase''
+that is not reproduced on new data.
 
-\item[--] AI assistants can access the dependency graph created for a 
-notebook. This allows the them to access the transformation history, which is the code 
-that has already been run during the analysis. The idea is that assistants may need to know the 
-context of transformations that have already been performed in order to decide what to suggest next.
-\end{itemize}
-
-The assistants should be self-contained in the sense that they interact only through the 
-data store. This restriction is designed to encourage the Unix "do one thing well" design for 
-assistants.
-
-There are two important design aspects of AI assistants. First, they always produce code, rather
-than being a black-box that transforms data. Second, they are designed to be used interactively.
+Assistants are therefore based on two different ideas:
 
@@
 that a person might write in a single cell of a Jupyter notebook.
-}
 
-\subsection{Types of AI assistants}
-{\color{blue}
-AI assistants can be imagined for every step of the data analytics process. We attempt a fairly c
-omprehensive list of the data analytics tasks for which can imagine targetting assistants. Then, 
-for each task, we list a few examples of specific agents. Even though the list of agents is 
-incomplete, it is still too long to implement fully in the current project.
+\vspace{1em}
+AI assistants have access to information in the Wrattler data store and can use two separate
+aspects for their guidance:
+%
+\begin{itemize}
+\item[--] AI assistants can access all data stored in the data
+store. This includes imported external data files, such as raw text or raw HTML content from which
+the analyst wants to extract data, as well as structured data frames imported from CSV files or
+produced as a result of previous computations. Moreover, an AI assistant can also access multiple
+versions of a file.
+\vspace{0.25em}
+
+\item[--] AI assistants can access the dependency graph created for a 
+notebook. This allows the them to access the transformation history, which is the code 
+that has already been run during the analysis. The idea is that assistants may need to know the 
+context of transformations that have already been performed in order to decide what to suggest next.
+\end{itemize}
+
+The assistants should be self-contained in the sense that they interact with each other
+only through the 
+data store. This restriction is necessary in order
+for provenance tracking and reproducibility.
+
+\subsection{The Assistant Landscape}
+{
+
+AI assistants can be imagined for every step of the data analytics process. 
+To give an idea of the breadth of this framework,
+we present a broad list of the data analytics tasks which have potential for assistant automation.
+This is an intentionally broad list, more so than
+could be fully explored in even a large research project, but gives a sense of the breadth of 
+potential use cases of our framework.
 
@@
 to recommend data sets that are relevant to the task at hand and offer to import these. For example,
-when analysing data about different countries, more information about countries can be offered.
+when analysing data about different countries, 
+the assistant could suggest joining in public
+demographic information from DBPedia.
 
-\paragraph{Data integration.} Assistants can identify pairs of variables across multiple data 
-sources that are likely to be coding the same variable.
+\paragraph{Data integration} is a problem as common
+as it is difficult. We envisage assistants
+that combine logical and statistical reasoning to assist
+with data  integration. A simple but useful
+example of a tool in this space is to
+ identifying pairs of variables across multiple data 
+sources that are likely to be daily temperatures.
+More sophisticated examples could include and extend the
+most cutting edge techniques in the data fusion and
+semantic technologies literatures.
 
-\paragraph{Schema inference.} Inferring types of variables and doing this probabilistically.
+\paragraph{Schema inference.} A sad fact of data
+science is that data sets are often not as well documented
+as we might like, even to the minimal level of
+documentation of the types of each column.
+There is an interesting opportunity to
+develop machine learning methods that inferring types of variables from their values, e.g.,
+by noticing that integer values between 0 and 30 in 
+the United Kingdom might describe an outdoor temperature, especially if other weather-related
+variables are detected in the data set.
 
 \paragraph{Data parsing.}
-Quick tool for specifying a scraper for web pages (related to wrapper induction, Sumit Gulwani 
-work, possibly this paper Robust and Noise Resistant Wrapper Induction
-Assistant to automatically infer format of csv files
-Assistants to extract tables from pdf's, html
-Identify which values are likely to be "special", like missing data markers
+A common task in data analysis is to convert
+data that is ``almost'' in structured form,
+such as HTML tables, into a relational table.
+Assistants in this space include
+tools specifying a scraper for web pages, related
+to wrapper induction \cite{Kushmerick1997WrapperIF},
+for inferring format of CSV files (while simple
+conceptually, they have no standard format), and
+for extracting tables from html and pdf files
+\cite{pinto03table}.
 
-\paragraph{Data cleaning}
-Identifying values that seem to be out of range, e.g. missing value indicators
-Identifying possible inconsistent coding of strings
-Identifying potential functional dependencies (1:1 correspondences) and returning violations
-Manual cleaning of subset of data (a la Sample Clean)
-Indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns
-Identifying distinct records that may refer to the same entity (record linkage)
+\paragraph{Data cleaning} is a well studied
+problem in the databases and data mining literature
+\cite{abedjan2016detecting,ilyas2015},
+but there are still a huge number of issues that have been unexplored.
+For example, we envisage assistants for
+Identifying values that seem to be out of range, e.g. missing value indicators;
+Identifying possible inconsistent coding of strings,
+such as different representations of acrynoms;
+indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
+identifying distinct records that may refer to the same entity (record linkage).
 
-\paragraph{Data monitoring.}
-Data diff: Returning a report of the differences between two data sets
-Monitoring classifier to detect shifts in its input distribution
-Detecting change points and covariate shift
-Montioring classifier to suggest when/how often new data points shoud be labelled for monitoring
+\paragraph{Data monitoring.} Once an analysis
+is completed, it may result in a machine learning
+model which will then be deployed on a data stream.
+Once a classifier is deployed, it must be monitored 
+to detect shifts in its input distribution,
+such as change points, covariate shift, and
+concept drift. This includes the problem 
+to suggesting when/how often new data points should be labelled for monitoring.
+Even for data analyses that do not involve
+deploying a ``production model'', there is
+often a need to run an analysis periodically on updated data, such as re-running last quarter's
+analysis on this quarter's data.
+In such cases, we might believe that the two data
+sets come from approximately the same distribution
+and format, so we can exploit
+divergences from this assumption to indicate
+potential issues with data quality.
+As part of the Wrattler effort, we have
+introduced the \emph{data diff} problem \cite{datadiff},
+which aims to improve the process of data wrangling by 
+returning a report of the differences between two data sets.
 
-\paragraph{Exploratory visualization.}
-Summarizing most representative rows of a data table
-Automatically styling scatterplots and bar charts
+\paragraph{Exploratory visualization} to summarize
+data sets, both the target data of the analysis,
+and the results of transformations that the analyst
+creates. Potential assistants here include ones
+to summarize most representative rows of a data table,
+and automatically choosing appropriate interactive
+visualizations, such as scatterplots and bar charts,
+with styling decisions such as axes automatically
+chosen to be most informative.
 
 \paragraph{Performing analytics.}
-Which classification methods should I use? e.g., by search through similar problems on Kaggle and Open ML (a la AutoML)
-Simiarly for feature selection methods and so on
-Hyperparameter exploration
+An increasing amount of attention is being
+given to the AutoML challenge \cite{guyon_review_2016}, which focuses
+on automatically choosing a classification
+methods, feature representations and hyperparameters
+for a given classification task, without the need for human intervention.
 
 \paragraph{Debugging and interpreting predictive analytics.}
-Debugging predictive analytics, e.g., deciding whether and how to add more features
-Error analysis assistant: Summarize the output on validation data
-Error analysis based on visualization of classifier score (see work of Saleema Amershi)
-Tools for explaining the predictions of a model on validation data (a la LIME) * Research directions: Improved model families for the interpretable explanation.
-Decision lists on marginal distribution rather than joint
-"Training set blame": Given an error in the validation data, what examples from the training set caused me to make that error?
-Explaining differences between predictions: why was instance 24601 treated differently than 24675?
+Every time a new machine learning model is trained,
+the first question an analyst has is often
+why its accuracy is not better. But
+debugging predictive analytics is notoriously difficult, and the remedies indirect and expensive,
+ e.g., labelling more data, adding more features,
+ using a more complex model. Debugging and improving
+ models is a rich area for potential assistants.
+ There is potentially low hanging fruit available
+ in how to aid the tasks of
+error analysis, exploring the
+predictions of the model 
+on validation data. Some early work in this direction is
+\cite{saleema}.
+There is a rich and growing literature on
+interpreting and explaining the predictions
+of a model \cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.
+Two potential directions from data are
+debugging by using
+"training set blame": Given an error in the validation data, what examples from the training set caused me to make that error? This might build on the work of \cite{percy}.
+Another potential avenue is explaining
+differences between predictions: why was instance 42 treated differently than instance 24601, even though they appear highly similar to a human?
 }
+
+\vspace*{1em}
+
+%%cs possibly we don't need to go into detail
+%% in this version of the report? Or we might
+%% decide to add this back.
+\begin{comment}
 \subsection{Possible AI assistants}
+
+Designing an AI assistant requires an intriguing
+combination of data science, AI, and user interaction
+thinking.
+As an example of the types of design considerations
+that arise, we go into more detail about the
+design issues that arise when considering
+three examples of the broader list of agents
+in the previous section.
 
@@
 }
+\end{comment}
 
@@
 
+\bibliographystyle{plain}
+\bibliography{paper}
 \end{document}

### 2018-02-28 report/paper.tex
fix snafus added in merge

@@
 
-The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
-individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
-a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
-depend on a data frame (hexagons) exported by an earlier cell. 
-When code in a cell changes, Wrattler updates the dependency graph, keeping existing
+As an example, the graph in Figure~\ref{fig:deps} represents the notebook in Figure~\ref{fig:notebook}.
+Top-level nodes (squares) represent individual notebook cells. Code (circles) is represented either as
+a single node (Python, JavaScript) or as a sub-graph with node for each operation (cleaning DSL
+understood by Wrattler). Any node of a cell can depend on a data frame (hexagons) exported by an
+earlier cell. When code in a cell changes, Wrattler updates the dependency graph, keeping existing
 nodes for entities that were unaffected by the change.
@@
 
+\noindent
+The format of the annotations as well as other features of the data store are discussed in
+more detail in Section~\ref{sec:datastore}.
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{l}
+\textit{\small{We retrieve data from a government web site using Python:}}\\[0.3em]
+\quad\ident{url}=\ident{open}(\str{http://data.gov.uk/\ldots/bb2014.csv})\\
+\quad\ident{bbRaw}=\ident{csv.reader}(\ident{url})\\[0.2em]
+\textit{\small{In the next block, we used an AI assistant to assist with cleaning}}\\[-0.1em]
+\textit{\small{and created a script that marks \#!NULL values as missing:}}\\[0.3em]
+\quad\kvd{let}~\ident{bbClean} = \ident{bbRaw}.\ident{missing}.\ident{as\_missing}(\str{\#!NULL})\\[0.3em]
+\textit{\small{Finally, we use JavaScript to create a histogram of upload speeds:}}\\[0.3em]
+\quad\ident{vega}.\ident{hsit}(\ident{bbClean}, \str{Upload speed})
+\end{array}
+\end{equation*}
+\vspace{-0.5em}
+\caption{\small{A small notebook consisting of three cells. The first Python cell exports
+\ident{bbRaw}; the second Wrattler cell uses it and exports \ident{bbClean} and the
+JavaScript cell visualizes \ident{bbClean}.}}
+\label{fig:notebook}
+\end{figure}
+\begin{figure}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+
+\caption{\small{Dependency graph for the notebook in Figure~\ref{fig:notebook}.
+Data acquisition is represented as a single Python cell. Data cleaning is understood
+by Wrattler and is the graph has a node for each call and constant. Hexagons represent
+data frames shared by cell and stored in the data store.}}
+\label{fig:deps}
+\vspace{-0.5em}
+\end{figure}
+
 \subsection{Language runtimes}
@@
 the data store. (Using a cache for efficiency.)
-
-\begin{figure}
-\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
-
-\caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
-opaque R/Python code cells. Data cleaning is written in a domain-specific langauge understood by
-Wrattler and represented by multiple cells. Latter two cells depend on data frames exported by
-earlier cells.}}
-\label{fig:deps}
-\vspace{-0.5em}
-\end{figure}
-
-\subsection{AI assistants}
-\label{sec:wrattler-ai}
-
-The Wrattler architecture enables integration of AI assistants in a number of ways. First, AI
-assistants have access to the data store (and the dependency graph) and can use those to provide
-guidance about both data and code. Second, Wrattler includes an extensible framework for defining
-domain-specific languages (DSLs) that can describe, for example, data extraction, cleaning, transofrmation or 
-visualization. An AI assistant can work in two ways:
-%
-\begin{itemize}
-\item[--] When invoked, the assistant creates a new frame in the data store with additional
-  information, e.g. probabilistic type inference (ptype) will annotate columns (and possibly
-  also cells) with their probabilistic types.
-\vspace{0.25em}
-\item[--] The assistant defines a DSL for the task at hand. When invoked from a notebook,
-  it guides the user in creating a script in the DSL that performs the required operation
-  (such as extracting data or reformatting a data frame).
-\end{itemize}
-%
-The second case is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
-store and returns suggestions in a small DSL to the notebook, which the user can review 
-(using a live preview mechanism) and accept. For example, an AI
-assistant based on datadiff can suggest the following script (written using the Wrattler DSL
-framework):
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{datadiff(broadband2014, broadband2015)}\\
-\quad.\ident{drop\_column}(\str{WT\_national})\\
-\quad.\ident{drop\_column}(\str{WT\_ISP})\\
-\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
-\end{array}  
-\end{equation*}
-%
-The script specifies that two columns should be dropped from the badly structured data frame and
-one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
-\strf{Urban} and \strf{Rural}).
-
-The above script is constructed interactively. When the user types the first line and types
-``.'' the datadiff assistant offers the most likely patches and the user can choose. While
-choosing, a preview of the result is computed on-the-fly, giving the user an immediate feedback.
-Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
-Once it is interactively constructed, it can be translated to multiple supported languages
-such as R, Python or JavaScript.
 
@@
 the course of an analysis, each of which specialize in a single kind of task.
-}
+
 \subsection{Architecture of AI assistants}
@@
  the first principle behind Wrattler, but
+ following the second,
  they must also be reproducible.
@@
 
+More specifically, an AI assistant defines a domain specific language (DSL) for the task at hand. When invoked from
+a notebook, it guides the user in creating a script in the DSL. Running the script performs
+the required operation (such as data cleaning). The assistant defines the semantics
+of those operations and, when the script is executed, stores the result into a data
+store. This new data frame can contain new data or additional information (such as their
+inferred probabilistic type).
+The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
+store and returns suggestions to the notebook, which the user can review and accept.
+
 Assistants are therefore based on two different ideas:
@@
 \paragraph{Domain specific languages.}
-The output of an assistant is code in a small domain-specific language designed for the task that
-the AI assistant is solving. The code can be used in two ways -- during the interaction with the
-AI assistant, the code is exposed through a Wrattler DSL framework as discussed in 
-Section~\ref{sec:wrattler-ai}. For production use, the code is translated to a language like R or 
-Python. A typical code produced by an AI assistant will implement a data transformation or
-produce a visualization.
+
+Wrattler includes a scripting language with simple syntax that is used by individual AI assistants
+to define domain specific languages. The syntax is shared across multiple AI assistnts, but the
+operations are provided by individual assistants.
+
+For example, an AI assistant based on datadiff
+guides the user in writing scripts that transform a malformed data frame into a correct format
+specified by a correct data frame. The resulting script might look as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{datadiff(broadband2014, broadband2015)}\\
+\quad.\ident{drop\_column}(\str{WT\_national})\\
+\quad.\ident{drop\_column}(\str{WT\_ISP})\\
+\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
+\end{array}
+\end{equation*}
+%
+The script specifies that two columns should be dropped from the badly structured data frame and
+one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
+\strf{Urban} and \strf{Rural}).
+
+The code produced by an AI assistant can be used in two ways. First, the analyst writes the code
+interactively during an interaction with the AI assistant. During this phase, Wrattler
+keeps a fine-grained representation of the code in the dependency graph (second cell in
+Figure~\ref{fig:deps}), which allows Wrattler to provie live previews of the result, giving the
+user an immediate feedback.
+
+At a later stage, when the code is used in production, the code can be translated to a language
+like R or Python. This translation uses the fine-grained representation of the code in the
+dependency graph and maps individual nodes to corresponding Python or R functions.
 
 \paragraph{Interactivity.}
+
 Assistants should usually contain user interaction in their design, and adapt rapidly to feedback
@@
 
-However, at the same time, our data preparation tools should be usable in "batch mode" by analysts 
-who already know what they want to do and don't want to mess with a wizard. We should aim to 
-produce the equivalent of scikit-learn for data preparation.
+In the above example, the interaction is provided through the ``.'' operator of the scripting
+language. When the analyst types the first line to invoke the AI assistant and types ``.''
+the datadiff assistant offers the most likely patches and the user can choose which ones
+to accept. Scripts constructed in such way are readable, allowing the data analyst to look at them
+and verify that the transformations accomplish the intended task (such verification further
+benefits from live previews that Wrattler provides when the analyst navigates through the
+individual transformations).
 
-The output snippet should be readable, and something that a data analyst can look at to verify 
-that the transformation accomplishes what they had wanted. As far as possible, the snippet 
-should be "reusable" in a sense that it could be copy-pasted into another script that was 
-processing a similar dataset and would still work. You can think of a snippet as something 
-that a person might write in a single cell of a Jupyter notebook.
+At the same time, AI assistants can be used in a fully automatic ``batch mode'' by analysts
+who already know what they want to achieve and know that the assistant produces the result they
+require. In this case, Wrattler will simply automatically accept the default recommendations made
+by an AI assistant. This allows packaging the assistant as an equivalent of scikit-learn for
+data preparation.
 
 \vspace{1em}
-AI assistants have access to information in the Wrattler data store and can use two separate
-aspects for their guidance:
+The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
+assistants have access to the data store and can use it to provide guidance about both data and
+code. Second, Wrattler includes an extensible framework for defining domain-specific languages
+(DSLs) that can describe, for example, data extraction, cleaning, transformation or visualization.
+The data store provides AI assistants with
+two sources of information that they can exploit
+to determine what guidance to provide the analyst:
 %
@@
 versions of a file.
-\vspace{0.25em}
 
@@
 notebook. This allows the them to access the transformation history, which is the code
-that has already been run during the analysis. The idea is that assistants may need to know the 
-context of transformations that have already been performed in order to decide what to suggest next.
+that has already been run during the analysis. The gives assistants information about the
+context such as transformations that have already been performed.
 \end{itemize}
 
+\medskip \noindent
 The assistants should be self-contained in the sense that they interact with each other
@@
 
-\vspace*{1em}
-
 %%cs possibly we don't need to go into detail
@@
 
-{\color{red}
-\textbf{TP: My rough notes from discussion with JG}
+The data store links together individual components of Wrattler. It stores the raw input
+data for data analysis such as downloaded web pages, raw data frames and also data frames
+produced as the result of running a script or invoking an AI assistant. In addition to data,
+the data store also stores multiple versions of the dependency graph created by Wrattler.
 
-~
+In the initial version of Wrattler, we restrict the data storage to raw input files and
+data frames, however we note that it would be desirable to include other data structures
+such graph or network data.
+
+In this section, we focus on the initial functionality that is required from the data store.
+We focus on the public interface exposed by the data store, but We do not discuss the technolgy
+used for data storage -- this can be decided during the development.
+
+Data frames stored in the data store can be annotated with metadata. Such metadata can
+communicate information between multiple AI assistants. For example, first AI assistant can
+infer that a type of a pair of columns represents GPS coordinate and produce a data frame
+with annotations specifying this information. A second AI assistant can then use this information
+and automatically provide a data visualization in the form of map. In the rest of this section,
+we discuss the structure of those annotations.
+
+\subsection{Semantic annotations}
+As noted in Section~\ref{sec:wrattler-ds}, annotations can be attached to individual cells,
+entire columns and rows, but also combinations of multiple columns or rows. As an example,
+consider the following small table that records data on broadband speed (and illustrates a number
+of real-world problems):
+
+\vspace{1em}
+\noindent
+\begin{tabular}{llc}
+\toprule
+\textbf{ISP}\qquad\qquad\qquad & \textbf{Postcode}\qquad\qquad & \textbf{Download} \\
+\midrule
+Virgin & CB43BE & 46.34 \\
+BT & CB43PT & N/A \\
+Virgin & NW53HL & 41.2 \\
+BT & Tottenham & 16.4 \\
+\bottomrule
+\end{tabular}
+\vspace{1em}
 
 \noindent
-Data store stores data frames with metadata. This is a good start for data science work,
-because most languages can provide mappings to/from data frames.
-}
+If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
+it will annotate the table with a number of information:
+\begin{itemize}
+\item[--] aa
+\end{itemize}
 
-\subsection{Semantics annotations}
 {\color{red}

### 2018-02-28 report/paper.tex
Data store format section

--- 2018-02-27 report/paper.tex
@@
 entire columns and rows, but also combinations of multiple columns or rows. As an example,
-consider the following small table that records data on broadband speed (and illustrates a number
-of real-world problems):
+consider the following small table obtained from a government web site that records data 
+on broadband speed (and illustrates a number of real-world problems):
 
 \vspace{1em}
-\noindent
 \begin{tabular}{llc}
@@
 Virgin & NW53HL & 41.2 \\
-BT & Tottenham & 16.4 \\
+BT & London & 16.4 \\
 \bottomrule
@@
 \noindent
-If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
-it will annotate the table with a number of information:
+As a first step, the data analyst might use the probabilistic type inference (ptype) assistant
+in order to infer the types of data in the data set. The assistant needs to be able to annotate
+the table with the following information:
+%
 \begin{itemize}
-\item[--] aa
+\item[--] With high certainty, the ISP column is a categorical column with values Virgin and BT;
+  with lower certainty, it could also be a string.
+\item[--] The column Postcode is most likely a postcode; with lower probability it is a city name
+  and with similarly low probability, it is a plain string.
+\item[--] The column Download is most likely a number or a missing value represented as N/A.  
+\item[--] The N/A cell is annotated as being most likely a missing value, or less likely a plain 
+  string. Other cells of the Download column are annotated as most likely being decimal numbers.
+\item[--] The cells of the Postcode column are annotated similarly. The first three are most
+  likely postcodes, while the last one is most likely a city name, but all also have a non-zero
+  probability of being an arbitrary string.
 \end{itemize}
 
-{\color{red}
-Data store provides a way of annotating cells, columns, combinations of columns and rows.
-For example, a column may be annotated with a type that indicates that its values are a mix of
-postcodes and city names. A cell can be annotated with an information saying that this value is 
-more likely a postcode than a city name. We also need to support annotaitons on a combination of
-columns, because three columns can jointly represent an address. A row can be annotated with a 
-provenance -- for example, when merging rows from two data sets with different sources.
-}
+\noindent
+Assuming the data frame is extracted from a government web page, each row of the data set
+could further be annotated with the URL of the data source. This makes it possible to track
+provenance at a fine-grained level. If we then append two such data frames, we will know
+which row comes from which government data set.
+
+It is worth noting that the data store does not need to fully understand the semantics of such 
+annotations. The probabilistic type inference assistant simply adds them and another assistant
+can then clean the data and, for example, drop values that do not conform to the most 
+probable type. We return to this topic in Section~\ref{sec:datastore-schema}, after discussing
+the format of such annotations.
 
 \subsection{Annotation format}
-{\color{red}
-We want a lightweight annotation format so that implementing an AI assistant or a language
-runtime that needs to communicate with the data store is not too hard.
-Our annotation format will follow something like \url{http://schema.org}, which defines
-a hierarchy of entity types and has a simple way for adding annotations to things.
-We will need to define our own hierarchy for schemas for data science.
-}
+The format of annotations needs to be light-weight so that it can be easily consumed and produced
+by AI assistants created in a variety of different programming languages. For this reason, it is
+desirable to use a standard file format such as JSON or XML. Furthermore, the format needs to 
+be extensible -- although Wrattler should provide a schema for basic annotations, new AI assistants
+should be able to freely define and use new kinds of annotations.
+
+An annotation format that satisfies the above constraints can take inspiration from the 
+Schema.org project (\url{http://schema.org}), which provides a way of annotating web content
+with semantic information. The focus of Schema.org is on higher-level entities (such as 
+a person, an event or a place), rather than lower-level information, but the format itself could
+be reused by Wrattler.
+
+As an example, consider the following Schema.org description of a place (the following shows a
+JSON encoding, but it is worth noting that Schema.org also supports RDF encoding):
+%
+\begin{equation*}
+\begin{array}{l}
+\{~~\; \str{@context}: \str{http://schema.org},\\
+\quad  \str{@type}: \str{Place},\\
+\quad  \str{name}: \str{The Black Lodge},\\
+\quad  \str{address}: \{\\
+\qquad    \str{@type}: \str{PostalAddress},\\
+\qquad    \str{addressLocality}: \str{London},\qquad\qquad\qquad\qquad\qquad~\\
+\qquad    \str{postalCode}: \str{NW53HL} ~\}~\}\\
+\end{array}
+\end{equation*}
+
+\noindent
+The \str{@context} attribute defines a namespace of a schema that defines the available
+entities. The default namespace, used here, includes two entities that we use in this 
+example, namely Place and PostalAddress.
+
+The namespace mechanism provides a convenient way of extending annotations -- Wrattler can
+define a core schema for common information that it needs, but new AI assistants can 
+extend it via custom namespaces.
+
+The Schema.org format also makes it easy to attach additional information to an existing
+value. In the example above, the original value might have been ``The Black Lodge'', but 
+an AI assistant -- possibly using entity linkage -- inferred and extended it with postal address.
 
 \subsection{Schemas for data science}
+\label{sec:datastore-schema}
+
 {\color{red}

### 2018-02-28 report/paper.tex

--- 2018-02-28 report/paper.tex
@@
 \noindent
-If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
-it will annotate the table with a number of information:
+As a first step, the data analyst might use the probabilistic type inference (ptype) assistant
+in order to infer the types of data in the data set. The assistant needs to be able to annotate
+the table with the following information:
+%
 \begin{itemize}
-\item[--] aa
+\item[--] With high certainty, the ISP column is a categorical column with values Virgin and BT;
+  with lower certainty, it could also be a string.
+\item[--] The column Postcode is most likely a postcode; with lower probability it is a city name
+  and with similarly low probability, it is a plain string.
+\item[--] The column Download is most likely a number or a missing value represented as N/A.  
+\item[--] The N/A cell is annotated as being most likely a missing value, or less likely a plain 
+  string. Other cells of the Download column are annotated as most likely being decimal numbers.
+\item[--] The cells of the Postcode column are annotated similarly. The first three are most
+  likely postcodes, while the last one is most likely a city name, but all also have a non-zero
+  probability of being an arbitrary string.
 \end{itemize}
 
-{\color{red}
-Data store provides a way of annotating cells, columns, combinations of columns and rows.
-For example, a column may be annotated with a type that indicates that its values are a mix of
-postcodes and city names. A cell can be annotated with an information saying that this value is
-more likely a postcode than a city name. We also need to support annotaitons on a combination of
-columns, because three columns can jointly represent an address. A row can be annotated with a
-provenance -- for example, when merging rows from two data sets with different sources.
-}
+Assuming the data frame is extracted from a government web page, each row of the data set
+could further be annotated with the URL of the data source. This makes it possible to track
+provenance at a fine-grained level. If we then append two such data frames, we will know
+which row comes from which government data set.
+
+It is worth noting that the data store does not need to fully understand the semantics of such 
+annotations. The probabilistic type inference assistant simply adds them and another assistant
+can then clean the data and, for example, drop values that do not conform to the most 
+probable type. We return to this topic in Section~\ref{sec:datastore-schema}, after discussing
+the format of such annotations.
 
 \subsection{Annotation format}
-{\color{red}
-We want a lightweight annotation format so that implementing an AI assistant or a language
-runtime that needs to communicate with the data store is not too hard.
-Our annotation format will follow something like \url{http://schema.org}, which defines
-a hierarchy of entity types and has a simple way for adding annotations to things.
-We will need to define our own hierarchy for schemas for data science.
-}
+The format of annotations needs to be light-weight so that it can be easily consumed and produced
+by AI assistants created in a variety of different programming languages. For this reason, it is
+desirable to use a standard file format such as JSON or XML. Furthermore, the format needs to 
+be extensible -- although Wrattler should provide a schema for basic annotations, new AI assistants
+should be able to freely define and use new kinds of annotations.
+
+An annotation format that satisfies the above constraints can take inspiration from the 
+Schema.org project (\url{http://schema.org}), which provides a way of annotating web content
+with semantic information. The focus of Schema.org is on higher-level entities (such as 
+a person, an event or a place), rather than lower-level information, but the format itself could
+be reused by Wrattler.
+
+As an example, consider the following Schema.org description of a place (the following shows a
+JSON encoding, but it is worth noting that Schema.org also supports RDF encoding):
+%
+\begin{equation*}
+\begin{array}{l}
+\{~~\; \str{@context}: \str{http://schema.org},\\
+\quad  \str{@type}: \str{Place},\\
+\quad  \str{name}: \str{The Black Lodge},\\
+\quad  \str{address}: \{\\
+\qquad    \str{@type}: \str{PostalAddress},\\
+\qquad    \str{addressLocality}: \str{London},\qquad\qquad\qquad\qquad\qquad~\\
+\qquad    \str{postalCode}: \str{NW53HL} ~\}~\}\\
+\end{array}
+\end{equation*}
+
+\noindent
+The \str{@context} attribute defines a namespace of a schema that defines the available
+entities. The default namespace, used here, includes two entities that we use in this 
+example, namely Place and PostalAddress.
+
+The namespace mechanism provides a convenient way of extending annotations -- Wrattler can
+define a core schema for common information that it needs, but new AI assistants can 
+extend it via custom namespaces.
+
+The Schema.org format also makes it easy to attach additional information to an existing
+value. In the example above, the original value might have been ``The Black Lodge'', but 
+an AI assistant -- possibly using entity linkage -- inferred and extended it with postal address.
 
 \subsection{Schemas for data science}
+\label{sec:datastore-schema}
+
 {\color{red}

### 2018-02-28 report/paper.tex
Minor tweaks throughout to make layout nicer

@@
 data and discover hidden structures or corner cases, but only a human can decide what those mean
-and decide how to handle them in data science scripts. Data science is often cited as an expensive
+and decide how to handle them in scripts. Data science is often cited as an expensive
 and time consuming task, especially due the costs of data cleaning and data wrangling.
@@
 \vspace{1em}
+\noindent
 To meet these challenges, we present Wrattler,
@@
 with new advances in AI systems for data wrangling and in provenance.
-The main design principles in Wrattler are:
+The main design principles are:
 
 \paragraph{Interactive.} Interactivity
-is a necessary because ``big data is big'', so the analyst learns about
+is a necessary because ``big data is big''. The analyst learns about
 the data set and the problem as she explores it.
@@
 Notebooks run in the browser, cache partial results of computations and provide previews
-of script results on-the-fly during development.
+of script results on-the-fly.
 
@@
 
-\begin{figure}
-\vspace{-0.5em}
-\includegraphics[scale=0.6]{diagram.pdf}
-\vspace{-0.5em}
-\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
-  Wrattler, those functions are separated and enriched with AI assistants.}}
-\label{fig:arch}
-\vspace{-0.5em}
-\end{figure}
-
 The architecture of Wrattler allows us to address these issues, as well as to provide a platform
@@
 in Figure~\ref{fig:arch}. The components of Wrattler are:
-%
+
+\vspace{-0.25em}
 \paragraph{Data store.} Imported external data, results of running scripts and of
@@
 
+\vspace{-0.25em}
 \paragraph{Language runtimes.} Scripts are evaluated by one or more language runtimes.
@@
 
+\vspace{-0.25em}
 \paragraph{AI assistants.} When invoked from the notebook, AI assistants read data
@@
 
+\vspace{-0.25em}
 \paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates
@@
 
-Wrattler consists of a notebook user interface running in the web browser, which communicates with
-a number of server-side components including language runtimes, the data store and AI assistants.
-In this section, we discuss the components in more detail, starting with the dependency graph
-that is maintained on the client-side, by the web browser. AI assistants are discussed later
-in Section~\ref{sec:ai}
+Wrattler consists of a user interface running in the web browser, which communicates with
+a number of server-side components. In this section, we discuss the components in more 
+detail, starting with the dependency graph that is maintained on the client-side, by the 
+web browser. AI assistants are discussed later in Section~\ref{sec:ai}.
 
@@
 
+\begin{figure}
+\includegraphics[scale=0.6]{diagram.pdf}
+\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
+  Wrattler, those functions are separated and enriched with AI assistants.}}
+\label{fig:arch}
+\end{figure}
+
 \subsection{Data store}
@@
 \begin{array}{l}
-\textit{\small{We retrieve data from a government web site using Python:}}\\[0.3em]
+\textit{\small{In the first block, we retrieve data from a government web site.}}\\[-0.1em]
+\textit{\small{We use Python to download the data and parse it as a CSV file.}}\\[0.3em]
 \quad\ident{url}=\ident{open}(\str{http://data.gov.uk/\ldots/bb2014.csv})\\
@@
 \end{figure}
-\begin{figure}
-\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
-
-\caption{\small{Dependency graph for the notebook in Figure~\ref{fig:notebook}.
-Data acquisition is represented as a single Python cell. Data cleaning is understood
-by Wrattler and is the graph has a node for each call and constant. Hexagons represent
-data frames shared by cell and stored in the data store.}}
-\label{fig:deps}
-\vspace{-0.5em}
-\end{figure}
 
@@
 the construction of dependency graph. They can run as server-side components (e.g.~for R and Python)
-or as client-side components (for JavaScript and small analytical DSLs used by AI assistants).
+or as client-side components (in case of JavaScript).
 Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
@@
 store. This new data frame can contain new data or additional information (such as their
-inferred probabilistic type).
+inferred types).
 The interaction is illustrated in Figure~\ref{fig:arch}. An AI assistant reads data from the data
@@
 Assistants are therefore based on two different ideas:
+
+\begin{figure}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+
+\caption{\small{Dependency graph for the notebook in Figure~\ref{fig:notebook}.
+Data acquisition is represented as a single Python cell. Data cleaning is understood
+by Wrattler and is the graph has a node for each call and constant. Hexagons represent
+data frames shared by cell and stored in the data store.}}
+\label{fig:deps}
+\vspace{-0.5em}
+\end{figure}
 
@@
 the datadiff assistant offers the most likely patches and the user can choose which ones
-to accept. Scripts constructed in such way are readable, allowing the data analyst to look at them
+to accept.
+
+Scripts consisting of a chain of simple method calls are readable, allowing the data analyst to look at them
 and verify that the transformations accomplish the intended task (such verification further
@@
 \vspace{1em}
+\noindent
 The Wrattler architecture supports the development of AI assistants in a number of ways. First, AI
@@
 (DSLs) that can describe, for example, data extraction, cleaning, transformation or visualization.
+
 The data store provides AI assistants with
@@
 
+\noindent
 Assuming the data frame is extracted from a government web page, each row of the data set

### 2018-02-28 report/paper.tex
Finish draft notes on the data store

@@
 
-{\color{red}
-This hierarchy will include some basic things that the data store understands and can
-convert between (integers, floats, unbounded size integers, ISO dates, etc.)
+Just like the default namespace of Schema.org defines common types of entities, Wrattler
+should define basic schema for the most common annotations. This includes two kinds of annotations.
+First, there are primitive type annotations that are understood by the data store. Second,
+there are frequent semantic annotations that are generally useful.
 
-In addition, we will define some purely semantic annotations that are commonly useful and
-we will define a hierarchy for those. This represents things like addresses, postcodes, countries,
-cities, etc.
+\paragraph{Schemas for primitive types.} The data store will store primitive values in the
+data frame in a suitable native format such as floating-point number, Boolean, string or 
+ISO 8601 date. However, this representation should remain internal and data store should be
+able to consume and expose data in formats required by the client (this is further discussed
+in Section~\ref{sec:datastore-negotiation}). This will be done through primitive type annotations.
+The data store will understand those and will be able to convert between them automatically.
+The primitive type annotations include:
+%
+\begin{itemize}
+\item[--] \textbf{Numerical formats}. The data store will be able to convert between a number of common numerical
+  formats such as integers, floats and unbounded size integers. In addition, values can also be
+  formatted differently. Examples include the scientific notation (1.34e10) or variations in 
+  decimal separators used in different cultures.
 
-Everyone can define their own schemas (like schema.org) by defining their own namespace (URL)
-and providing entities with their custom namespace. The data store will simply save and return
-this information - without doing any checking.}
+\item[--] \textbf{Dates}. The data store will be able to recognise and convert between a number of
+  standard date formats (such as ISO 8601 format, dates in US and non-US formats).
+
+\item[--] \textbf{Missing values}. The data store will record missing values internally and will be
+  able to return them in a variety of formats, depending on the client request (NaN, special 
+  strings such as empty or \#!NULL, etc).
+
+\item[--] \textbf{Geo coordinates}. Finally, the list of primitive types might also include geographical
+  coordinates, in which case the data store will recognize basic formats such as GPS coordinates
+  and Northings and Eastings.  
+\end{itemize}
+
+\noindent
+The above primitive types need to be understood by the data store, in order to enable easy
+communication between clients written in a variety of programming environments that use different
+primitive types. The key idea is that each client will only need to handle, for example, one format 
+for dates. The data store will be able to communicate using this format, even when the value was
+written in another format by another client.
+
+In addition, Wrattler should define a set of annotations that can be used by a wide range of
+standard data sets. It is useful to provide a basic schema for such common annotations as this
+will enable a range of AI assistants and other tools to exchange information without first having
+to define a common custom schema. Those should include annotations for places (addresses, 
+countries, cities, postcodes), physical units (meters, killograms, etc.), contact information
+(e-mails, phone numbers) and similar. We will investigate whether Schema.org or other existing
+standards developed by the semantic web community already cover those areas.
 
 \subsection{Content negotiation}
-{\color{red}
-The only bit where the data store does something clever is that it will be able to return data
-in a format that the client asked for (a bit like HTTP negotiation). The client can say, ``I only
-understand semantic information defined by these schema namespaces`` and the data store will do
-its best to return data in the required format.
+\label{sec:datastore-negotiation}
 
-For simple types (e.g. dates) that the data store understands, it will convert data accordingly.
-For other semantic information that the data store does not understand (e.g. information that
-a certain cell is missing with a probability $p$), the data store will filter out the extra
-information and just return the raw number.
-}
+As noted earlier, the data store will understand a range of primitive data types and will be 
+able to convert between them. This will be done using a content negotiation protocol,
+similar to the Accept/Content-Type header pair in the HTTP protocol. Rather than using 
+content type as in HTTP, the negotiation will be based on the schema. For example, a client may
+specify that it accepts data with a certain namespace or that it recognizes a specified list of
+types.
+
+For primitive types that the data store understands, it will convert data according to the
+request made by the client. For example, a date can be automatically converted between 
+mm/dd and dd/mm formats. For additional semantic annotations that the data store does not
+understand, it will simply drop the extra information not required by the client.
+
+This makes it possible to exchange data efficiently. For example, a data frame where each cell
+is annotated with a probabilistically inferred type will be significantly larger than a data
+frame contianing just raw numbers. The content negotiation protocol allows the client to 
+request a data frame with just raw numbers, significantly reducing the size.
 

### 2018-03-06 report/paper.tex
Fix latex bugs, generate nice PDF
Minor style tweaks

@@
 \usepackage{enumitem}
-\usepackage[shortlabels]{enumerate}
+\usepackage{enumerate}
 \setlist{leftmargin=6mm}
-
-\newcounter{thc}
-\newcounter{dfc}
-
-\theoremstyle{plain}
-\newtheorem{lem}[thc]{Lemma}
-\newtheorem{theorem}[thc]{Theorem}
-
-\theoremstyle{definition}
-\newtheorem{definition}[dfc]{Definition}
 
@@
 
-\author{Authors}
+\author{AIDA Team, The Alan Turing Institute}
 %\affiliation{
@@
 \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
 \newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
@@
 
-As an example, consider the following Schema.org description of a place (the following shows a
-JSON encoding, but it is worth noting that Schema.org also supports RDF encoding):
-%
+\begin{figure}
 \begin{equation*}
@@
 \end{equation*}
+\caption{\small{Example of a Schema.org annotation using a JSON encoding (it is worth noting 
+that Schema.org also supports RDF).}}
+\label{fig:schema}
+\vspace{-1em}
+\end{figure}
 
-\noindent
+As an example, Figure~\ref{fig:schema} shows a Schema.org description of a place.
 The \str{@context} attribute defines a namespace of a schema that defines the available

### 2018-03-12 report/paper.tex
fixing typos

@@
 the vast array of disparate use cases of data science,
-ranging form relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.
+ranging from relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.
 
@@
 \paragraph{Interactive.} Interactivity
-is a necessary because ``big data is big''. The analyst learns about
+is necessary because ``big data is big''. The analyst learns about
 the data set and the problem as she explores it.
 Wrattler enables an efficient interaction by bringing computation closer to the human.
-Notebooks run in the browser, cache partial results of computations and provide previews
+Wrattler notebooks run in the browser, cache partial results of computations and provide previews
 of script results on-the-fly.
@@
 
-Notebook systems such as Jupyter became a popular programming environment for data science, because
+Notebook systems such as Jupyter have become a popular programming environment for data science, because
 they support gradual data exploration and provide a convenient way of interleaving code with
-comments and visualizations. However, notebooks suffer from a number of issues that hanper
+comments and visualizations. However, notebooks suffer from a number of issues that hamper
 reproducibility and limit the possible interaction model.
@@
 
-The data store provides a way for persistently stroing data and enables communication between
+The data store provides a way for persistently storing data and enables communication between
 individual Wrattler components. Data store keeps external data files imported into Wrattler,
@@
 \vspace{0.25em}
-\item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or and address.
+\item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or address.
 \vspace{-0.85em}
@@
 Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
-the data store. (Using a cache for efficiency.)
+the data store (using a cache for efficiency).
 
@@
 
-Wrattler includes a scripting language with simple syntax that is used by individual AI assistants
-to define domain specific languages. The syntax is shared across multiple AI assistnts, but the
-operations are provided by individual assistants.
+Wrattler includes a scripting language with simple syntax that is used
+by individual AI assistants to define domain specific languages. The
+syntax is shared across multiple AI assistants, but the operations are
+provided by individual assistants.
 
@@
 keeps a fine-grained representation of the code in the dependency graph (second cell in
-Figure~\ref{fig:deps}), which allows Wrattler to provie live previews of the result, giving the
+Figure~\ref{fig:deps}), which allows Wrattler to provide live previews of the result, giving the
 user an immediate feedback.
@@
 example of a tool in this space is to
- identifying pairs of variables across multiple data
+ identify pairs of variables across multiple data
 sources that are likely to be daily temperatures.
@@
 but there are still a huge number of issues that have been unexplored.
-For example, we envisage assistants for
-Identifying values that seem to be out of range, e.g. missing value indicators;
-Identifying possible inconsistent coding of strings,
+For example, we envisage assistants for:
+identifying values that seem to be out of range, e.g. missing value indicators;
+identifying possible inconsistent coding of strings,
 such as different representations of acrynoms;
-indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
+identifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
 identifying distinct records that may refer to the same entity (record linkage).
@@
 concept drift. This includes the problem
-to suggesting when/how often new data points should be labelled for monitoring.
+of suggesting when/how often new data points should be labelled for monitoring.
 Even for data analyses that do not involve
 deploying a ``production model'', there is
-often a need to run an analysis periodically on updated data, such as re-running last quarter's
-analysis on this quarter's data.
+often a need to run an analysis periodically on updated data, such as
+re-running last quarter's analysis on this quarter's data.
 In such cases, we might believe that the two data
@@
 \paragraph{Debugging and interpreting predictive analytics.}
-Every time a new machine learning model is trained,
-the first question an analyst has is often
-why its accuracy is not better. But
-debugging predictive analytics is notoriously difficult, and the remedies indirect and expensive,
- e.g., labelling more data, adding more features,
- using a more complex model. Debugging and improving
- models is a rich area for potential assistants.
- There is potentially low hanging fruit available
- in how to aid the tasks of
-error analysis, exploring the
-predictions of the model
-on validation data. Some early work in this direction is
-\cite{saleema}.
-There is a rich and growing literature on
-interpreting and explaining the predictions
-of a model \cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.
-Two potential directions from data are
-debugging by using
-"training set blame": Given an error in the validation data, what examples from the training set caused me to make that error? This might build on the work of \cite{percy}.
-Another potential avenue is explaining
-differences between predictions: why was instance 42 treated differently than instance 24601, even though they appear highly similar to a human?
-}
+Every time a new machine learning model is trained, the first question
+an analyst has is often why its accuracy is not better. But debugging
+predictive analytics is notoriously difficult, and the remedies
+indirect and expensive, e.g., labelling more data, adding more
+features, using a more complex model. Debugging and improving models
+is a rich area for potential assistants.  There is potentially low
+hanging fruit available in how to aid the tasks of error analysis,
+exploring the predictions of the model on validation data. Some early
+work in this direction is \cite{saleema}.  There is a rich and growing
+literature on interpreting and explaining the predictions of a model
+\cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.  Two
+potential directions from are debugging by using "training set
+blame": Given an error in the validation data, what examples from the
+training set caused me to make that error? This might build on the
+work of \cite{percy}.  Another potential avenue is explaining
+differences between predictions: why was instance 42 treated
+differently than instance 24601, even though they appear highly
+similar to a human?  }
 
@@
 
-In this section, we focus on the initial functionality that is required from the data store.
-We focus on the public interface exposed by the data store, but We do not discuss the technolgy
-used for data storage -- this can be decided during the development.
+In this section, we focus on the initial functionality that is
+required from the data store.  We focus on the public interface
+exposed by the data store, but we do not discuss the technology used
+for data storage -- this can be decided during the development.
 
-Data frames stored in the data store can be annotated with metadata. Such metadata can
-communicate information between multiple AI assistants. For example, first AI assistant can
-infer that a type of a pair of columns represents GPS coordinate and produce a data frame
-with annotations specifying this information. A second AI assistant can then use this information
-and automatically provide a data visualization in the form of map. In the rest of this section,
-we discuss the structure of those annotations.
+Data frames stored in the data store can be annotated with
+metadata. Such metadata can communicate information between multiple
+AI assistants. For example, first AI assistant can infer that a pair
+of columns represent GPS lat/lon coordinates and produce a data frame
+with annotations specifying this information. A second AI assistant
+can then use this information and automatically provide a data
+visualization in the form of map. In the rest of this section, we
+discuss the structure of those annotations.
 
@@
 
-\item[--] \textbf{Geo coordinates}. Finally, the list of primitive types might also include geographical
-  coordinates, in which case the data store will recognize basic formats such as GPS coordinates
-  and Northings and Eastings.  
+\item[--] \textbf{Geo coordinates}. The list of primitive types might also include geographical
+  coordinates, in which case the data store will recognize basic formats such as GPS coordinates  and Northings and Eastings.
+
 \end{itemize}
@@
 to define a common custom schema. Those should include annotations for places (addresses, 
-countries, cities, postcodes), physical units (meters, killograms, etc.), contact information
+countries, cities, postcodes), physical units (meters, kilograms, etc.), contact information
 (e-mails, phone numbers) and similar. We will investigate whether Schema.org or other existing

### 2018-03-15 tapp/paper.tex
Adding TaPP draft

\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\setcopyright{none}

\bibliographystyle{ACM-Reference-Format}

\begin{document}
\title{Wrattler: \textnormal{Making notebooks reproducible, live and smart}}

\author{Tomas Petricek}
\affiliation{
  \institution{The Alan Turing Institute}
  \city{London}
  \country{UK}
}
\email{tomas@tomasp.net}
\author{Charles Sutton}
\affiliation{
  \institution{University of Edinburgh}
  \city{Edinburgh}
  \country{UK}
}
\email{csutton@inf.ed.ac.uk}
\author{James Geddes}
\affiliation{
  \institution{The Alan Turing Institute}
  \city{London}
  \country{UK}
}
\email{jgeddes@turing.ac.uk}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}

\begin{abstract}
Notebook systems such as Jupyter became a popular programming environment for data science, because 
they support interactive data exploration and provide a convenient way of interleaving code, 
comments and visualizations. However, notebooks also suffer from reproducibility issues and make
versioning and provenance tracking difficult.

In this paper, we present Wrattler, a new notebook system with an architecture that addresses the
above issues. Wrattler stores all state in a data store and separates state management from 
script evaluation. This makes it possible to support versioning, guarantee reproducibility, track 
data provenance, but also allow richer forms of interactivity and integrate AI tools that help
automate routine data wrangling tasks.
\end{abstract}
\keywords{notebook, dependency graph, live coding, AI}
\maketitle

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a computer
system and a human. Notebook systems support this interaction model by making it easy to run 
snippets of code and see results without leaving the notebook. However, typical notebook systems,
such as Jupyter suffer from a number of problems:

\paragraph{Limited reproducibility.} Cells can be executed out-of-order or modified without re-evaluating 
the notebook. This means that running all cells from scratch may give different results than those 
originally seen in the notebook.

\paragraph{Opaque state management.} All state is managed by the execution engine, or \emph{kernel}.
This makes the state invisible to tools such as version control systems and makes it hard to 
track provenance in a notebook.

\paragraph{Limited interaction.} Notebooks execute code at the granularity of an entire cell. 
This means that even simple code change may trigger a long-running computation and, as a result, 
notebooks do not always provide rapid feedback.

\paragraph{Single language.} When using a notebook, the state is managed by a single \emph{kernel}.
This makes it difficult to combine multiple programming languages in a single notebook, or build
third-party tools that would inspect the state and provide data analysts with hints about data.

\vspace{1em}
\noindent
In this paper, we present Wrattler, a new notebook system that addresses the above issues.
This is made possible by two key aspects of the Wrattler architecture (Section~\ref{sec:arch}).

First, Wrattler maintains a dependency graph between cells and often also function calls inside
a single cell (Section~\ref{sec:comp-deps}). When a cell is changed, relevant parts of a graph
are invalidated. This guarantees reproducibility and it enables more efficient re-computation, 
because values of unaffected nodes can be reused.
Second, Wrattler separates state management from code execution (Section~\ref{sec:comp-data}). 
The \emph{data store} is responsible for managing state. It handles versioning, provenance tracking 
and enables multiple independent languages and tools to operate on the state of a notebook.

Together, these two changes to the standard architecture of notebook systems make Wrattler notebooks
(Section~\ref{sec:results}) reproducible (with an easy and reliable state rollback), live (with 
efficient re-computation on change) polyglot and smart (allowing the integration of multiple 
execution engines and third-party AI tools).

\begin{figure}[b]
\vspace{-1em}
\includegraphics[scale=0.6]{diagram.pdf}
\vspace{-0.5em}
\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are split between data store and language runtimes. Data in the data store
  can be also accessed by other thrd-party tools.}}
\label{fig:arch}
\end{figure}

\section{Wrattler architecture}
\label{sec:arch}

In standard notebook systems, such as Jupyter, the state and execution is handled by a 
\emph{kernel}. The notebook running in a browser sends commands to the kernel to evaluate 
cells selected by the user. As illustrated in Figure~\ref{fig:arch}, Wrattler splits the 
server functionality between the following components:

\paragraph{Data store.} Imported external data, results of running scripts and of 
running third-party tools are stored in the data store. The data store keeps version history
and annotates data with metadata such as types, inferred semantics and provenance.

\paragraph{Language runtimes.} Code in notebook cells is evaluated by language runtimes.
The runtimes read input data from and write results back to the data store. Wrattler supports
language runtimes that run code on the server (similar to Jupyter kernels), but also language
runtimes that parse and evaluate code directly in the browser.

\paragraph{Third-party tools.} In addition to language runtimes, other tools can contribute
to data analysis by analysing data in the data store. In Wrattler, this includes AI assistants 
that help, for example, with data cleaning, type inference and data extraction from raw 
input data.

\paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates 
all other components. The browser builds a dependency graph between cells or individual 
expressions in the cells. It calls language runtimes to evaluate code that has changed,
third-party tools to provide hints and reads data from the data store to display results.

\begin{figure}
\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
\caption{\small{Dependency graph of a sample notebook with three cells. The cells are written
in Python, Wrattler script and JavaScript. The browser builds a fine-grained dependency graph
for Wrattler script that can be parsed and analysed in the browser.}}
\label{fig:deps}
\vspace{-0.5em}
\end{figure}

\section{Wrattler components}
\label{sec:comp}

Wrattler consists of a notebook user interface running in the web browser, which communicates 
with a number of server-side components including language runtimes, the data store and third-party 
tools. In this section, we discuss the components in more detail.

\subsection{TheGamma script}
\label{sec:comp-gamma}

The Wrattler architecture supports languages that require running a separate language runtime
as a server, but also languages that can be fully evaluated in the browser. To illustrate this
feature, we integrated Wrattler with TheGamma [X], a simple browser-based language for data
exploration.

As an example, we use a data set with broadband speed published by the UK government [X].
The following TheGamma script loads the raw CSV file, groups rows by whether they represent
rural or urban area and then calculates average download speed for both of the kinds of areas:
%
\begin{equation*}
\begin{array}{l}
\ident{web.loadTable}(\str{https://.../bb2014.csv}).\ident{explore}.\\
\quad\qident{group data}.\qident{by Urban/rural}.\\
\qquad\qident{average Download speed (Mbit/s) 24 hrs}  
\end{array}
\end{equation*}
%
Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} [X] 
and the user typically interacts by selecting one of generated members through an auto-complete
(hence the long explicity names are acceptable). For the purpose of this paper, the most 
important aspect is that TheGamma scripts can be parsed, checked and evaluated in the browser.

\subsection{Dependency graph}
\label{sec:comp-deps}

When opening a notebook, Wrattler parses the source of the notebook (consisting of text cells and
code cells) and constructs a dependency graph, which serves as a runtime representation of the
notebook.

As an example, consider a simpler notebook with three cells. The first cell, written in Python,
downloads data and exports the result as a data frame; the second cell is written in TheGamma script
and performs data cleaning and the third cell creates a visualization in JavaScript. The
resulting dependency graph is shown in Figure~\ref{fig:deps}.

Wrattler creates at least two nodes for each cell, representing the cell as a whole and its source
code. In addition, it creates a node for each data frame exported by a cell (e.g.~the rightmost
node in the first cell). Code in subsequent cells can depend on data frames exported from earlier
cell, which is captured as a dependency in the graph.

For code snippets that can be analysed in 
the browser (such as TheGamma cells), Wrattler creates a more fine-grained dependency graph with
a node for each operation (such as member access or a function call). This is the case in the 
second cell in Figure~\ref{fig:deps} and it allows more efficient live update of results when 
code changes as we only need to re-evaluate parts of code in a cell. We briefly outline how the
dependency graph is constructed and used.

\paragraph{Dependency graph construction.} 
The dependency graph is reconstructed after every change in any of the cells. This is an efficient
process, because the changes are typically small. Wrattler first parses each cell, producing a
single syntactic element for Python, JavaScript and R cells (representing the entire source code)
and a syntax tree for TheGamma (representing individual member accesses and other constructs).
The syntax tree for the entire notebook is then a list of elements produced for each cell.

After parsing, Wrattler walks over the syntax tree recursively and binds a dependency graph node
to each syntactic element in the tree. The \emph{antecedents} of a node are the nodes that it
depends on. This typically includes inputs for an operation or instance on which an operation
is invoked. The binding procedure is decribed in Figure~\ref{fig:bind}.

\paragraph{Checking and evaluation.} 
Nodes in the dependency graph can be annotated with information such as the evaluated value
of the syntactic element that the node represents. An important property of the binding process
is that, if there is no change in antecedents of a node, binding will return the same node
as before. This means that if we evaluate a value for a given node in a dependency graph and
attach it to a node, the value will be cached and reused.

Wrattler does not automatically re-evalaute the entire dependency graph, but allows the user to
trigger re-evaluation. This provides a user experience familiar to other notebook systems. 
However, the displayed results and visualizations always reflect the current source code in the
notebook. When the evaluation of a cell is requested, Wrattler recursively evaluates all the 
antecedents of the node and then evaluates the value of the node.

The evaluation is delegated to a language
runtime associated with the language of the node and it can be done in several ways:
%
\begin{enumerate}
\item For Python or R nodes, the language runtime ensures that the values of the dependencies
  are stored in the data store and sends the source code, together with its dependencies, to a
  server that retrieves the dependencies and evaluates the code.
\item For TheGamma and JavaScript nodes, the language runtime collects values of the 
  dependencies and runs the operation that the node represents in the browser.
\end{enumerate}
%
Wrattler also uses the dependency graph for type checking. Just like evaluation, type checking
is done by recursively walking over the dependency graph and annotating the nodes with their 
type. This step is only relevant for cells written in statically typed languages and provide
quicker feedback, because it does not need to evaluate the cells.

\subsection{Data store}
\label{sec:comp-data}

The data store enables communication between individual Wrattler components and provides a way for 
persistently stroing data. The data stored in the data store is associated with the hash 
produced by the binding process outlined in Figure~\ref{fig:bind} and is immutable. When the
notebook changes, new nodes with new hashes are created and so the new results are appended,
rather than overwriting existing state.

The data store in Wrattler (currently) supports two data formats -- external data files imported
into Wrattler notebooks (such as downloaded web pages) and data frames. The latter are used for
communication between multiple cells of a Wrattler notebook. A cell that exports a data frame
so that it can be used by subsequent cells needs to store the data frame into a data store (the
key is the hash of the node representing the data frame).

The data store additionally also supports a mechanism for annotating data frames with additional 
semantic information. Columns can be annotated with (and stored as) primitive data type such as 
date or floating-point number. Columns can also be annotated with semantic annotation that indicates
the meaning of the column -- for example, address or longitude and latitude. Finally, columns,
rows and individual cells of the data frame can be annotated with other metadata such as their
provenance.

\begin{figure}
\vspace{-1em}
\begin{equation*}
\hspace{-6em}\begin{array}{l}
\kvd{procedure}~\ident{bind}(\textit{cache}, \texit{syn})~=\\
\quad \kvd{let}~h = \{ \ident{hash}(\ident{kind}(\textit{syn})) \} \\  
\quad \qquad \cup~ \{\ident{hash}(c) \,|\, c\in\ident{antecedents}(\textit{syn}) \}\\
\quad \kvd{if}~\ident{not}~\ident{contains}(\textit{cache}, h)~\kvd{then}\\
\quad \quad \kvd{let}~n = \emph{fresh node}\\
\quad \quad \ident{value}(n) \leftarrow \ident{Unevaluated} \\
\quad \quad \ident{hash}(n) \leftarrow h\\
\quad \quad \ident{set}(\textit{cache}, h, n) \\
\quad \ident{lookup}(\textit{cache}, h)\\
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{\small{When binding a graph node to a syntactic element, Wrattler first computes
  a set of hashes that uniquely represent the node. This includes hash of the kind of the 
  node (e.g. the source code of a Python node or member name in TheGamma) and hashes
  of all antecedents. If a node with a given hash does not exist in cache, a new node
  is created. We set its hash, indicate that its value has not been evaluated and
  add it to the cache.}}
\label{fig:bind}
\end{figure}

In addition to storing the raw data, the data store also persistently stores the current and
multiple past versions of the dependency graph constructed from the notebook (saved by an 
explicit checkpoint). This makes it possible to analyse the history of a notebook and track how
data is transformed by the computation in a notebook.

\subsection{Third-party tools}
\label{sec:comp-ai}

The Wrattler architecture enables integration of third-party tools that analyse the notebook
or data it uses and provide hints to the data analyst. Most notably, such tools include AI 
assistants that help with tedious data cleaning and pre-processing tasks.

As an example, \emph{datadiff} is an AI assistant that reads two specified data frames from the
data store and suggests a script that transforms the data in the second data frame to match the
format of the first data frame. The AI assistant does not automatically transofrm the data, but
instead, produces a new cell in TheGamma script that can be reviewed by the data analyst. 
For example:
%
\begin{equation*}
\begin{array}{l}
\ident{datadiff(broadband2014, broadband2015)}\\
\quad.\ident{drop\_column}(\str{WT\_national})\\
\quad.\ident{drop\_column}(\str{WT\_ISP})\\
\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
\end{array}  
\end{equation*}
%
The script specifies that two columns should be dropped from the corrupted second data frame and
one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
\strf{Urban} and \strf{Rural}).

\section{Results}
\label{sec:results}

\paragraph{Properties.}
The dependency graph enables several features that are difficult for most notebook systems:
%
\begin{itemize}
\item[--] When code changes, Wrattler only needs to recompute a small part of the graph.
  This makes it possible to provide live previews, especially for simple data analytical DSLs.
\item[--] Refactoring can extract parts of the notebook that
  are needed to compute a specified output.
\vspace{0.25em}
\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
\vspace{0.25em}
\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
  or code cleanup.
\end{itemize}

Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
Once it is interactively constructed, it can be translated to multiple supported languages
such as R, Python or JavaScript.

\section{Related work}

dataflow notebooks

\newpage
~
\newpage

Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice 
based on statistical analysis of the data and discover hidden structures or corner cases, but only
a human can decide what those mean.
A data science environment of the future thus needs to support an interaction model that 
keeps the human in the loop, allows an efficient human-computer interaction 
and can be extended with new AI asistants that provide advice about data.
In this report, we present Wrattler, a notebook system that is:

\paragraph{Interactive.}
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly during development.

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state.
A data store tracks the provenance and semantics of data, supports versioning and keeps 
the history, making the data analyses fully reproducible.

\paragraph{Polyglot.}
Multiple languages can be used in a single notebook and share data via the data store.
Analysts can use R and Python, but also interactive languages for data exploration
that run in the browser and provide live previews.

\paragraph{Smart.}
Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
help with data cleaning and joining, but also help data exploration by finding typical and atypical 
data points and automatically visualizing data.

\section{Wrattler components}

Wrattler consists of a notebook user interface running in the web browser, which communicates with 
a number of server-side components including language runtimes, the data store and AI assistants. 
In this section, we discuss the components in more detail, starting with the dependency graph
that is maintained on the client-side, by the web browser.

\subsection{Dependency graph}

When opening a notebook, Wrattler parses the source of the notebook (consisting of text cells and 
code cells) and constructs a dependency graph, which serves as a runtime representation of the
notebook.

The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
depend on a data frame (hexagons) exported by an earlier cell. 
When code in a cell changes, Wrattler updates the dependency graph, keeping existing
nodes for entities that were unaffected by the change.

The dependency graph enables several features that are difficult for most notebook systems:
%
\begin{itemize}
\item[--] When code changes, Wrattler only needs to recompute small part of the graph.
  This makes it possible to provide live previews, especially for simple data analytical DSLs.
\vspace{-0.85em}
\item[--] Refactoring can extract parts of the notebook that
  are needed to compute a specified output.
\vspace{0.25em}
\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
\vspace{0.25em}
\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
  or code cleanup.
\end{itemize}

\subsection{AI assistants}

\bibliography{paper}
\end{document}

### 2018-03-15 report/paper.tex
fixing some typos

@@
 \item[--] AI assistants can access the dependency graph created for a
-notebook. This allows the them to access the transformation history, which is the code
-that has already been run during the analysis. The gives assistants information about the
+notebook. This allows them to access the transformation history, which is the code
+that has already been run during the analysis. This gives assistants information about the
 context such as transformations that have already been performed.
@@
 is annotated with a probabilistically inferred type will be significantly larger than a data
-frame contianing just raw numbers. The content negotiation protocol allows the client to 
+frame containing just raw numbers. The content negotiation protocol allows the client to 
 request a data frame with just raw numbers, significantly reducing the size.

### 2018-03-16 report/paper.tex
update wrt typos

@@
 \setcopyright{none}
-\bibliographystyle{ACM-Reference-Format}
 
@@
 
-\author{AIDA Team, The Alan Turing Institute}
+\author{AIDA Team, The Alan Turing Institute \\ 16 March 2018}
+
 %\affiliation{
@@
 \paragraph{Big data is big,}
-so the analyst doesn't understand it all.*
-Even if a data set is small enough fits on one computer, it's still too large to fit in
+so the analyst doesn't understand it all.
+Even if a data set is small enough to fit on one computer, it is still too large to fit in
 an analyst's working memory.
@@
 The data store provides a way for persistently storing data and enables communication between
-individual Wrattler components. Data store keeps external data files imported into Wrattler,
+individual Wrattler components. The data store keeps external data files that have been
+imported into Wrattler,
 as well as computed data frames (and, potentially, other data structures). It is immutable
@@
 \item[--] Columns can be annotated with (and stored as) primitive data type such as date or floating-point number.
-\vspace{0.25em}
 \item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or address.
-\vspace{-0.85em}
 \item[--] Columns, rows and cells of the data frame can be annotated with other metadata such as provenance.
@@
 \textit{\small{Finally, we use JavaScript to create a histogram of upload speeds:}}\\[0.3em]
-\quad\ident{vega}.\ident{hsit}(\ident{bbClean}, \str{Upload speed})
+\quad\ident{vega}.\ident{hist}(\ident{bbClean}, \str{Upload speed})
 \end{array}
@@
 by Wrattler and is the graph has a node for each call and constant. Hexagons represent
-data frames shared by cell and stored in the data store.}}
+data frames shared by multiple cells and stored in the data store.}}
 \label{fig:deps}
@@
 given to the AutoML challenge \cite{guyon_review_2016}, which focuses
-on automatically choosing a classification
+on automatically choosing classification
 methods, feature representations and hyperparameters
@@
 exploring the predictions of the model on validation data. Some early
-work in this direction is \cite{saleema}.  There is a rich and growing
+work in this direction is that of \citet{ren2016squares}.  There is a rich and growing
 literature on interpreting and explaining the predictions of a model
-\cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.  Two
+\cite{lipton:mythos,doshi-velez17,Ribeiro2016lime,darksight}.  Two
 potential directions from are debugging by using "training set
@@
 training set caused me to make that error? This might build on the
-work of \cite{percy}.  Another potential avenue is explaining
+work of \citet{koh2017understanding}.  Another potential avenue is explaining
 differences between predictions: why was instance 42 treated
@@
 exposed by the data store, but we do not discuss the technology used
-for data storage -- this can be decided during the development.
+for data storage; this is to be decided during the development.
 
@@
 metadata. Such metadata can communicate information between multiple
-AI assistants. For example, first AI assistant can infer that a pair
-of columns represent GPS lat/lon coordinates and produce a data frame
+AI assistants. For example, suppose that the first AI assistant  infers that a pair
+of columns represent GPS latitude/longitude coordinates and produces a data frame
 with annotations specifying this information. A second AI assistant
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{paper}

### 2018-03-18 tapp/paper.tex
Finishing TaPP draft

@@
 \documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\usepackage{flushend}
+
 \acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
@@
 
-In standard notebook systems, such as Jupyter, the state and execution is handled by a 
+In standard notebook systems, such as Jupyter \cite{jupyter}, the state and execution is handled by a 
 \emph{kernel}. The notebook running in a browser sends commands to the kernel to evaluate 
@@
 as a server, but also languages that can be fully evaluated in the browser. To illustrate this
-feature, we integrated Wrattler with TheGamma [X], a simple browser-based language for data
+feature, we integrated Wrattler with TheGamma \cite{thegamma}, a simple browser-based language for data
 exploration.
 
-As an example, we use a data set with broadband speed published by the UK government [X].
+As an example, we use a data set with broadband speed published by the UK government.
 The following TheGamma script loads the raw CSV file, groups rows by whether they represent
@@
 %
-Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} [X] 
+Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} \cite{providers} 
 and the user typically interacts by selecting one of generated members through an auto-complete
@@
 \hspace{-6em}\begin{array}{l}
-\kvd{procedure}~\ident{bind}(\textit{cache}, \texit{syn})~=\\
+\kvd{procedure}~\ident{bind}(\textit{cache}, \textit{syn})~=\\
 \quad \kvd{let}~h = \{ \ident{hash}(\ident{kind}(\textit{syn})) \} \\  
@@
 
-As an example, \emph{datadiff} is an AI assistant that reads two specified data frames from the
+As an example, datadiff \cite{datadiff} is an AI assistant that reads two specified data frames from the
 data store and suggests a script that transforms the data in the second data frame to match the
@@
 
-\section{Results}
+\section{Properties of Wrattler}
 \label{sec:results}
 
-\paragraph{Properties.}
-The dependency graph enables several features that are difficult for most notebook systems:
-%
-\begin{itemize}
-\item[--] When code changes, Wrattler only needs to recompute a small part of the graph.
-  This makes it possible to provide live previews, especially for simple data analytical DSLs.
-\item[--] Refactoring can extract parts of the notebook that
-  are needed to compute a specified output.
-\vspace{0.25em}
-\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
-\vspace{0.25em}
-\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
-  or code cleanup.
-\end{itemize}
+The Wrattler architecture outlined in Section~\ref{sec:arch} leads to a number of properties that
+are difficult to obtain with traditional notebook systems. In this section, we discuss the 
+properties and briefly discuss our prototype implementation.
 
-Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
-Once it is interactively constructed, it can be translated to multiple supported languages
-such as R, Python or JavaScript.
+\subsection{Reproducible, live and smart}
+
+The two most important aspects of the Wrattler architecture are that it separates the state from 
+the language runtime (using a data store) and that it keeps a dependency graph based on the current
+notebook source code (on the client). This enables a number of properties.
+
+\paragraph{Reproducibility.} The evaluation outputs displayed in Wrattler notebook always reflect
+the current source code. When code changes, Wrattler updates the dependency graph and hides 
+invalidated visualizations. Because the data store caches earlier results, it is always possible
+to go back without re-evaluating the whole notebook.
+
+\paragraph{Provenance.} The graph allows us to track data provenance. For code written in TheGamma
+script, we can track provenance at the level of individual variables and operations; for R, Python
+or JavaScript, we can currently only track provenance at cell level. This also enables implementing
+refactoring of notebooks, for example, to extract only code necessary to produce a given visualization.
+
+\paragraph{Interactivity.} Finally, the use of dependency graph makes it possible to update 
+displayed results more efficiently, because only values for new nodes in the graph need to be
+calculated. The fine-grained structure of the dependency graph for TheGamma script makes it 
+possible to often show live previews instantly.
+
+\paragraph{Polyglot.} Sharing the state via data store makes it possible to combine multiple
+langauge runtimes, as long as they support sharing data via data frames. In our prototype, this
+includes R, JavaScript and TheGamma script, but the extensibility model allows adding further
+languages.
+
+\paragraph{Smart and explainable.} The data store makes it possible to build third-party components, 
+that analyze data in the notebook and provide automatic suggestion, such as how to merge entities
+across multiple data sets. Furthermore, the data store does not allow direct modification of the 
+data, so such tools result in code that explains the operation and can be reviewed by the data
+analyst.
+
+\subsection{Wrattler prototype}
+
+A prototype implementation of Wrattler is available on GitHub (\url{http://github.com/wrattler}).
+The prototype implements language runtimes for R, JavaScript and TheGamma script 
+(Section~\ref{sec:comp-gamma}). It builds a dependency graph (Section~\ref{sec:comp-deps})
+and uses it to evaluate results of cells. The data store (Section~\ref{sec:comp-data}) 
+stores data in Microsoft Azure in JSON format. Support for meta-data annotations and big data
+is not yet implemented. The protoype comes with one third-party tool -- the datadiff AI 
+assistant mentioned above (Section~\ref{sec:comp-data}). Figure~\ref{fig:proto} shows the 
+first cell of a sample Wrattler notebook.
+
+\begin{figure}
+\vspace{-0.5em}
+\includegraphics[scale=0.15]{screen.png}
+\vspace{-0.5em}
+\caption{\small{TheGamma script that downloads and aggregates UK government data, running in 
+  Wrattler notebook with a live preview.}}
+\label{fig:proto}
+\end{figure}
 
@@
 
-dataflow notebooks
+The work in this paper directly follows the work on IPython and Jupyter notebook systems
+\cite{ipython}\cite{jupyter}. Wrattler shares many properties with those and aims to address some
+of their limitations. To address reproducibility, some Jupyter extensions and systems such
+as R markdown \cite{rmarkdown} lock cells after evaluation. Dataflow notebooks 
+\cite{dataflow} are more sophisticated and attach unique hashes to cell evaluations.
+Scientific workflow systems \cite{taverna,kepler} manage evaluation of workflows over a 
+dependency graph similarly to Wrattler, but often allow editing it directly via a GUI, rather
+than through code in a notebook.
 
-\newpage
-~
-\newpage
+Our method of dependency grpah construction is inspired by Roslyn \cite{roslyn} and extends our
+earlier work on TheGamma \cite{thegamma}. It is simiar to methods used in live programming 
+languages \cite{live,subtext}, as well as incremental compilation \cite{incremental} and
+partial evaluation \cite{partial}. Wrattler adapts those methods to a notebook-like data
+science environment.
 
-Data science is an iterative, exploratory process that requires a collaboration between a 
-computer system and a human. A computer can provide advice 
-based on statistical analysis of the data and discover hidden structures or corner cases, but only
-a human can decide what those mean.
-A data science environment of the future thus needs to support an interaction model that 
-keeps the human in the loop, allows an efficient human-computer interaction 
-and can be extended with new AI asistants that provide advice about data.
-In this report, we present Wrattler, a notebook system that is:
+\section{Summary}
 
-\paragraph{Interactive.}
-Wrattler enables an efficient interaction by bringing computation closer to the human.
-Notebooks run in the browser, cache partial results of computations and provide previews
-of script results on-the-fly during development.
+This paper presents early work on Wrattler -- a new notebook system for data science that makes
+notebooks reproducible, live and smart. The properties of Wrattler follow from two changes to 
+the standard architecture of notebook systems.
 
-\paragraph{Reproducible.} 
-Wrattler separates the task of running scripts from the task of managing state.
-A data store tracks the provenance and semantics of data, supports versioning and keeps 
-the history, making the data analyses fully reproducible.
-
-\paragraph{Polyglot.}
-Multiple languages can be used in a single notebook and share data via the data store.
-Analysts can use R and Python, but also interactive languages for data exploration
-that run in the browser and provide live previews.
-
-\paragraph{Smart.}
-Wrattler serves as a platform for AI assistants that use machine learning to provide suggestions 
-about data. Such AI assistants connect to the data store to infer types and meaning of data, provide 
-help with data cleaning and joining, but also help data exploration by finding typical and atypical 
-data points and automatically visualizing data.
-
-\section{Wrattler components}
-
-Wrattler consists of a notebook user interface running in the web browser, which communicates with 
-a number of server-side components including language runtimes, the data store and AI assistants. 
-In this section, we discuss the components in more detail, starting with the dependency graph
-that is maintained on the client-side, by the web browser.
-
-\subsection{Dependency graph}
-
-When opening a notebook, Wrattler parses the source of the notebook (consisting of text cells and 
-code cells) and constructs a dependency graph, which serves as a runtime representation of the
-notebook.
-
-The structure is ilustrated in Figure~\ref{fig:deps}. Top-level nodes (squares) represent
-individual notebook cells. Code (circles) is represented either as a single node (R, Python) or as 
-a sub-graph with node for each operation (DSLs understood by Wrattler). Any node of a cell can 
-depend on a data frame (hexagons) exported by an earlier cell. 
-When code in a cell changes, Wrattler updates the dependency graph, keeping existing
-nodes for entities that were unaffected by the change.
-
-The dependency graph enables several features that are difficult for most notebook systems:
-%
-\begin{itemize}
-\item[--] When code changes, Wrattler only needs to recompute small part of the graph.
-  This makes it possible to provide live previews, especially for simple data analytical DSLs.
-\vspace{-0.85em}
-\item[--] Refactoring can extract parts of the notebook that
-  are needed to compute a specified output.
-\vspace{0.25em}
-\item[--] Refactoring can translate nodes of an analytical DSLs (generated by an AI assistant) into R or Python.
-\vspace{0.25em}
-\item[--] The graph can be used for other analyses and refactorings, such as provenance tracking
-  or code cleanup.
-\end{itemize}
-
-\subsection{AI assistants}
+First, Wrattler separates the state management from code execution. This allows provenance 
+tracking, polyglot notebooks and integration of third-party tools that can work directly with
+the data store. Second, Wrattler keeps a dependency graph on the client (web browser) and uses 
+it to control evaluation. This guarantees reproducibility and allows faster feedback during
+development.
 

### 2018-03-26 tapp/paper.tex
Adding submitted version

@@
 \begin{document}
-\title{Wrattler: \textnormal{Making notebooks reproducible, live and smart}}
+\title{Wrattler: \textnormal{Reproducible, live and polyglot notebooks}}
 
@@
   \institution{The Alan Turing Institute}
-  \city{London}
-  \country{UK}
 }
@@
 \affiliation{
-  \institution{University of Edinburgh}
-  \city{Edinburgh}
-  \country{UK}
+  \institution{The University of Edinburgh}
+}
+\additionalaffiliation{
+  \institution{The Alan Turing Institute and Google}
 }
@@
 they support interactive data exploration and provide a convenient way of interleaving code, 
-comments and visualizations. However, notebooks also suffer from reproducibility issues and make
-versioning and provenance tracking difficult.
+comments and visualizations. However, most notebook systems use an architecture that makes
+reproducibility and versioning difficult and limits the interaction model.
 
-In this paper, we present Wrattler, a new notebook system with an architecture that addresses the
-above issues. Wrattler stores all state in a data store and separates state management from 
-script evaluation. This makes it possible to support versioning, guarantee reproducibility, track 
-data provenance, but also allow richer forms of interactivity and integrate AI tools that help
-automate routine data wrangling tasks.
+In this paper, we present Wrattler, a new notebook system built around provenance that addresses
+the above issues. Wrattler separates state management from script evaluation and controls the 
+evaluation using a dependency graph maintained in the web browser. This allows richer forms of 
+interactivity, an efficient evaluation through caching, guarantees reproducibility and makes it 
+possible to support versioning.
 \end{abstract}
@@
 \section{Introduction}
-Data science is an iterative, exploratory process that requires a collaboration between a computer
-system and a human. Notebook systems support this interaction model by making it easy to run 
-snippets of code and see results without leaving the notebook. However, typical notebook systems,
-such as Jupyter suffer from a number of problems:
+Notebooks \cite{ipython,jupyter} are literate programming \cite{literate} develoment
+environments that allow data analyst to combine text, code and code outputs. Notebooks make
+it easy to run code and see results. To further aid reproducible, iterative and exploratory data 
+science, notebook systems should also support:
 
-\paragraph{Limited reproducibility.} Cells can be executed out-of-order or modified without re-evaluating 
-the notebook. This means that running all cells from scratch may give different results than those 
-originally seen in the notebook.
+\paragraph{Richer interaction model.}
+Web browsers are increasingly powerful and allow moving parts of data exploration to the client-side. 
+Simple code changes should be evaluated efficiently using a cache and notebooks
+should give live previews when writing code to perform simple data exploration.
 
-\paragraph{Opaque state management.} All state is managed by the execution engine, or \emph{kernel}.
-This makes the state invisible to tools such as version control systems and makes it hard to 
-track provenance in a notebook.
+\paragraph{Transparent state management.}
+The state maintained by a notebook should be transparent and accessible to external tools. This
+would allow versioning of state and development of tools to explore and manipulate the notebook state.
 
-\paragraph{Limited interaction.} Notebooks execute code at the granularity of an entire cell. 
-This means that even simple code change may trigger a long-running computation and, as a result, 
-notebooks do not always provide rapid feedback.
+\paragraph{Multiple languages and tools.} 
+A notebook should make it easy to combine multiple programming languages. A cell written in one
+language should be able to automatically access data frames defined in other languages.
 
-\paragraph{Single language.} When using a notebook, the state is managed by a single \emph{kernel}.
-This makes it difficult to combine multiple programming languages in a single notebook, or build
-third-party tools that would inspect the state and provide data analysts with hints about data.
+\paragraph{Improved reproducibility.} 
+Changing code in a cell should invalidate results that depend on data frames defined in 
+the cell. Reverting a change should immediately revert the result to the previous one
+and show it immediately using a cache.
 
-\vspace{1em}
+\newpage
+~
+
 \noindent
-In this paper, we present Wrattler, a new notebook system that addresses the above issues.
-This is made possible by two key aspects of the Wrattler architecture (Section~\ref{sec:arch}).
+In this paper, we present Wrattler, a new notebook system that supports
+the above features. We follow a line of work combining provenance tracking with notebooks 
+\cite{noworkflow,dataflow}, but rather than extending existing systems, we revisit
+two key aspects of the notebook architecture (Section~\ref{sec:arch}).
 
-First, Wrattler maintains a dependency graph between cells and often also function calls inside
-a single cell (Section~\ref{sec:comp-deps}). When a cell is changed, relevant parts of a graph
-are invalidated. This guarantees reproducibility and it enables more efficient re-computation, 
-because values of unaffected nodes can be reused.
-Second, Wrattler separates state management from code execution (Section~\ref{sec:comp-data}). 
-The \emph{data store} is responsible for managing state. It handles versioning, provenance tracking 
-and enables multiple independent languages and tools to operate on the state of a notebook.
+First, Wrattler uses a \emph{dependency graph} to track provenance between cells and even function calls 
+inside individual cells (Section~\ref{sec:comp-deps}). When a cell is changed, relevant parts of a graph
+are invalidated. This guarantees reproducibility and enables more efficient re-computation.
+Second, Wrattler introduces a \emph{data store} that separates state management from code execution 
+(Section~\ref{sec:comp-data}). The data store handles versioning and simplifies the support for polyglot programming.
 
 Together, these two changes to the standard architecture of notebook systems make Wrattler notebooks
-(Section~\ref{sec:results}) reproducible (with an easy and reliable state rollback), live (with 
-efficient re-computation on change) polyglot and smart (allowing the integration of multiple 
-execution engines and third-party AI tools).
+(Section~\ref{sec:results}) polyglot, reproducible (with an easy and reliable state rollback) and live 
+(with efficient re-computation on change that enables live preview during data exploration).
 
@@
 \caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
-  Wrattler, those functions are split between data store and language runtimes. Data in the data store
-  can be also accessed by other thrd-party tools.}}
+  Wrattler, those functions are split between data store and language runtimes. Language runtimes 
+  can run on the server-side (e.g.~Python) or client-side (e.g.~TheGamma).}}
 \label{fig:arch}
@@
 \label{sec:arch}
+Standard notebook architecture consists of a \emph{notebook} and a \emph{kernel}. The kernel
+runs on a server, evaluates code snippets and maintains state they use.
+Notebook runs in a browser and sends commands to the kernel in order to evaluate 
+cells selected by the user. As illustrated in Figure~\ref{fig:arch}, Wrattler splits the 
+server functionality between two components:
 
-In standard notebook systems, such as Jupyter \cite{jupyter}, the state and execution is handled by a 
-\emph{kernel}. The notebook running in a browser sends commands to the kernel to evaluate 
-cells selected by the user. As illustrated in Figure~\ref{fig:arch}, Wrattler splits the 
-server functionality between the following components:
-
-\paragraph{Data store.} Imported external data, results of running scripts and of 
-running third-party tools are stored in the data store. The data store keeps version history
-and annotates data with metadata such as types, inferred semantics and provenance.
+\paragraph{Data store.} Imported external data and results of running scripts  
+are stored in the data store. The data store keeps version history and annotates data with 
+metadata such as types, inferred semantics and provenance information.
 
@@
 
-\paragraph{Third-party tools.} In addition to language runtimes, other tools can contribute
-to data analysis by analysing data in the data store. In Wrattler, this includes AI assistants 
-that help, for example, with data cleaning, type inference and data extraction from raw 
-input data.
-
 \paragraph{Notebook.} The notebook is displayed in a web browser and orchestrates 
 all other components. The browser builds a dependency graph between cells or individual 
-expressions in the cells. It calls language runtimes to evaluate code that has changed,
-third-party tools to provide hints and reads data from the data store to display results.
-
-\begin{figure}
-\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
-\caption{\small{Dependency graph of a sample notebook with three cells. The cells are written
-in Python, Wrattler script and JavaScript. The browser builds a fine-grained dependency graph
-for Wrattler script that can be parsed and analysed in the browser.}}
-\label{fig:deps}
-\vspace{-0.5em}
-\end{figure}
+calls. It invokes language runtimes to evaluate code that has changed,
+and reads data from the data store to display results.
 
@@
 Wrattler consists of a notebook user interface running in the web browser, which communicates 
-with a number of server-side components including language runtimes, the data store and third-party 
-tools. In this section, we discuss the components in more detail.
+with the data store and multiple server-side language runtimes. We also provide an integration 
+with a simple data exploration language TheGamma \cite{thegamma} that has a fully client-side language
+runtime.
 
@@
 The Wrattler architecture supports languages that require running a separate language runtime
-as a server, but also languages that can be fully evaluated in the browser. To illustrate this
-feature, we integrated Wrattler with TheGamma \cite{thegamma}, a simple browser-based language for data
-exploration.
+as a server, but also languages that can be parsed and evaluated in the browser. To illustrate this, 
+we integrated Wrattler with TheGamma~\cite{thegamma}, a simple browser-based language 
+for data exploration.
 
-As an example, we use a data set with broadband speed published by the UK government.
-The following TheGamma script loads the raw CSV file, groups rows by whether they represent
-rural or urban area and then calculates average download speed for both of the kinds of areas:
-%
+As an example, we use a data set with broadband speed published by the UK government \cite{ofcom}.
+The following script written using TheGamma loads the online CSV file, groups rows by whether they represent
+rural or urban area and then calculates average download speed for both area kinds:
+
+\noindent
 \begin{equation*}
+\hspace{-3em}
 \begin{array}{l}
@@
 \end{equation*}
-%
-Here, identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} \cite{providers} 
+
+\noindent
+Identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} \cite{providers} 
 and the user typically interacts by selecting one of generated members through an auto-complete
@@
 code cells) and constructs a dependency graph, which serves as a runtime representation of the
-notebook.
+notebook. The parsing of code cells is done by language runtimes that parse and statically analyse 
+code and constructs parts of the dependency graph.
+
+Wrattler uses a \emph{data frame} as a primary format for passing data between cells.
+Data frames represent tables with multiple (named) columns and (anonymous) rows. Data frames
+have a first-class support in most data science languages.
+
+\begin{figure}
+\vspace{-0.5em}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+\vspace{-0.5em}
+\caption{\small{Dependency graph of a sample notebook with three cells. The cells are written
+in Python, Wrattler script and JavaScript. The browser builds a fine-grained dependency graph
+for Wrattler script that can be parsed and analysed in the browser.}}
+\label{fig:deps}
+\vspace{-1em}
+\end{figure}
 
@@
 
-Wrattler creates at least two nodes for each cell, representing the cell as a whole and its source
+Wrattler creates two nodes for each cell, representing the cell as a whole and its source
 code. In addition, it creates a node for each data frame exported by a cell (e.g.~the rightmost
@@
 
-For code snippets that can be analysed in 
-the browser (such as TheGamma cells), Wrattler creates a more fine-grained dependency graph with
-a node for each operation (such as member access or a function call). This is the case in the 
-second cell in Figure~\ref{fig:deps} and it allows more efficient live update of results when 
-code changes as we only need to re-evaluate parts of code in a cell. We briefly outline how the
-dependency graph is constructed and used.
+The dependency graph is updated after each code change and so it needs to be constructed efficiently 
+in the browser. Language runtime that run on the client-side create a more fine-grained 
+dependency graph with a node for each operation (such as member access or a function call). This 
+is the case in the  second cell in Figure~\ref{fig:deps}. This allows live update of results when 
+code changes as we only need to re-evaluate small parts of code in a cell.
 
 \paragraph{Dependency graph construction.} 
-The dependency graph is reconstructed after every change in any of the cells. This is an efficient
-process, because the changes are typically small. Wrattler first parses each cell, producing a
-single syntactic element for Python, JavaScript and R cells (representing the entire source code)
-and a syntax tree for TheGamma (representing individual member accesses and other constructs).
-The syntax tree for the entire notebook is then a list of elements produced for each cell.
+The dependency graph is updated after every code change. Wrattler first parses each 
+cell, producing a single syntactic element for Python and JavaScript cells (representing the entire 
+source code) and a syntax tree for TheGamma (representing member accesses and other 
+constructs). The syntax tree for the entire notebook is then a list of elements produced for each 
+cell.
 
 After parsing, Wrattler walks over the syntax tree recursively and binds a dependency graph node
-to each syntactic element in the tree. The \emph{antecedents} of a node are the nodes that it
-depends on. This typically includes inputs for an operation or instance on which an operation
-is invoked. The binding procedure is decribed in Figure~\ref{fig:bind}.
+to each syntactic element in the tree using a process decribed in Figure~\ref{fig:bind}. 
+The \emph{antecedents} of a node are the nodes that it depends on. This typically includes 
+inputs for an operation or instance on which a member access is performed.
 
@@
 of the syntactic element that the node represents. An important property of the binding process
-is that, if there is no change in antecedents of a node, binding will return the same node
-as before. This means that if we evaluate a value for a given node in a dependency graph and
-attach it to a node, the value will be cached and reused.
+(Figure~\ref{fig:bind}) is that, if there is no change in antecedents of a node, binding will 
+return the same node as before. This means that if we evaluate a value for a given node in a 
+dependency graph and attach it to a node, the value will be cached and reused.
 
-Wrattler does not automatically re-evalaute the entire dependency graph, but allows the user to
-trigger re-evaluation. This provides a user experience familiar to other notebook systems. 
-However, the displayed results and visualizations always reflect the current source code in the
-notebook. When the evaluation of a cell is requested, Wrattler recursively evaluates all the 
-antecedents of the node and then evaluates the value of the node.
-
-The evaluation is delegated to a language
-runtime associated with the language of the node and it can be done in several ways:
+Wrattler re-evaluates parts of the dependency graph on demand and the displayed results and 
+visualizations always reflect the current source code in the notebook. When the evaluation 
+of a cell is requested, Wrattler recursively evaluates all the antecedents of the node and 
+then evaluates the value of the node. The evaluation is delegated to a language
+runtime associated with the language of the node:
 %
 \begin{enumerate}
-\item For Python or R nodes, the language runtime ensures that the values of the dependencies
-  are stored in the data store and sends the source code, together with its dependencies, to a
-  server that retrieves the dependencies and evaluates the code.
+\item For Python nodes, the language runtime sends the source code, together with its 
+  dependencies, to a server that retrieves the dependencies and evaluates the code.
 \item For TheGamma and JavaScript nodes, the language runtime collects values of the 
@@
 is done by recursively walking over the dependency graph and annotating the nodes with their 
-type. This step is only relevant for cells written in statically typed languages and provide
-quicker feedback, because it does not need to evaluate the cells.
+type. This step is only relevant for cells written in statically typed languages, but it provides
+a quick feedback, because it does not need to evaluate the code.
 
@@
 The data store enables communication between individual Wrattler components and provides a way for 
-persistently stroing data. The data stored in the data store is associated with the hash 
-produced by the binding process outlined in Figure~\ref{fig:bind} and is immutable. When the
-notebook changes, new nodes with new hashes are created and so the new results are appended,
-rather than overwriting existing state.
+persistently stroing input data. It stores raw input data, data frames and multiple versions of the 
+dependency graph. Data frames stored in the data store are associated with the hash produced by the 
+binding process outlined in Figure~\ref{fig:bind} and are immutable. When the notebook changes, 
+new nodes with new hashes are created. New data frames are appended, rather than overwriting 
+existing state.
 
-The data store in Wrattler (currently) supports two data formats -- external data files imported
-into Wrattler notebooks (such as downloaded web pages) and data frames. The latter are used for
-communication between multiple cells of a Wrattler notebook. A cell that exports a data frame
-so that it can be used by subsequent cells needs to store the data frame into a data store (the
+External input data files imported into Wrattler notebooks (such as downloaded web pages) are 
+stored as binary blobs. Data frames are currently stored in JSON format (as an array of records),
+but we intend to use a suitable database system in the future. Data frames are used for
+communication between multiple cells of a Wrattler notebook. To make a data frame available to 
+subsequent cells, a cell can export and store a data frame into the data store (the
 key is the hash of the node representing the data frame).
 
-The data store additionally also supports a mechanism for annotating data frames with additional 
-semantic information. Columns can be annotated with (and stored as) primitive data type such as 
-date or floating-point number. Columns can also be annotated with semantic annotation that indicates
+The data store also supports a mechanism for annotating data frames with 
+semantic information. Columns can be annotated with primitive data type such as 
+date or floating-point number. Columns can be annotated with semantic annotation that indicates
 the meaning of the column -- for example, address or longitude and latitude. Finally, columns,
 rows and individual cells of the data frame can be annotated with other metadata such as their
-provenance.
+data source.
 
@@
 \subsection{Reproducible, live and smart}
-
 The two most important aspects of the Wrattler architecture are that it separates the state from 
 the language runtime (using a data store) and that it keeps a dependency graph based on the current
-notebook source code (on the client). This enables a number of properties.
+notebook source code (on the client). The provenance information that is available thanks to this
+arhcitecture enable a number of properties.
 
@@
 
-\paragraph{Provenance.} The graph allows us to track data provenance. For code written in TheGamma
-script, we can track provenance at the level of individual variables and operations; for R, Python
-or JavaScript, we can currently only track provenance at cell level. This also enables implementing
-refactoring of notebooks, for example, to extract only code necessary to produce a given visualization.
+\paragraph{Refactoring.} The dependency graph allows us to implement notebook refactoring. 
+For example, it is possible to extract only code necessary to produce a given visualization.
+For code written in TheGamma, this extracts individual operations; for Python or JavaScript,
+we can currently extract code at cell-level granularity.
 
-\paragraph{Interactivity.} Finally, the use of dependency graph makes it possible to update 
-displayed results more efficiently, because only values for new nodes in the graph need to be
-calculated. The fine-grained structure of the dependency graph for TheGamma script makes it 
-possible to often show live previews instantly.
+\paragraph{Live previews.} The dependency graph makes it possible to give 
+live previews during development. When code changes, only values for new nodes in the graph 
+need to be calculated. The fine-grained structure of the dependency graph for TheGamma 
+makes it possible to update previews instantly.
 
@@
 
-\paragraph{Smart and explainable.} The data store makes it possible to build third-party components, 
-that analyze data in the notebook and provide automatic suggestion, such as how to merge entities
-across multiple data sets. Furthermore, the data store does not allow direct modification of the 
-data, so such tools result in code that explains the operation and can be reviewed by the data
-analyst.
-
 \subsection{Wrattler prototype}
 
-A prototype implementation of Wrattler is available on GitHub (\url{http://github.com/wrattler}).
-The prototype implements language runtimes for R, JavaScript and TheGamma script 
+A prototype implementation of the Wrattler system is available on GitHub (\url{http://github.com/wrattler}).
+The prototype implements language runtimes for JavaScript, TheGamma script and R 
 (Section~\ref{sec:comp-gamma}). It builds a dependency graph (Section~\ref{sec:comp-deps})
-and uses it to evaluate results of cells. The data store (Section~\ref{sec:comp-data}) 
+and uses it to evaluate results of cells.
+
+The data store (Section~\ref{sec:comp-data}) 
 stores data in Microsoft Azure in JSON format. Support for meta-data annotations and big data
@@
 assistant mentioned above (Section~\ref{sec:comp-data}). Figure~\ref{fig:proto} shows the 
-first cell of a sample Wrattler notebook.
+first cell of a sample Wrattler notebook analysing the UK broadband data.
 
 \begin{figure}
-\vspace{-0.5em}
 \includegraphics[scale=0.15]{screen.png}
-\vspace{-0.5em}
 \caption{\small{TheGamma script that downloads and aggregates UK government data, running in 
@@
 
-The work in this paper directly follows the work on IPython and Jupyter notebook systems
-\cite{ipython}\cite{jupyter}. Wrattler shares many properties with those and aims to address some
+The work in this paper directly follows the work on IPython and Jupyter systems
+\cite{ipython,jupyter}. Wrattler shares many properties with those and aims to address some
 of their limitations. To address reproducibility, some Jupyter extensions and systems such
-as R markdown \cite{rmarkdown} lock cells after evaluation. Dataflow notebooks 
-\cite{dataflow} are more sophisticated and attach unique hashes to cell evaluations.
+as R markdown \cite{rmarkdown} lock cells after evaluation.
+
+Dataflow notebooks \cite{dataflow} are more sophisticated and attach unique hashes to cell 
+evaluations. This allows the user to refer to dependencies explicitly and, in effect, construct
+a dependecy graph similar to ours manually.
 Scientific workflow systems \cite{taverna,kepler} manage evaluation of workflows over a 
@@
 than through code in a notebook.
+The noWorkflow project \cite{noworkflow} links the two approaches by instrumenting Jupyter 
+kernel with a mechanism for capturing provenance based on light-weight code annotations.
 
@@
 This paper presents early work on Wrattler -- a new notebook system for data science that makes
-notebooks reproducible, live and smart. The properties of Wrattler follow from two changes to 
-the standard architecture of notebook systems.
+notebooks reproducible, live and polyglot. The properties of Wrattler are enabled by provenance
+information that is maintained thanks to two changes to the standard architecture of notebook 
+systems.
 
-First, Wrattler separates the state management from code execution. This allows provenance 
-tracking, polyglot notebooks and integration of third-party tools that can work directly with
+First, Wrattler separates the state management from code execution. This allows versioning,
+polyglot notebooks and integration of third-party tools that can work directly with
 the data store. Second, Wrattler keeps a dependency graph on the client (web browser) and uses 
@@
 
+\begin{acks}
+We thank to our colleagues from the AIDA (Artificial Intelligence for data analytics) project
+Chris Williams, Zoubin Ghahramani and Ian Horrocks and also numerous attendees of a recent 
+AIDA workshop in The Alan Turing Institute. This work was in part supported by The Alan Turing 
+Institute under the EPSRC grant EP/N510129/1.
+\end{acks}
+
 \bibliography{paper}

### 2018-03-27 tapp/paper.tex
Adding Vizer reference

@@
 \begin{figure}
+\vspace{-0.5em}
 \includegraphics[scale=0.15]{screen.png}
@@
 \label{fig:proto}
+\vspace{-0.5em}
 \end{figure}
@@
 
-Dataflow notebooks \cite{dataflow} are more sophisticated and attach unique hashes to cell 
+Dataflow notebooks \cite{dataflow} are attach unique hashes to cell 
 evaluations. This allows the user to refer to dependencies explicitly and, in effect, construct
@@
 than through code in a notebook.
+
 The noWorkflow project \cite{noworkflow} links the two approaches by instrumenting Jupyter 
-kernel with a mechanism for capturing provenance based on light-weight code annotations.
+kernel with a mechanism for capturing provenance based on light-weight annotations.
+Vizer \cite{vizer} focuses on integrating notebooks with spreadsheet-like interface.
+It internally uses a data store component similar to ours, but does not keep dependency graph
+on the client.
 
-Our method of dependency grpah construction is inspired by Roslyn \cite{roslyn} and extends our
-earlier work on TheGamma \cite{thegamma}. It is simiar to methods used in live programming 
-languages \cite{live,subtext}, as well as incremental compilation \cite{incremental} and
-partial evaluation \cite{partial}. Wrattler adapts those methods to a notebook-like data
-science environment.
+Our dependency grpah construction is inspired by Roslyn \cite{roslyn} and extends an
+earlier work on TheGamma \cite{livegamma}. It is simiar to methods used in live programming 
+languages \cite{live,subtext}, incremental compilation \cite{incremental} and
+partial evaluation \cite{partial}. Wrattler adapts those methods to a notebook environment.
 
@@
 We thank to our colleagues from the AIDA (Artificial Intelligence for data analytics) project
-Chris Williams, Zoubin Ghahramani and Ian Horrocks and also numerous attendees of a recent 
-AIDA workshop in The Alan Turing Institute. This work was in part supported by The Alan Turing 
+Chris Williams, Zoubin Ghahramani and Ian Horrocks and attendees of a recent 
+AIDA workshop. This work was in part supported by The Alan Turing 
 Institute under the EPSRC grant EP/N510129/1.

### 2018-06-05 tapp/paper.tex
Camera ready version of TaPP

@@
 
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+% What is the type of a cell in Python? Only the result? You say exports must 
+% be explicitly defined - so the type of the cell is the type of the export?
+
+% The authors say (page 2), "The data store keeps version
+% history and annotates data with metdata such as types, inferred
+% semantics and provenance information."  How these semantics and
+% provenance information are inferred and represented, however, is
+% not elaborated on (but only very briefly mentioned in Section 3.3,
+% bottom of page 3).  The role of provenance thus remains unclear to me.
+
+% How does Wrattler compare to other notebook systems in terms of storage and compute overhead, 
+% since a) extra data needs to be maintained, and 
+%  b) the state and the runtime are now separated? It would be great to include some (preliminary) numbers in this regard.
+
+\acmConference[TaPP 2018]{10th USENIX Workshop on The Theory and Practice of Provenance}{July 11--12, 2018}{London, UK}
 \acmYear{2018}
@@
 \startPage{1}
-\setcopyright{none}
+\setcopyright{rightsretained}
 
@@
 \affiliation{
+  \institution{University of Kent}
+}
+\affiliation{
   \institution{The Alan Turing Institute}
@@
 \email{tomas@tomasp.net}
-\author{Charles Sutton}
-\affiliation{
-  \institution{The University of Edinburgh}
-}
-\additionalaffiliation{
-  \institution{The Alan Turing Institute and Google}
-}
-\email{csutton@inf.ed.ac.uk}
 \author{James Geddes}
@@
 \email{jgeddes@turing.ac.uk}
+\author{Charles Sutton}
+\affiliation{
+  \institution{The University of Edinburgh}
+}
+\affiliation{
+  \institution{The Alan Turing Institute and Google}
+}
+\email{csutton@inf.ed.ac.uk}
 
@@
 \begin{abstract}
-Notebook systems such as Jupyter became a popular programming environment for data science, because 
+Notebooks such as Jupyter became a popular environment for data science, because 
 they support interactive data exploration and provide a convenient way of interleaving code, 
-comments and visualizations. However, most notebook systems use an architecture that makes
-reproducibility and versioning difficult and limits the interaction model.
+comments and visualizations. Alas, most notebook systems use an architecture that leads to a 
+limited model of interaction and makes reproducibility and versioning difficult.
 
@@
 \end{abstract}
-\keywords{notebook, dependency graph, live coding, AI}
+%\keywords{notebook, dependency graph, live coding}
 \maketitle
@@
 \section{Introduction}
-Notebooks \cite{ipython,jupyter} are literate programming \cite{literate} develoment
-environments that allow data analyst to combine text, code and code outputs. Notebooks make
-it easy to run code and see results. To further aid reproducible, iterative and exploratory data 
-science, notebook systems should also support:
+Notebooks \cite{ipython,jupyter} are literate programming \cite{literate} systems
+that allow interleaving text, code and outputs. To aid reproducible, 
+exploratory data science, notebook systems should provide:
 
+\vspace{-0.3em}
 \paragraph{Richer interaction model.}
 Web browsers are increasingly powerful and allow moving parts of data exploration to the client-side. 
-Simple code changes should be evaluated efficiently using a cache and notebooks
-should give live previews when writing code to perform simple data exploration.
+Notebooks should leverage this and give live previews when writing code to perform simple data exploration.
 
+\vspace{-0.3em}
 \paragraph{Transparent state management.}
 The state maintained by a notebook should be transparent and accessible to external tools. This
-would allow versioning of state and development of tools to explore and manipulate the notebook state.
+would allow versioning of state and development of tools that provide hints based on the notebook state.
 
+\vspace{-0.3em}
 \paragraph{Multiple languages and tools.} 
@@
 
+\vspace{-0.3em}
 \paragraph{Improved reproducibility.} 
@@
 
-\newpage
-~
+\vspace{0.5em}
+\noindent
+Supporting these is a challenge that combines several research areas. We need programming 
+language techniques to efficiently update live previews during editing, provenance methods to 
+track data dependencies and data representation that can be shared across langauges.
 
-\noindent
-In this paper, we present Wrattler, a new notebook system that supports
-the above features. We follow a line of work combining provenance tracking with notebooks 
-\cite{noworkflow,dataflow}, but rather than extending existing systems, we revisit
-two key aspects of the notebook architecture (Section~\ref{sec:arch}).
-
+Wrattler is a new notebook system that supports the above features. We follow a line of work combining 
+provenance tracking with notebooks \cite{noworkflow,dataflow}, but rather than extending existing 
+systems, we revisit two core aspects of the notebook architecture (Section~\ref{sec:arch}). 
 First, Wrattler uses a \emph{dependency graph} to track provenance between cells and even function calls 
@@
 \begin{figure}[b]
-\vspace{-1em}
+\vspace{-2em}
 \includegraphics[scale=0.6]{diagram.pdf}
-\vspace{-0.5em}
+\vspace{-1.5em}
 \caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
@@
 The runtimes read input data from and write results back to the data store. Wrattler supports
-language runtimes that run code on the server (similar to Jupyter kernels), but also language
-runtimes that parse and evaluate code directly in the browser.
+language runtimes that run code on the server (similar to Jupyter), but also browser-based 
+langauge runtimes.
 
@@
 
+\begin{figure}
+\vspace{-0.5em}
+\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
+\vspace{-0.5em}
+\caption{\small{Dependency graph of a sample notebook: The first (Python) cell downloads data and 
+exports the result as a data frame; the second (TheGamma) cell performs data cleaning and the third 
+(JavaScript) cell creates a visualization. Language runtime for TheGamma runs in the browser and 
+creates a fine-grained graph (which allows an efficient live previews), while Python and JavaScript 
+runtimes create just one node for the whole source code.}}
+\label{fig:deps}
+\vspace{-1em}
+\end{figure}
+
 \section{Wrattler components}
@@
 
-Wrattler consists of a notebook user interface running in the web browser, which communicates 
-with the data store and multiple server-side language runtimes. We also provide an integration 
-with a simple data exploration language TheGamma \cite{thegamma} that has a fully client-side language
-runtime.
-
-\subsection{TheGamma script}
-\label{sec:comp-gamma}
-
-The Wrattler architecture supports languages that require running a separate language runtime
-as a server, but also languages that can be parsed and evaluated in the browser. To illustrate this, 
-we integrated Wrattler with TheGamma~\cite{thegamma}, a simple browser-based language 
-for data exploration.
-
-As an example, we use a data set with broadband speed published by the UK government \cite{ofcom}.
-The following script written using TheGamma loads the online CSV file, groups rows by whether they represent
-rural or urban area and then calculates average download speed for both area kinds:
-
-\noindent
-\begin{equation*}
-\hspace{-3em}
-\begin{array}{l}
-\ident{web.loadTable}(\str{https://.../bb2014.csv}).\ident{explore}.\\
-\quad\qident{group data}.\qident{by Urban/rural}.\\
-\qquad\qident{average Download speed (Mbit/s) 24 hrs}  
-\end{array}
-\end{equation*}
-
-\noindent
-Identifiers such as \qident{by Urban/rural} are generated by a \emph{type provider} \cite{providers} 
-and the user typically interacts by selecting one of generated members through an auto-complete
-(hence the long explicity names are acceptable). For the purpose of this paper, the most 
-important aspect is that TheGamma scripts can be parsed, checked and evaluated in the browser.
+Wrattler runs in the web browser and communicates with the data store and language runtimes
+that may run on the server or in the browser. An example of browser-based language runtime is
+TheGamma \cite{thegamma}, discussed in Section~\ref{sec:comp-gamma}.
 
@@
 
-When opening a notebook, Wrattler parses the source of the notebook (consisting of text cells and
-code cells) and constructs a dependency graph, which serves as a runtime representation of the
-notebook. The parsing of code cells is done by language runtimes that parse and statically analyse 
-code and constructs parts of the dependency graph.
+At runtime, Wrattler maintains a dependency graph that is composed from sub-graphs
+created by individual language runtimes. The graph is acyclic and a node can only depend on earlier 
+nodes. Each node has a value which may be:
+%
+\begin{itemize}
+\item A language-specific value that Wrattler does not understand. This is kept in the browser
+  and re-computed when the notebook is re-opened.
+\item A data frame. Data frame is stored in the data store and browser keeps a reference (URL) 
+  of the frame. This is understood by all language runtimes and provides a way of exchanging
+  data between multiple languages. 
+\end{itemize}
+%
+Figure~\ref{fig:deps} shows a sample dependency graph. Wrattler creates two nodes for each cell
+(representing the cell and its source code) and a node for each data frame exported by a cell 
+(e.g.~the rightmost node in the first cell). Nodes in subsequent cells may depend on data frames
+exported by earlier cells.
 
-Wrattler uses a \emph{data frame} as a primary format for passing data between cells.
-Data frames represent tables with multiple (named) columns and (anonymous) rows. Data frames
-have a first-class support in most data science languages.
-
-\begin{figure}
-\vspace{-0.5em}
-\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
-\vspace{-0.5em}
-\caption{\small{Dependency graph of a sample notebook with three cells. The cells are written
-in Python, Wrattler script and JavaScript. The browser builds a fine-grained dependency graph
-for Wrattler script that can be parsed and analysed in the browser.}}
-\label{fig:deps}
-\vspace{-1em}
-\end{figure}
-
-As an example, consider a simpler notebook with three cells. The first cell, written in Python,
-downloads data and exports the result as a data frame; the second cell is written in TheGamma script
-and performs data cleaning and the third cell creates a visualization in JavaScript. The
-resulting dependency graph is shown in Figure~\ref{fig:deps}.
-
-Wrattler creates two nodes for each cell, representing the cell as a whole and its source
-code. In addition, it creates a node for each data frame exported by a cell (e.g.~the rightmost
-node in the first cell). Code in subsequent cells can depend on data frames exported from earlier
-cell, which is captured as a dependency in the graph.
-
-The dependency graph is updated after each code change and so it needs to be constructed efficiently 
-in the browser. Language runtime that run on the client-side create a more fine-grained 
-dependency graph with a node for each operation (such as member access or a function call). This 
-is the case in the  second cell in Figure~\ref{fig:deps}. This allows live update of results when 
-code changes as we only need to re-evaluate small parts of code in a cell.
+Wrattler treats data frames in a special way. They are stored in data store and each langauge 
+runtime is responsible for loading them into a native language representation (e.g. pandas in 
+Python and array of records in JavaScript).
 
 \paragraph{Dependency graph construction.} 
-The dependency graph is updated after every code change. Wrattler first parses each 
-cell, producing a single syntactic element for Python and JavaScript cells (representing the entire 
-source code) and a syntax tree for TheGamma (representing member accesses and other 
-constructs). The syntax tree for the entire notebook is then a list of elements produced for each 
-cell.
+The dependency graph is updated after every code change. Wrattler invokes individual langauge 
+runtimes to parse each cell. Language runtimes that run in the browser (e.g.~TheGamma) produce
+a fine-grained syntax tree. The result of parsing the whole notebook is then a list of elements
+obtained for each cell.
 
-After parsing, Wrattler walks over the syntax tree recursively and binds a dependency graph node
-to each syntactic element in the tree using a process decribed in Figure~\ref{fig:bind}. 
+Wrattler then walks over the syntax tree and binds a dependency graph node
+to each syntactic element using a process decribed in Figure~\ref{fig:bind}. 
 The \emph{antecedents} of a node are the nodes that it depends on. This typically includes 
@@
 (Figure~\ref{fig:bind}) is that, if there is no change in antecedents of a node, binding will 
-return the same node as before. This means that if we evaluate a value for a given node in a 
-dependency graph and attach it to a node, the value will be cached and reused.
+return the same node as before. As a result, previously evaluated values attached to nodes
+in the graph are reused.
 
@@
 \end{enumerate}
-%
-Wrattler also uses the dependency graph for type checking. Just like evaluation, type checking
-is done by recursively walking over the dependency graph and annotating the nodes with their 
-type. This step is only relevant for cells written in statically typed languages, but it provides
-a quick feedback, because it does not need to evaluate the code.
 
@@
 The data store enables communication between individual Wrattler components and provides a way for 
-persistently stroing input data. It stores raw input data, data frames and multiple versions of the 
-dependency graph. Data frames stored in the data store are associated with the hash produced by the 
-binding process outlined in Figure~\ref{fig:bind} and are immutable. When the notebook changes, 
-new nodes with new hashes are created. New data frames are appended, rather than overwriting 
-existing state.
+persistently storing input data. Data frames stored in the data store are associated with the hash 
+produced by the binding process outlined in Figure~\ref{fig:bind} and are immutable. When the 
+notebook changes, new nodes with new hashes are created and appended to the data store. This means 
+that language runtimes can cache them and avoid fetching data from data store each time they need 
+to evaluate a code snippet.
 
-External input data files imported into Wrattler notebooks (such as downloaded web pages) are 
-stored as binary blobs. Data frames are currently stored in JSON format (as an array of records),
-but we intend to use a suitable database system in the future. Data frames are used for
-communication between multiple cells of a Wrattler notebook. To make a data frame available to 
-subsequent cells, a cell can export and store a data frame into the data store (the
-key is the hash of the node representing the data frame).
+External inputs imported into Wrattler notebooks (such as downloaded web pages) are stored as 
+binary blobs. Data frames are stored in JSON format (as an array of records),
+but we intend to use a suitable database in the future. During the binding process 
+(Section~\ref{sec:comp-deps}), a langauge runtime identifies imported and exported data frames 
+for each cell (e.g.~by static analysis of the code). Those are then represented as hashes (keys)
+referring to a location in the data store.
 
 The data store also supports a mechanism for annotating data frames with 
-semantic information. Columns can be annotated with primitive data type such as 
-date or floating-point number. Columns can be annotated with semantic annotation that indicates
-the meaning of the column -- for example, address or longitude and latitude. Finally, columns,
-rows and individual cells of the data frame can be annotated with other metadata such as their
-data source.
+semantic information. Columns can be annotated with primitive data types (date, floating-point 
+number) and semantic annotation indicating their meaning (address or longitude and latitude). 
+Columns, rows and individual cells of the data frame can also be annotated with custom metadata 
+such as their data source or accuracy.
 
@@
 \begin{equation*}
-\hspace{-6em}\begin{array}{l}
+\hspace{-3em}\begin{array}{l}
 \kvd{procedure}~\ident{bind}(\textit{cache}, \textit{syn})~=\\
-\quad \kvd{let}~h = \{ \ident{hash}(\ident{kind}(\textit{syn})) \} \\  
-\quad \qquad \cup~ \{\ident{hash}(c) \,|\, c\in\ident{antecedents}(\textit{syn}) \}\\
+\quad \kvd{let}~h = \ident{hash}(\{\ident{kind}(\textit{syn}) \}\cup \ident{antecedents}(\textit{syn}))\\
 \quad \kvd{if}~\ident{not}~\ident{contains}(\textit{cache}, h)~\kvd{then}\\
 \quad \quad \kvd{let}~n = \emph{fresh node}\\
-\quad \quad \ident{value}(n) \leftarrow \ident{Unevaluated} \\
-\quad \quad \ident{hash}(n) \leftarrow h\\
+\quad \quad \ident{value}(n), \ident{hash}(n) \leftarrow \ident{Unevaluated}, h \\
 \quad \quad \ident{set}(\textit{cache}, h, n) \\
@@
 \label{fig:bind}
+\vspace{-1em}
 \end{figure}
@@
 
-\subsection{Third-party tools}
-\label{sec:comp-ai}
+\subsection{TheGamma script}
+\label{sec:comp-gamma}
 
-The Wrattler architecture enables integration of third-party tools that analyse the notebook
-or data it uses and provide hints to the data analyst. Most notably, such tools include AI 
-assistants that help with tedious data cleaning and pre-processing tasks.
+The Wrattler architecture supports languages that can be parsed and evaluated in the browser.
+To illustrate this, we integrated Wrattler with TheGamma~\cite{thegamma}, a simple browser-based 
+language for data exploration.
 
-As an example, datadiff \cite{datadiff} is an AI assistant that reads two specified data frames from the
-data store and suggests a script that transforms the data in the second data frame to match the
-format of the first data frame. The AI assistant does not automatically transofrm the data, but
-instead, produces a new cell in TheGamma script that can be reviewed by the data analyst. 
-For example:
+The Figure~\ref{fig:gamma} shows TheGamma cell in Wrattler during editing. The example uses
+broadband speed data published by the UK government \cite{ofcom} and calculates average download
+speed in urban and rural areas, respectively. TheGamma supports a rich interactive model in two ways:
 %
-\begin{equation*}
-\begin{array}{l}
-\ident{datadiff(broadband2014, broadband2015)}\\
-\quad.\ident{drop\_column}(\str{WT\_national})\\
-\quad.\ident{drop\_column}(\str{WT\_ISP})\\
-\quad.\ident{recode\_column}(\str{URBAN}, [\num{1}, \num{2}], [\str{Urban},\str{Rural}])
-\end{array}  
-\end{equation*}
-%
-The script specifies that two columns should be dropped from the corrupted second data frame and
-one column needs to be recoded (turning $\num{1}$ and $\num{2}$ into strings
-\strf{Urban} and \strf{Rural}).
+\begin{itemize}
+\item The script is evaluated on-the-fly during editing and a live preview is shown 
+  (below the code editor).
+\item All code can be written using autocomplete that offers available members (representing
+  aggregation operations). Rather than writing code, user repeatedly selects one of the offered 
+  members (which are provided by a type provider \cite{providers} running in the browser).
+\end{itemize}
+
+For the purpose of this paper, the most important aspect of TheGamma is that scripts can be 
+parsed and evaluated in the browser. This allows more interactive style of data exploration
+without round-trips to re-evaluate modified code.
 
@@
 
-The Wrattler architecture outlined in Section~\ref{sec:arch} leads to a number of properties that
-are difficult to obtain with traditional notebook systems. In this section, we discuss the 
-properties and briefly discuss our prototype implementation.
+The Wrattler architecture outlined in Section~\ref{sec:arch} allowed us to develop a prototype 
+system with a number of properties that are difficult to obtain with traditional notebooks.
 
@@
 \paragraph{Polyglot.} Sharing the state via data store makes it possible to combine multiple
-langauge runtimes, as long as they support sharing data via data frames. In our prototype, this
+language runtimes, as long as they support sharing data via data frames. In our prototype, this
 includes R, JavaScript and TheGamma script, but the extensibility model allows adding further
@@
 A prototype implementation of the Wrattler system is available on GitHub (\url{http://github.com/wrattler}).
-The prototype implements language runtimes for JavaScript, TheGamma script and R 
-(Section~\ref{sec:comp-gamma}). It builds a dependency graph (Section~\ref{sec:comp-deps})
+The prototype implements language runtimes for TheGamma script (Section~\ref{sec:comp-gamma}), R
+and JavaScript. It builds a dependency graph (Section~\ref{sec:comp-deps})
 and uses it to evaluate results of cells.
 
-The data store (Section~\ref{sec:comp-data}) 
-stores data in Microsoft Azure in JSON format. Support for meta-data annotations and big data
-is not yet implemented. The protoype comes with one third-party tool -- the datadiff AI 
-assistant mentioned above (Section~\ref{sec:comp-data}). Figure~\ref{fig:proto} shows the 
-first cell of a sample Wrattler notebook analysing the UK broadband data.
+The data store (Section~\ref{sec:comp-data}) stores data in Microsoft Azure in JSON format. Support 
+for meta-data annotations and big data is not yet implemented. Storing notebook state in the data
+store also allowed us to develop an integration with Data diff \cite{datadiff},
+which provides data cleaning recommendations.
 
@@
 \vspace{-0.5em}
-\includegraphics[scale=0.15]{screen.png}
+\hspace{-1em}
+\includegraphics[scale=0.335]{gamma.png}
 \caption{\small{TheGamma script that downloads and aggregates UK government data, running in 
   Wrattler notebook with a live preview.}}
-\label{fig:proto}
+\label{fig:gamma}
 \vspace{-0.5em}
 \end{figure}
+
+%\begin{figure}
+%\vspace{-0.5em}
+%\includegraphics[scale=0.15]{screen.png}
+%\caption{\small{TheGamma script that downloads and aggregates UK government data, running in 
+%  Wrattler notebook with a live preview.}}
+%\label{fig:proto}
+%\vspace{-0.5em}
+%\end{figure}
 
@@
 of their limitations. To address reproducibility, some Jupyter extensions and systems such
-as R markdown \cite{rmarkdown} lock cells after evaluation.
+as R markdown %\cite{rmarkdown} 
+lock cells after evaluation.
 
-Dataflow notebooks \cite{dataflow} are attach unique hashes to cell 
+Dataflow notebooks \cite{dataflow} attach unique hashes to cell 
 evaluations. This allows the user to refer to dependencies explicitly and, in effect, construct
-a dependecy graph similar to ours manually.
-Scientific workflow systems \cite{taverna,kepler} manage evaluation of workflows over a 
-dependency graph similarly to Wrattler, but often allow editing it directly via a GUI, rather
+a dependecy graph manually.
+Scientific workflow systems \cite{taverna,kepler} manage evaluation over a 
+dependency graph similarly to Wrattler, but allow editing it directly via a GUI, rather
 than through code in a notebook.
@@
 kernel with a mechanism for capturing provenance based on light-weight annotations.
-Vizer \cite{vizer} focuses on integrating notebooks with spreadsheet-like interface.
+Vizer~\cite{vizer} focuses on integrating notebooks with spreadsheet-like interface.
 It internally uses a data store component similar to ours, but does not keep dependency graph
@@
 
-Our dependency grpah construction is inspired by Roslyn \cite{roslyn} and extends an
+Our binding process is inspired by Roslyn~\cite{roslyn} and extends an
 earlier work on TheGamma \cite{livegamma}. It is simiar to methods used in live programming 
@@
 \begin{acks}
-We thank to our colleagues from the AIDA (Artificial Intelligence for data analytics) project
-Chris Williams, Zoubin Ghahramani and Ian Horrocks and attendees of a recent 
-AIDA workshop. This work was in part supported by The Alan Turing 
-Institute under the EPSRC grant EP/N510129/1.
+The authors would like to acknowledge the funding provided by the UK Government's Defence \& 
+Security Programme in support of the Alan Turing Institute and the EPSRC grant EP/N510129/1.
+We thank to our colleagues Chris Williams, Zoubin Ghahramani and Ian Horrocks and attendees 
+of a recent AIDA project workshop.
 \end{acks}

### 2018-07-26 ai-assistants/paper.tex
What is an AI assistant?

\documentclass{article}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tgtermes}
\usepackage{forest}
\usepackage{fancyvrb}

\forestset{
  tt/.style={
    for tree={
      parent anchor=south,
      child anchor=north,
      grow=270,
      align=center,
      inner sep=0pt
    }
  },
  nice empty nodes up/.style={
    for tree={calign=fixed edge angles},
    delay={where content={}{shape=coordinate,for parent={for children={anchor=south}}}{}}
  }
}
\begin{document}
\title{What is an AI assistant?}
\author{Tomas Petricek}
\maketitle

\newcommand{\ident}[1]{\textnormal{\sffamily #1}}

\section{Introduction}
The idea of an \emph{AI assistant} has been discussed in many contexts in the AIDA project.
The Wrattler notebook system aims to provide a platform through which AI assistants can be 
easily used; the datadiff project developed an AI assistant which automates the process of 
reconciling inconsistencies between pairs of tabular datasets. However, we never clearly defined
what an AI assistant is. This document aims to collect some notes to help answer this question.
In the current version, these are mostly based on the needs of Wrattler and datadiff integration
in Wrattler -- as such, it clearly does not give the final answer, but merely one (limited, but
hopefully useful) perspective.

A simple way of trying to address the data wrangling problem would be to build a set of functions
that take some dirty datasets and produce one or more clean datasets. The user could then call these
functions on their dirty data and would get a clean result (conceivably, this could also be further 
automated). However, this simple approach has two problems. First, real-world data sets often 
contain numerous corner cases that require some human insight. Second, we want the data analysis
to be transparent and be able to understand how have the data wrangling problems been solved. 
A simple function does not address these because it does not encourage human interaction\footnote{At
best, users can run the function again with different parameters, but they get no suggestion about 
suitable values of the parameters.} and it is not transparent -- it produces a result without 
indicating how and why.

\section{What makes AI assistants unique}

AI assistants address the above issues as follows. First, AI assistants are interactive. Given 
some input data, they perform some analysis and provide the user with several options to choose 
from (those can be most likely solutions to some problem, possible ways of resolving ambiguous
corner cases, or a choice of next steps the AI assistant can follow). The options can be a finite
list (where the user has to choose one) or open-ended (where the user can specify some input).
They should also be sorted by likelihood (if we always choose the most likely option, the AI
assistant will behave as a fully automatic function).

Second, AI assistants do not directly transform data, but instead, generate scripts (or, more 
formally, expressions) in simple languages (domain-specific languages, or DSLs) that define how 
data should be transformed. The AI assistant can still use sophisticated machine learning 
algorithms (that do not produce an explanation of how they work) behind the scenes, but the end
result will always be a simple script that can be reviewed and understood by the user.

Note that each AI assistant can use its own domain-specific language -- the language should be
as small as possible, i.e. it should be able to express the data transformations that the AI
assistant can recommend, but nothing more.

The combination of interactivity and domain-specific languages allows us to do a number of 
interesting things:

\begin{itemize}
\item Scripts generated by AI assistant can be translated to R or Python (and it is easy to 
  add support for other languages), which means that one can use an AI assistant during data
  analysis and exploration, but then end up with code that can be used in production in any
  programming language.

\item The interactive nature of AI assistants makes it possible to give immediate feedback to
  users and let them correct potential issues early. For example, in Wrattler, we display the
  options that AI assistant provides together with a preview of results (of applying the
  generated expression to a sample dataset). 
\end{itemize}

\section{AI assistants as trees}
We can think of an AI assistant as a function that takes some input data and generates
a tree where each node offers a finite list of options that the user can choose from. (For
simplicity, we ignore the possibility of open-ended options where the user specifies some input 
and we also ignore the fact that the AI assistant might take some initial parameters in addition
to the input). Formally:
%
\begin{equation*}
\begin{array}{rcll}
a &=& \ident{Data} \rightarrow t &\textnormal{(AI assistant)}\\
t &=& (n_1, p_1, t_1), \ldots (n_k, p_k, e_k, t_k) &\textnormal{(Provided options)}\\
\mathit{eval} &=& \ident{Data}\times e \rightarrow \ident{Data}&\textnormal{(Evaluator for the DSL)}
\end{array}
\end{equation*}
%
An AI assistant $a$ is a function that takes some input data and returns a root tree node $t$. 
A tree node contains a list of options $(n_i, p_i, e_i, t_i)$ where $n_i$ is a name of the node (shown
to the user, but otherwise unimportant), $p_i$ is a number that gives the likelihood of the 
option (between 0 and 1), $e_i$ is an expression in the domain-specific language that represents
the data transformation attached to the node and $t_i$ is a sub-tree that can be empty or can offer 
more follow-up options. Note that this sub-tree can be generated lazily -- it is only needed if the 
user chooses the option with which it is associated and so we do not need to evaluate the full tree 
upfront.

The nature of expressions $e$ depends on the particular AI assistant and so we leave that abstract
for now. However, we certainly need to be able to apply the transformation on sample dataset, 
which is captured by the $\mathit{eval}$ function, which takes data together with a DSL expression
and returns transformed data. An (abstract) example tree might look as follows:

\vspace{0.5em}
\begin{forest}, tt  
  [
    [{$\mathbf{n_1}$\\$p=0.9$\\$e=e_1$}
      [{$\quad\mathbf{n_{11}}\quad$\\$p=0.9$\\$e=e_{11}$}]
      [{$\quad\mathbf{n_{12}}\quad$\\$p=0.7$\\$e=e_{12}$}]
    ]
    [{$\mathbf{n_2}$\\$p=0.8$\\$e=e_2$}
      [{$\quad\mathbf{n_{21}}\quad$\\$p=0.9$\\$e=e_{21}$}]
      [{$\quad\mathbf{n_{22}}\quad$\\$p=0.8$\\$e=e_{22}$}]
    ]
    [{$\mathbf{n_3}$\\$p=0.7$\\$e=e_3$}
      [{$\quad\mathbf{n_{31}}\quad$\\$p=0.8$\\$e=e_{31}$}]
      [{$\quad\mathbf{n_{32}}\quad$\\$p=0.6$\\$e=e_{32}$}]
    ]
  ]
\end{forest}
\vspace{0.5em}

\noindent
Here, the root offers three options represented by nodes $n_1, n_2, n_3$. The most likely one
as recommended by the AI assistant is $n_1$. The expression attached to the node is $e_1$ and
when the user selects the node, we can display a preview using sample data $\mathit{sample}$
by evaluating $\mathit{eval}(\mathit{sample}, e_1)$.

Assuming the user chooses the most likely option $n_1$, they will then be presented with two
subsequent options, $n_{11}, n_{12}$. Even though the AI assistant marks the first node as more
likely, the user may review the previews $\mathit{eval}(\mathit{sample}, e_{11})$,
$\mathit{eval}(\mathit{sample}, e_{12})$ and decide that the latter is actually the correct
data cleaning operation. They can then choose the node $n_{12}$ as the final one use the 
cleaning script represented by the DSL expression $e_{12}$.

\section{Examples of AI assistants}

What would a sample AI assistant look like when seen through the above model? We consider two
examples. The first is hypothetical AI assistant for extracting data from CSV files. The second
is the datadiff assistant. For datadiff, we discuss the current implementation in Wrattler, but
also briefly sketch a possible improvement (that improves the interactivity aspect).

\subsection{Cleaning CSV files}

As an example, consider the following broken CSV file. The fle starts with two meta-data lines
(one specifying the name and another specifying copyright information). This is followed by headers
and actual data.

\begin{Verbatim}[numbers=left,xleftmargin=8mm]
Counts of things
Copyright (c) Poor data export company, London, UK
Name, City, Count
Joe, London, 3
Jane, Edinburgh, 16
Jim, Cardiff, na
\end{Verbatim}

A hypothetical minimalistic AI assistant for cleaning CSV files could generate tree with two
levels of nodes. In the first level, it will aim to infer where the actual data is in the CSV
file (and it will offer its best guesses to the user). In the second level, it will aim to infer
the types of the columns.

The domain-specific language of data cleaning operations consists of two functions named
$\mathit{range}$ and $\mathit{types}$ that specify the range of data in the file and the types
of the columns: $\mathit{range}(\mathit{lines}=l_n\ldots l_m, \mathit{cols}=k)$ denotes that
the data starts on line $l_n$ and ends on line $l_m$ and there are $k$ columns; 
$\mathit{types}(t_1,\ldots, t_k)$ specifies that the types of the $k$ columns are 
$t_1, \ldots, t_k$. The types are either $\ident{string}$ or $\ident{int}$.

Given the above file as an input, the AI assistant might produce the following tree:
%
\begin{equation*}
\begin{array}{l}
p=0.9,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3)\\[0.2em]
\qquad\longrightarrow p=0.9,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
\qquad\longrightarrow p=0.3,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
p=0.7,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3)\\[0.2em]
\qquad\longrightarrow p=0.8,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
\qquad\longrightarrow p=0.7,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
p=0.3,~ e=\mathit{range}(\mathit{lines}=1\ldots 6, \mathit{cols}=1)\\[0.2em]
\qquad\longrightarrow p=0.9,~ e=\mathit{range}(\mathit{lines}=1\ldots 6, \mathit{cols}=1); \mathit{types}(\ident{string})\\[0.2em]
\end{array}
\end{equation*}
%
At the first level, the AI assistant gives us three options. The first, most likely, one skips just 
the first line and identifies that there are three columns (perhaps because the second line contains 
two commas and can thus be interpreted as data row). The second skips first two lines and recognises
three columns. The AI assistant might find this less likely, but it is the option we want to choose
as users. Finally, the third option takes all lines, but then it has to join all columns into one
which is unlikely.

The user can choose the second option, at which point a user interface can evaluate the 
expression $\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3)$ and show a preview of 
the cleaning result. Next, the user has a choice of two ways of inferring types -- the most 
likely one (and the right one) is that the third column has a type \ident{int} (which means that
the value {\texttt na} in the last row will be treated as missing).

The resulting expression 
$\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})$
can then be applied to the input data using the $\mathit{eval}$ function, which gives us 
columns \ident{Name} of type \ident{string}, \ident{City} of type \ident{string}
and \ident{Count} of type \ident{int} with data:
%
\begin{equation*}
\begin{array}{l}
(\textnormal{\sffamily "\hspace{-0.2em}Joe"}, \textnormal{\sffamily "\hspace{-0.2em}London"}, 3)\\
(\textnormal{\sffamily "\hspace{-0.2em}Jane"}, \textnormal{\sffamily "\hspace{-0.2em}Edinburgh"}, 16)\\
(\textnormal{\sffamily "\hspace{-0.2em}Jim"}, \textnormal{\sffamily "\hspace{-0.2em}Cardiff\hspace{0.1em}"}, \mathit{NaN})
\end{array}
\end{equation*}

\subsection{Datadiff (now)}

\subsection{Datadiff (better)}

\section{Evaluating AI assistants}

\subsection{Overall}
Following the best probability, we get the best result 
(this is equivalent to running a function)

\subsection{Interactivity}

Imagine there are some annomalies that we cannot detect.
Does the AI assistant provide an easy way for the user to address them?

\end{document}

### 2018-07-27 ai-assistants/paper.tex
Finish note on AI assistants

@@
 a &=& \ident{Data} \rightarrow t &\textnormal{(AI assistant)}\\
-t &=& (n_1, p_1, t_1), \ldots (n_k, p_k, e_k, t_k) &\textnormal{(Provided options)}\\
+t &=& (n_1, p_1, e_k, t_1), \ldots (n_k, p_k, e_k, t_k) &\textnormal{(Provided options)}\\
 \mathit{eval} &=& \ident{Data}\times e \rightarrow \ident{Data}&\textnormal{(Evaluator for the DSL)}
@@
 The domain-specific language of data cleaning operations consists of two functions named
-$\mathit{range}$ and $\mathit{types}$ that specify the range of data in the file and the types
-of the columns: $\mathit{range}(\mathit{lines}=l_n\ldots l_m, \mathit{cols}=k)$ denotes that
+$\ident{range}$ and $\ident{types}$ that specify the range of data in the file and the types
+of the columns: $\ident{range}(\ident{lines}=l_n\ldots l_m, \ident{cols}=k)$ denotes that
 the data starts on line $l_n$ and ends on line $l_m$ and there are $k$ columns; 
-$\mathit{types}(t_1,\ldots, t_k)$ specifies that the types of the $k$ columns are 
+$\ident{types}(t_1,\ldots, t_k)$ specifies that the types of the $k$ columns are 
 $t_1, \ldots, t_k$. The types are either $\ident{string}$ or $\ident{int}$.
@@
 \begin{array}{l}
-p=0.9,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
-\qquad\longrightarrow p=0.3,~ e=\mathit{range}(\mathit{lines}=2\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
-p=0.7,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3)\\[0.2em]
-\qquad\longrightarrow p=0.8,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
-\qquad\longrightarrow p=0.7,~ e=\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
-p=0.3,~ e=\mathit{range}(\mathit{lines}=1\ldots 6, \mathit{cols}=1)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ e=\mathit{range}(\mathit{lines}=1\ldots 6, \mathit{cols}=1); \mathit{types}(\ident{string})\\[0.2em]
+p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3)\\[0.2em]
+\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
+\qquad\longrightarrow p=0.3,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
+p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)\\[0.2em]
+\qquad\longrightarrow p=0.8,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
+\qquad\longrightarrow p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
+p=0.3,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1)\\[0.2em]
+\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1); \ident{types}(\ident{string})\\[0.2em]
 \end{array}
@@
 The user can choose the second option, at which point a user interface can evaluate the 
-expression $\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3)$ and show a preview of 
+expression $\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)$ and show a preview of 
 the cleaning result. Next, the user has a choice of two ways of inferring types -- the most 
@@
 The resulting expression 
-$\mathit{range}(\mathit{lines}=3\ldots 6, \mathit{cols}=3); \mathit{types}(\ident{string},\ident{string},\ident{int})$
+$\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})$
 can then be applied to the input data using the $\mathit{eval}$ function, which gives us 
@@
 
-\subsection{Datadiff (now)}
+\subsection{Current design of datadiff AI assistant}
 
-\subsection{Datadiff (better)}
+The datadiff project takes two tabular datasets that are in inconsistent formats, but represent
+two sets of observations about the same entity (such as broadband speed in two different years).
+It generates a list of patches that can be applied to one dataset in order to convert it to the
+format used by the second dataset. The patches form a domain-specific language (DSL) of the 
+data transformation suggested by datadiff. A patch $p$ can be one of the following
+(the details are not important for this document, but it provides a nice example of a DSL):
+
+\begin{equation*}
+\begin{array}{lcll}
+p &=& \ident{permute}(\pi) &\textnormal{(Reorder columns using a given permutation)}\\
+  &|& \ident{recode}(i, [s_1\mapsto s_2, \ldots]) &\textnormal{(Recode categorical values using in column $i$)}\\
+  &|& \ident{linear}(i, a, b)&\textnormal{(Apply linear transform $va+b$ to a numerical column $i$)}\\
+  &|& \ident{delete}(i)&\textnormal{(Drop the column at index $i$)}\\
+  &|& \ident{insert}(i, d)&\textnormal{(Insert a new column at index $i$ initialized to values $d$)}
+\end{array}
+\end{equation*}
+%
+The current implementation of datadiff is a non-interactive R package that takes two datasets
+and produces a list of patches. It takes a couple of parameters that specify which patches should
+datadiff attempt to generate (together with various weights to tweak the optimization).
+
+The current datadiff AI assistant as implemnented in Wrattler generates a tree that allows the
+user to select and apply individual patches. The user can review each patch to make sure that it
+is a reasonable transformation. For example, consider the following two datasets:
+
+\vspace{0.5em}
+\begin{minipage}[t]{0.5\textwidth}
+\begin{Verbatim}[numbers=left,xleftmargin=3mm]
+Name, City
+Joe, London
+Jane, Edinburgh
+Jim, London
+\end{Verbatim}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+\begin{Verbatim}[numbers=left,xleftmargin=3mm]
+City, Name, Count
+Cardiff, Alice, 1
+Cardiff, Bob, na
+Edinburgh, Bill, 2
+\end{Verbatim}
+\end{minipage}
+\vspace{1em}
+
+\noindent
+If we ask datadiff for a list of patches to transform the dataset on the right-hand side
+to the format of the dataset on the left-hand side, we might get the following patches
+(assuming columns are indexed from $1$):
+
+\[
+\ident{delete}(3);~ \ident{permute}(2,1);~ \ident{recode}(2, [\textnormal{\sffamily "\hspace{-0.2em}Cardiff\hspace{0.1em}"}\mapsto\textnormal{\sffamily "\hspace{-0.2em}London"}])
+\]
+
+\noindent
+Datadiff correctly infers that we need to drop the newly added {\ttfamily Count} column and that
+the order of {\ttfamily Name} and {\ttfamily City} has been switched. It might also suggest that
+{\ttfamily City} is a categorical column where the encoding has been changed, which is not the 
+case here (but this would be a reasonable guess if one dataset contained values ``true'' and ``false''
+while the other contained ``yes'' and ``no'').
+
+The datadiff AI assistant generates the following tree based on the three inferred patches 
+(we omit the parameters of the patches):
+%
+\begin{equation*}
+\begin{array}{l}
+p=0.9,~ e=\ident{delete}(\ldots)\\[0.2em]
+\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{permute}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{recode}(\ldots); \ident{permute}(\ldots)\\[0.4em]
+p=0.8,~ e=\ident{permute}(\ldots)\\[0.2em]
+\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{delete}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{recode}(\ldots); \ident{delete}(\ldots)\\[0.4em]
+(\ldots)
+\end{array}
+\end{equation*}
+
+At the first level, the user can choose one of the three patches (we only show sub-trees for the first two). 
+The patches are sorted by their likelihood and so \ident{delete} is offered before \ident{permute}. In the
+second level, the user can choose one of the remaining patches. Assuming the user follows the first
+branch and chooses \ident{delete}, they can now choose to apply \ident{permute} or \ident{recode} (the
+first option being more likely). In the last third level, the AI assistant offers the remaining patch. 
+However, the user can stop before reaching the leaf and choose just
+$\ident{delete}(3);~ \ident{permute}(2,1)$ as the final data cleaning script. As before, Wrattler 
+can use the \emph{eval} function to give previews of the transformed datasets while the user is 
+navigating through the tree.
+
+\subsection{Alternative design of datadiff AI assistant}
+
+The current design of the datadiff AI assistant is not optimal, because it performs a global
+optimization up-front and then merely allows the user to choose some of the patches. This is 
+problematic, because the global optimization can be slow and it does not use the user input to 
+improve the results.
+
+For example, if the user could specify that a certain column should not be
+recoded, datadiff could run the global optimization taking this constraint into account and, say,
+match and reorder columns differently (this is currently not possible, because the current version
+runs the global optimization once up-front and then merely offers a choice of produced patches).
+
+An alternative datadiff AI assistant would generate expressions in the same domain-specific 
+language as the current one (list of patches $p$). However, the tree would be generated and 
+structured differently. This could be done by maintaining a list of constraints and generating
+the best possible patch given certain constraints, together with a choice of most likely constraints
+the user might want to add. A constraint $c$ would be as follows:
+%
+\begin{equation*}
+\begin{array}{lcll}
+c &=& \ident{norecode}(i) &\textnormal{(Values in a categorical column $i$ should not be recoded)}\\
+  &|& \ident{notransform}(i)&\textnormal{(Values in a numerical column $i$ should not be transformed)}\\
+  &|& \ident{nodelete}(i)&\textnormal{(Column $i$ should not be deleted)}\\
+  &|& \ident{match}(i,j)&\textnormal{(Column $i$ should be mapped to sample column $j$)}
+\end{array}
+\end{equation*}
+%
+The first three constraints specify that certain patches should not be generated. The last one is
+more interesting as it allows the user to explicitly give a mapping for a column (but unless other
+constraints are given, datadiff can still attempt to transform or recode the column to get a 
+better match). This is likely not a complete set of constraints, but it is sufficient to illustrate
+the idea.
+
+The AI assistant would then implement a function that takes two datasets together with a set of 
+constraints and generates a cleaning script together with a list of constraints (sorted by their
+likelihood) that the user might want to add. (If we allowed nodes with open-ended options, the
+user could write an arbitrary constraint, but for now, we just assume that the AI assistant can
+generate all sensible options.)
+
+Using the above function, we can produce a tree as follows. We start with an empty set of 
+constraints $C=\{\}$ and call the function to obtain a data cleaning DSL expression $e$ and
+potential constraints $c_1, \ldots, c_k$. For each constraint, we add it to the set $C$ and
+generate a sub-tree by calling the AI assistant function recursively. In other words, following
+a path through the tree would add more and more constraints to the set $C$ and each node will
+come with the best DSL expression that datadiff can generate under the constraints determined
+by the path taken through the tree. The following illustrates (a small part of) the tree that 
+the revised AI assistant for datadiff would generate for the above two files. (In addition to a DSL 
+expression $e$ and likelihood $p$, we also include a set of constriants $C$ associated with each 
+node):
+%
+\begin{equation*}
+\begin{array}{l}
+C=[], p=1,~ e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{norecode}(2)], p=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);\\
+\qquad\qquad\longrightarrow C=[\ident{norecode}(2); \ident{nodelete}(3)], p=0.2, e=(\ldots)\\
+\qquad\longrightarrow C=[\ident{nodelete}(3)], p=0.2, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{match}(1,1)], p=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{match}(1,2)], p=0.8, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\end{array}
+\end{equation*}
+%
+The tree has only one root node which was generated without any constraints. The generated expression
+is the best guess by datadiff, which drops the {\texttt Count} column, rearranges columns and recodes
+(incorrectly) the {\texttt City} column. The nodes at the second level allow the user to add a
+number of constraints.
+
+The first two sub-nodes prevent datadiff from generating two patches that it employed. The first one
+disallows recoding the column {\texttt City}. The resulting expression drops {\texttt Count} and
+rearranges the columns, which is the result we were hoping to obtain. 
+The second sub-node disallows deleting the {\texttt Count} node. In this case, datadiff has to
+find other ways of reconcilling the data sets, perhaps by recoding {\texttt Code} and dropping
+another column. Finally, the last two sub-nodes allow us to explicitly specify a mapping between
+columns. The listing only shows the first two (map column $1$ to column $1$; map column $1$ to column $2$),
+but the AI assistant would likely generate the full range, sorted by likelihood.
 
@@
 
-\subsection{Overall}
-Following the best probability, we get the best result 
-(this is equivalent to running a function)
+When building an AI assistant, we want to be able to evaluate how good it is. For an AI tool that 
+is implemented as a data cleaning function, we can run it on sample inputs and asses how good 
+the resulting outputs are. (Even this is not so simple, but we might know what the optimal results 
+look like for some inputs, or we could generate inputs by introducing errors into clean datasets 
+and see if those were eliminated.)
 
-\subsection{Interactivity}
+The situation with AI assistants is more complicated as they are interactive and aim to produce
+cleaning scripts that humans can understand. We could evaluate AI assistants using user studies 
+(and this would be, no doubt, worthwhile), but there are a couple of things that we can say about
+AI assistants in a formal way too.
 
-Imagine there are some annomalies that we cannot detect.
-Does the AI assistant provide an easy way for the user to address them?
+\subsection{Evaluating AI assistant in an automatic way}
+First of all, we can treat an AI assistant as a fully automatic tool. To do this, we automatically
+choose the most likely option provided by the AI assistant until we reach a leaf node (or, perhaps,
+until we reach a case where the sub-node has a likelihood $p\leq 0.5$).
+
+More formally, given a dataset $d$ and an AI assistant $a$ (as defined on page 2), we run the 
+AI assistant $a(d)$ to obtain a list of options $(n_1, p_1, e_k, t_1), \ldots (n_k, p_k, e_k, t_k)$.
+We choose the most likely node $(n_i, p_i, e_i, t_i)$ (i.e. $p_i \geq p_j$ for all $j$). If 
+$t_i$ is empty or the likelihoods of all options inside $t_i$ are less than $0.5$, we return
+an expression $e_i$; otherwise, we recursively follow the sub-tree $t_i$. Finally, we evaluate
+the resulting data cleaning expression using $\mathit{eval}(d, e_i)$ and asses how good the 
+cleaned data set is using whatever methods we would use for an automatic cleaning function.
+
+This way, we can guarantee that the interactive AI assistant is as good as an automatic cleaning 
+function would be. This is a reasonable baseline test, but it does not tell us much about how
+good the interactive aspects are.
+
+\subsection{Evaluating interactivity for annomaly handling}
+
+One of the motivations for the idea of AI assistants was that real-world data cleaning tasks
+often have ad-hoc issues that require some form of manual intervention. In other words, the 
+file might have an annomaly that a human can spot, but that an automatic data cleaning algorithm
+would miss. For example, consider the following two CSV files:
+
+\vspace{0.5em}
+\begin{minipage}[t]{0.5\textwidth}
+\begin{Verbatim}[numbers=left,xleftmargin=3mm]
+Name, Count, ID
+Jim, 45, 1
+Jane, 52, 2
+Joe, 67, 3
+\end{Verbatim}
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+\begin{Verbatim}[numbers=left,xleftmargin=3mm]
+Name, Code, Value
+Alice, 35, 4
+Bill, 38, 5
+Bob, 42, 6
+\end{Verbatim}
+\end{minipage}
+\vspace{1em}
+
+\noindent
+As humans, we can see that the column {\texttt ID} should be aligned with {\texttt Code}
+and {\texttt Count} should likely match {\texttt Value}. However, unless the AI assistant
+considers possible meanings of the file headers, it has no way of discovering this.
+
+We might call such annomalies \emph{adversary annomaly}. There is very likely no way of 
+characterising adversary annomalies in general for all AI assistants, but it should be possible
+to characterise them for concrete data wrangling problems. Examples might include rows that
+are not outliers in the statistical sense, textual values that have the right format, but wrong 
+meaning such as ``Refugee Olympic Athletes'' among the names of countries\footnote{This is a name
+for an actual team of 10 athletes competing in the 2016 Rio Olympic games.}, or a suspicious 
+numerical value such as $123.45678$ in a numerical column with values 
+$190.52136, 58.68297, 123.45678, 67.47067, 47.69846$.
+
+An important quality of an AI assistant is tht it makes it possible to eliminate adversary
+annomalies with a small number of interactions, ideally just by choosing an appropriate 
+non-default offered option at one point when traversing the tree. For example, given the
+above example, the alternative datadiff AI assistant would allow us to choose a tree node that
+adds a constraint $\ident{match}(2,3)$ and we would get the correct result with just one manual
+intervention (the other column would be reordered automatically).
+
+To summarise, we want to be able to claim that an AI assistant makes it easy to handle ad-hoc
+corner cases that are not discovered by the machine learning algorithm behind it. To do this,
+we introduce adversary annomalies into the data -- those are plausible real-world data annomalies 
+that are not discovered by the algorithm emplyed by the AI assistant. A good AI assistant allows
+the user to eliminate these with a minimal number of interactions, i.e.~choices of other than
+the most likely nodes in the tree.
+
+\subsection{Easy discoverability of best results}
+
+Finally, the current integration for AI assistants in Wrattler exposes the tree through a
+user interface. The user calls the AI assistant and then chooses a sub-tree to explore 
+(by typing `.' and choosing one of the options in a drop-down menu). This means that it is
+desirable to make the most likely result (domain-specific language expression) closer to the
+root rather than somewhere deep in the tree.
+
+As an example, consider the two designs of the datadiff assistant. In the first design, the
+user has to manually accept all the desired patches. If the optimal patch needs to drop 15
+extra columns from a dataset, the user will have to make 15 choices explicitly. The
+second design of the datadiff assistant is better because the most likely set of patches
+is available immediately in the root node (and users only need to continue exploring a
+sub-tree if they need to address an adversary annomaly).
 

### 2018-11-02 ai-assistants/paper.tex
More work on AI assistants report

@@
 \usepackage{fancyvrb}
+\usepackage{setspace}
+\usepackage[font=small,labelfont=bf]{caption}
 
@@
 \begin{document}
-\title{What is an AI assistant?}
+\title{\textbf{AI assistants}: A framework for semi-automated, transparent, tooling-rich data wrangling}
 \author{Tomas Petricek}
 \maketitle
+% \begin{abstract}
+%   zz
+% \end{abstract}
 
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
+\definecolor{tpclr}{rgb}{0.6,0,0.3}
+\newcommand{\tpnote}[1]{{\marginpar{
+  \setstretch{0.8}\raggedright\textcolor{tpclr}{ \footnotesize{\textbf{Tomas:} #1 } }}
+}}
+~
 
 \section{Introduction}
-The idea of an \emph{AI assistant} has been discussed in many contexts in the AIDA project.
-The Wrattler notebook system aims to provide a platform through which AI assistants can be 
+\tpnote{This part is written as an intro to a potential paper about AI assistants 
+(using many ideas and formulations from Charles et al.)}
+
+According to some reports, 80\% of data analysis is spent on \emph{data wrangling} -- an activity
+that encompasses obtaining data from a diversity of semi-structured sources, making sense of data
+encoding, joining multiple datasets with mismatched structure, correcting erroneous records and 
+filling missing data. Despite the recent rise of artificial intelligence and machine learning, 
+data wrangling has, so far, largely eluded automation and remains a tedious and time-consuming 
+manual task.
+
+There are a number of reasons why data wrangling is hard to automate. First, data is often big and 
+diverse and hide hard-to-find special cases that can be only handled with some human insight.
+Second, real-world data is heterogeneous. Automatic methods developed for one kind of data will 
+often fail when more data is added. Third, data cleaning often requires tuning of many parameters 
+and you often cannot know what will work until you review the results. Finally, making an error
+during data wrangling can have a disasterous effect on a project. An automatic tool can easily 
+confuse interesting outliers for an uninteresting noise in cases where a quick human glance would
+immediately spot the distinction.
+
+For these rasons, we argue that a major advance in data wrangling is best achieved by developing
+semi-automatic methods that allow us to effectively combine automation and scalability of machine 
+learning methods with human insight. To capture the human-computer interaction pattern that is
+required for data wrangling, we introduce the concept of an \emph{AI assistant}. An AI assistant
+is a component that interacts with a data analyst to guide them through a single data extraction,
+collection, integration, preparation or exploration task. AI assistants are semi-automated, 
+transparent and tooling-rich:
+%
+\begin{enumerate}
+\item \textbf{Semi-automated.} After being invoked on input data, the AI assistant runs some automated
+  data wrangling algorithm (based on AI, machine learning or statistical analysis) and offers a list
+  of suggestions to the data analyst. The analyst can review the results, choose one of the provided
+  options or, more generally, provide feedback back to the algorithm. The loop is repeated until 
+  the AI assistant has no further suggestions or until the analyst accepts one of the suggested results.
+
+\vspace{-0.2em}
+\item \textbf{Transparent.} An AI assistant needs to explain what cleaning operations it recommends. 
+  At each step, the result of running an AI assistant is a short data wrangling script, written 
+  in a domain-specific language, designed for each individual assistant. The data analyst can 
+  review the script to assess if the recommended cleaning operation is reasonable, before running
+  the script on the input data.
+
+\vspace{-0.2em}
+\item \textbf{Tooling-rich.} Finally, an AI assistant needs to play well with modern data science
+  tooling. This means that offering the suggestions to the data analyst should be done through an 
+  interactive programming mechanism such as auto-complete. When reviewing suggestions, the data 
+  analyst should also immediately see a preview of running the script on the actual input data as,
+  for example, when running code in Jupyter notebooks.
+\end{enumerate}
+%
+Several tools that satisfy the above definition of an AI assistant have been built in the past, 
+but our main motivation is to describe the general structure of an AI assistant abstractly and
+discuss the methodology of developing and evaluating AI assistants.
+
+A data analyst will typically need to interact with many separate assistants over the course of an
+anlysis, each of which specializes in a single kind of task. Having a clear abstract definition
+of what an AI assistant is, together with several examples, will provide a common framework that 
+developers of AI assistants and authors of data science tools can use in order to build a 
+flourishing ecosystem for semi-automated, transparent, tooling-rich data wrangling.
+
+\tpnote{I think this is good structure of contributions for a real paper. The current version does not quite do all this yet though!}
+This paper explains what an AI assistant is, discusses how it can be integrated in a notebook
+environment for data science and gives a number of examples of AI assistants: 
+%
+\begin{itemize}
+\item We give a formal description that models AI assistants as trees (Section~\ref{sec:trees}). 
+  An AI assistant offers user a choice by generating multiple sub-trees and the user makes a choice 
+  and provides an input to the AI assistant by choosing one branch.
+\vspace{-0.2em}
+\item We discuss the integration of AI assistants in the Wrattler notebook system (Section~\ref{sec:notebook}). 
+  This makes it easy to use AI assistants interactively and Wrattler also gives rapid feedback when 
+  cleaning data using an AI assistant. 
+\vspace{-0.2em}
+\item We give two examples of AI assistants and explain how the tree-based model captures those
+  examples (Section~\ref{sec:examples}). We consider \emph{Datadiff}, which automates the process of reconcilling 
+  inconsistencies between pairs of tabular data sets and a \emph{CSV parser} that extracts typed
+  data from real-world messy CSV files.
+\vspace{-0.2em}
+\item We consider two ways of evaluating AI assistants (Section~\ref{sec:evaluation}). An AI 
+  assistant should provide the best answer when used fully automatically, but user interaction 
+  should allow correcting it when the answer obtained automatically is not right.
+\end{itemize}
+
+\section{Background}
+\label{sec:background}
+\tpnote{These are comments for AIDA team readers only, somewhat overlapping with the introduction.
+We'll need to rewrite this for a real paper. }
+
+The idea of an \emph{AI assistant} appeared in many contexts in the AIDA project.
+Wrattler aims to provide a platform through which AI assistants can be 
 easily used; the datadiff project developed an AI assistant which automates the process of 
 reconciling inconsistencies between pairs of tabular datasets. However, we never clearly defined
-what an AI assistant is. This document aims to collect some notes to help answer this question.
-In the current version, these are mostly based on the needs of Wrattler and datadiff integration
-in Wrattler -- as such, it clearly does not give the final answer, but merely one (limited, but
-hopefully useful) perspective.
+what an AI assistant is.
 
@@
 that take some dirty datasets and produce one or more clean datasets. The user could then call these
-functions on their dirty data and would get a clean result (conceivably, this could also be further 
-automated). However, this simple approach has two problems. First, real-world data sets often 
-contain numerous corner cases that require some human insight. Second, we want the data analysis
+functions on their dirty data and would get a clean result. The function would likely take some
+parameters that specify how cleaning should be done. (Conceivably, this could be further automated
+and a system could invoke such functions automatically.)
+
+However, this simple approach has two problems. First, real-world data sets often 
+contain numerous corner cases that require some human insight and so getting the right results 
+might involve laborious fiddling with function parameters. Second, we want the data analysis
 to be transparent and be able to understand how have the data wrangling problems been solved. 
-A simple function does not address these because it does not encourage human interaction\footnote{At
-best, users can run the function again with different parameters, but they get no suggestion about 
-suitable values of the parameters.} and it is not transparent -- it produces a result without 
-indicating how and why.
+A simple function does not address these because it does not encourage human interaction
+and it is not transparent -- it produces a result without indicating how and why.
 
-\section{What makes AI assistants unique}
+\subsection{Why are assistants more than functions}
 
-AI assistants address the above issues as follows. First, AI assistants are interactive. Given 
-some input data, they perform some analysis and provide the user with several options to choose 
+\tpnote{I like the idea of contrasting `cleaning function' with an `assistant' but this text
+repeats ideas from the intrduction and does not do a very good job overall.}
+AI assistants address the above issues. First, AI assistants are interactive. Given 
+some data, they perform some analysis and provide the user with several options to choose 
 from (those can be most likely solutions to some problem, possible ways of resolving ambiguous
-corner cases, or a choice of next steps the AI assistant can follow). The options can be a finite
-list (where the user has to choose one) or open-ended (where the user can specify some input).
-They should also be sorted by likelihood (if we always choose the most likely option, the AI
-assistant will behave as a fully automatic function).
+corner cases, or a choice of next steps the AI assistant can follow). In this paper, the options are 
+a finite list (where the user has to choose one), but they could also be open-ended (where the user 
+can specify some input). The list of options is also sorted by quality (if we always choose the 
+most likely option, the assistant will behave as an automatic function).
 
@@
 as small as possible, i.e. it should be able to express the data transformations that the AI
-assistant can recommend, but nothing more.
-
-The combination of interactivity and domain-specific languages allows us to do a number of 
-interesting things:
+assistant can recommend, but nothing more. The combination of interactivity and domain-specific 
+languages allows us to do a number of interesting things:
 
@@
 
-\section{AI assistants as trees}
-We can think of an AI assistant as a function that takes some input data and generates
-a tree where each node offers a finite list of options that the user can choose from. (For
-simplicity, we ignore the possibility of open-ended options where the user specifies some input 
-and we also ignore the fact that the AI assistant might take some initial parameters in addition
-to the input). Formally:
+\begin{figure}
+\vspace{-1em}
+\includegraphics[scale=0.29]{wrattler.png}
+\caption{The datadiff AI assistant running inside Wrattler, a notebook-like environment for
+data science. The completion list shows data cleaning operations offered by an AI assistant 
+and the preview shows the result of running the provided cleaning script.}
+\label{fig:wrattler}
+\end{figure}
+
+\section{AI assistants in a notebook environment}
+\label{sec:notebook}
+
+Before looking at the formal model which aims to explain what an AI assistant is, it is useful 
+to explore how AI assistants can be used in a notebook programming environment for data science 
+such as Jupyter. Figure~\ref{fig:wrattler} shows a screenshot of a datadiff AI assistant, running
+in the Wrattler notebook system.
+
+In the example, a user obtained two datasets, \texttt{bb2015} and \texttt{bb2014} that contain 
+information about broadband speeds for years 2015 and 2014. The two data sets record observations
+over two years, containing a common subset of data, but in a mismatched format -- the column names
+and order differ. The data analyst invokes the datadiff AI 
+assistant in order to turn the format of 2015 data into the format used by the 2014 dataset.
+To do this, they type \texttt{datadiff.adapt(bb2015, bb2014)}, which invokes the AI assistant.
+They then type ``.'' to get a data cleaning recommendation.
+
+The AI assistant analyses the data and determines a list of patches that needs to be applied in
+order to reformat the 2015 dataset. It then offers those patches in the autocompletion list, 
+sorted from the most likely patch to the least likely patch. The user can proceed by selecting 
+the first patch repeatedly (as in the above screenshot). After selecting a patch, the source code
+in the cell changes and the preview is updated, showing the result of running the current cleaning
+script. This interactivity makes it easy to see possible data cleaning errors. If applying a patch
+does not lead to an expected result, the user can easily go back and select the second most likely
+patch.
+
+Modelling an AI assistant as a tree, we can see that the code entered in the editor corresponds to
+a path through the tree (where the user provides information for the AI asistant; here, by 
+accepting some of the recommended patches). The options shown in an auto-complete list correspond 
+to the possible sub-trees in the current tree node (here, the AI assistant is asking for an user 
+input or a recommendation on how to continue).
+
+\section{Formal model of AI assistants}
+\label{sec:trees}
+
+An AI assistant is a concept that spans multiple research areas. In order to understand how to 
+integrate AI assistants into programming tools such as notebooks, we need to use methods from 
+programming language research. To implement interesting AI assistants, we need to use machine 
+learning and statistical models.
+
+From a programming language perspective, we are mainly interested in the abstract structure of the
+AI assistant. A programming langauge model will capture how data is passed to the assistant, how
+is the assistant invoked and how does it interact with the analyst. From a machine learning
+perspective, we are interested in how to build AI assistants. For example, many assistants may fit
+in the Bayesian learning model where the AI assistant updates an initial model -- a prior --
+based on inputs from the analyst.
+
+\tpnote{It would be nice to have more about Bayesian model of AI assistants - but I'm not qualified
+to write that. We'd need to show how a Bayesian model defines ``AI assistant as a tree structure'' 
+(probably quite easy) and give some Bayesian AI assistant examples... (could be quite simple, but
+we don't have them)}
+However, it is conceivable that an AI assistant can be implemented in other ways. It might provide
+suggestions based on a simple frequentist statistical analysis, or it might use models based on 
+neural networks. The programming language model of what is an AI assistant should be able to 
+accommodate all those different implementations. Modelling AI assistants as trees, as discussed
+in the next section, satisfies this requirement as it allows a range of different concrete
+implementations.
+
+\subsection{Programming abstraction: AI assistants as trees}
+\label{sec:trees-def}
+
+To discuss properties of AI assistants from a programming language perspective, we model an AI 
+assistant as a function that takes some input data and generates a tree where each node offers a 
+finite list of options that the user can choose from. For simplicity, we ignore the possibility of 
+open-ended options where the user specifies some input and we also ignore the fact that the AI 
+assistant might take some initial parameters in addition to the input. Assuming \ident{Data}
+is a representation of input data:
 %
@@
 a &=& \ident{Data} \rightarrow t &\textnormal{(AI assistant)}\\
-t &=& (n_1, p_1, e_k, t_1), \ldots (n_k, p_k, e_k, t_k) &\textnormal{(Provided options)}\\
+t &=& (n_1, r_1, e_k, t_1), \ldots (n_k, r_k, e_k, t_k) &\textnormal{(Provided options)}\\
 \mathit{eval} &=& \ident{Data}\times e \rightarrow \ident{Data}&\textnormal{(Evaluator for the DSL)}
@@
 %
-An AI assistant $a$ is a function that takes some input data and returns a root tree node $t$. 
-A tree node contains a list of options $(n_i, p_i, e_i, t_i)$ where $n_i$ is a name of the node (shown
-to the user, but otherwise unimportant), $p_i$ is a number that gives the likelihood of the 
-option (between 0 and 1), $e_i$ is an expression in the domain-specific language that represents
-the data transformation attached to the node and $t_i$ is a sub-tree that can be empty or can offer 
-more follow-up options. Note that this sub-tree can be generated lazily -- it is only needed if the 
-user chooses the option with which it is associated and so we do not need to evaluate the full tree 
-upfront.
+An AI assistant $a$ is a function that takes some input data and returns a root tree node~$t$. 
+A tree node consists of a list of branches $(n_i, r_i, e_i, t_i)$ where $n_i$ is a name of the branch 
+(shown to the user, but otherwise unimportant), $r_i$ is a number that gives a rating of the option
+(between 0 and 1), $e_i$ is an expression in the domain-specific language that represents
+the data transformation attached to the node. Finally, $t_i$ is a sub-tree that can be empty or can 
+offer more follow-up options. Note that this sub-tree can be generated lazily -- it is only needed 
+if the user chooses the option with which it is associated and so we do not need to evaluate the 
+full tree upfront.
 
 The nature of expressions $e$ depends on the particular AI assistant and so we leave that abstract
-for now. However, we certainly need to be able to apply the transformation on sample dataset, 
-which is captured by the $\mathit{eval}$ function, which takes data together with a DSL expression
-and returns transformed data. An (abstract) example tree might look as follows:
+for now. However, we certainly need to be able to apply the transformation on sample dataset. 
+This is captured by the $\mathit{eval}$ function, which takes data together with an expression $e$
+in the domain-specific langauge and returns transformed data.
 
-\vspace{0.5em}
-\begin{forest}, tt  
+\newpage
+
+\begin{figure}
+\hspace{4em}\begin{forest}, tt  
   [
     [{$\mathbf{n_1}$\\$p=0.9$\\$e=e_1$}
-      [{$\quad\mathbf{n_{11}}\quad$\\$p=0.9$\\$e=e_{11}$}]
-      [{$\quad\mathbf{n_{12}}\quad$\\$p=0.7$\\$e=e_{12}$}]
+      [{$\quad\mathbf{n_{11}}\quad$\\$r=0.9$\\$e=e_{11}$}]
+      [{$\quad\mathbf{n_{12}}\quad$\\$r=0.7$\\$e=e_{12}$}]
     ]
     [{$\mathbf{n_2}$\\$p=0.8$\\$e=e_2$}
-      [{$\quad\mathbf{n_{21}}\quad$\\$p=0.9$\\$e=e_{21}$}]
-      [{$\quad\mathbf{n_{22}}\quad$\\$p=0.8$\\$e=e_{22}$}]
+      [{$\quad\mathbf{n_{21}}\quad$\\$r=0.9$\\$e=e_{21}$}]
+      [{$\quad\mathbf{n_{22}}\quad$\\$r=0.8$\\$e=e_{22}$}]
     ]
     [{$\mathbf{n_3}$\\$p=0.7$\\$e=e_3$}
-      [{$\quad\mathbf{n_{31}}\quad$\\$p=0.8$\\$e=e_{31}$}]
-      [{$\quad\mathbf{n_{32}}\quad$\\$p=0.6$\\$e=e_{32}$}]
+      [{$\quad\mathbf{n_{31}}\quad$\\$r=0.8$\\$e=e_{31}$}]
+      [{$\quad\mathbf{n_{32}}\quad$\\$r=0.6$\\$e=e_{32}$}]
     ]
@@
 \end{forest}
-\vspace{0.5em}
+\caption{An illustration of a formal model that views AI assistants as trees.}
+\label{fig:aitree}
+\vspace{-1em}
+\end{figure}
 
 \noindent
-Here, the root offers three options represented by nodes $n_1, n_2, n_3$. The most likely one
-as recommended by the AI assistant is $n_1$. The expression attached to the node is $e_1$ and
-when the user selects the node, we can display a preview using sample data $\mathit{sample}$
-by evaluating $\mathit{eval}(\mathit{sample}, e_1)$.
+Figure~\ref{fig:aitree} illustrates the formal model and shows an (abstract) example of a tree
+that can be generated by an AI assistant. The root node offers three options represented by nodes 
+$n_1, n_2, n_3$. The most likely one as recommended by the AI assistant is $n_1$ with rating $r=0.9$. 
+The expression attached to the node is $e_1$ and when the user selects the node, we can display a 
+preview using sample data $\mathit{sample}$ using the $\mathit{eval}$ function by evaluating 
+$\mathit{eval}(\mathit{sample}, e_1)$.
 
@@
 data cleaning operation. They can then choose the node $n_{12}$ as the final one use the 
-cleaning script represented by the DSL expression $e_{12}$.
+cleaning script represented by the DSL expression $e_{12}$. The expression $e_{12}$ can be used
+to get the clean data by evaluating $\mathit{eval}(\mathit{sample}, e_{12})$, but it can also be
+translated to a programming language such as R or Python and used in a larger data wrangling 
+workflow.
 
 \section{Examples of AI assistants}
+\label{sec:examples}
 
-What would a sample AI assistant look like when seen through the above model? We consider two
-examples. The first is hypothetical AI assistant for extracting data from CSV files. The second
+\tpnote{The examples here illustrate the idea, but are not detailed enough for a proper paper
+(we should also have only the better version of datadiff). It would be ideal to have CSV, better
+datadiff and some Bayesian example. Possibly also ptype, but it might not fit the model that well --
+we need to think about that.}
+
+We now consider two AI assistants through the lens of our abstract tree model.
+The first is hypothetical AI assistant for extracting data from CSV files. The second
 is the datadiff assistant. For datadiff, we discuss the current implementation in Wrattler, but
@@
 
-As an example, consider the following broken CSV file. The fle starts with two meta-data lines
+\tpnote{This is inspired by Gerrit's work, but it's very basic at this point.}
+
+As an example, consider the following broken CSV file. The file starts with two meta-data lines
 (one specifying the name and another specifying copyright information). This is followed by headers
@@
 
-A hypothetical minimalistic AI assistant for cleaning CSV files could generate tree with two
-levels of nodes. In the first level, it will aim to infer where the actual data is in the CSV
-file (and it will offer its best guesses to the user). In the second level, it will aim to infer
-the types of the columns.
+A minimalistic AI assistant for cleaning CSV files generates tree with two levels of nodes. 
+In the first level, it infers where the actual data is in the CSV file (and it offers 
+its best guesses to the user). In the second level, it infers the types of the columns.
 
-The domain-specific language of data cleaning operations consists of two functions named
+The domain-specific language for CSV parsing consists of just two functions named
 $\ident{range}$ and $\ident{types}$ that specify the range of data in the file and the types
@@
 
-Given the above file as an input, the AI assistant might produce the following tree:
-%
+\begin{figure}
 \begin{equation*}
 \begin{array}{l}
-p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
-\qquad\longrightarrow p=0.3,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
-p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)\\[0.2em]
-\qquad\longrightarrow p=0.8,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
-\qquad\longrightarrow p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
-p=0.3,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1); \ident{types}(\ident{string})\\[0.2em]
+r=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3)\\[0.2em]
+\qquad\longrightarrow r=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
+\qquad\longrightarrow r=0.3,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
+r=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)\\[0.2em]
+\qquad\longrightarrow r=0.8,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
+\qquad\longrightarrow r=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
+r=0.3,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1)\\[0.2em]
+\qquad\longrightarrow r=0.9,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1); \ident{types}(\ident{string})\\[0.2em]
 \end{array}
 \end{equation*}
-%
+\caption{A tree generated by an AI assistant for parsing CSV files. Each node has a rating $r$ and
+a data cleaning expression $e$ written in a small domain-specific language for cleaning CSV files.}
+\label{fig:csvtree}
+\end{figure}
+
+Given the above CSV file as an input, the AI assistant produces a tree shown in Figure~\ref{fig:csvtree}.
 At the first level, the AI assistant gives us three options. The first, most likely, one skips just 
-the first line and identifies that there are three columns (perhaps because the second line contains 
-two commas and can thus be interpreted as data row). The second skips first two lines and recognises
+the first line and identifies that there are three columns (this error is made because the second line contains 
+two commas and can thus be interpreted as data row). The second option skips first two lines and recognises
 three columns. The AI assistant might find this less likely, but it is the option we want to choose
@@
 expression $\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)$ and show a preview of 
-the cleaning result. Next, the user has a choice of two ways of inferring types -- the most 
+the cleaning result as a parsed table, but without types. Next, the user has a choice of two ways of inferring types -- the most 
 likely one (and the right one) is that the third column has a type \ident{int} (which means that
-the value {\texttt na} in the last row will be treated as missing).
+the value {\texttt na} in the last row will be treated as missing). If {\texttt na} was not 
+representing a missing value, the user could choose the second option which infers the type of the 
+last column as \ident{string}.
 
@@
 
-The datadiff project takes two tabular datasets that are in inconsistent formats, but represent
-two sets of observations about the same entity (such as broadband speed in two different years).
-It generates a list of patches that can be applied to one dataset in order to convert it to the
+\tpnote{I describe current version of datadiff first and then a possible improvement. If we turn
+this into a paper, we should have just one (preferably the latter) version. This section might
+be useful for the AIDA team though.}
+The datadiff package takes two tabular datasets that are in inconsistent formats, but represent
+two sets of observations about the same entity. An example is measurements of broadband speed in
+two different yers, stored as two tabular files with missing or additional columns, using different
+order of columns and different column names.
+
+Datadiff generates a list of patches that can be applied to one dataset in order to convert it to the
 format used by the second dataset. The patches form a domain-specific language (DSL) of the 
-data transformation suggested by datadiff. A patch $p$ can be one of the following
-(the details are not important for this document, but it provides a nice example of a DSL):
-
+data transformation suggested by datadiff. A patch $p$ can be one of the following:
+%
 \begin{equation*}
@@
 (assuming columns are indexed from $1$):
-
+%
 \[
@@
 
-The datadiff AI assistant generates the following tree based on the three inferred patches 
-(we omit the parameters of the patches):
-%
+\begin{figure}
 \begin{equation*}
 \begin{array}{l}
-p=0.9,~ e=\ident{delete}(\ldots)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{permute}(\ldots)\\[0.2em]
-\qquad\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
-\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
-\qquad\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{recode}(\ldots); \ident{permute}(\ldots)\\[0.4em]
-p=0.8,~ e=\ident{permute}(\ldots)\\[0.2em]
-\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{delete}(\ldots)\\[0.2em]
-\qquad\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
-\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
-\qquad\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{recode}(\ldots); \ident{delete}(\ldots)\\[0.4em]
+r=0.9,~ e=\ident{delete}(\ldots)\\[0.2em]
+\qquad\longrightarrow r=0.9,~ \ident{delete}(\ldots); \ident{permute}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow r=0.6,~ \ident{delete}(\ldots); \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\longrightarrow r=0.6,~ \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow r=0.9,~ \ident{delete}(\ldots); \ident{recode}(\ldots); \ident{permute}(\ldots)\\[0.4em]
+r=0.8,~ e=\ident{permute}(\ldots)\\[0.2em]
+\qquad\longrightarrow r=0.9,~ \ident{permute}(\ldots); \ident{delete}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow r=0.6,~ \ident{permute}(\ldots); \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\longrightarrow r=0.6,~ \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
+\qquad\qquad\longrightarrow r=0.9,~ \ident{permute}(\ldots); \ident{recode}(\ldots); \ident{delete}(\ldots)\\[0.4em]
 (\ldots)
@@
 \end{equation*}
+\caption{A tree generated by a basic datadiff AI assistant. The parameters of the patches are
+omitted -- they are the same as in the full list of patches shown above. }
+\label{fig:ddtree}
+\end{figure}
 
+The tree generated by the datadiff AI assistant based on the three inferred patches is shown 
+in Figure~\ref{fig:ddtree}.
 At the first level, the user can choose one of the three patches (we only show sub-trees for the first two). 
-The patches are sorted by their likelihood and so \ident{delete} is offered before \ident{permute}. In the
+The patches are sorted by their rating and so \ident{delete} is offered before \ident{permute}. In the
 second level, the user can choose one of the remaining patches. Assuming the user follows the first
@@
 first option being more likely). In the last third level, the AI assistant offers the remaining patch. 
-However, the user can stop before reaching the leaf and choose just
-$\ident{delete}(3);~ \ident{permute}(2,1)$ as the final data cleaning script. As before, Wrattler 
-can use the \emph{eval} function to give previews of the transformed datasets while the user is 
-navigating through the tree.
+The user can stop before reaching the leaf and choose just $\ident{delete}(3);~ \ident{permute}(2,1)$ 
+as the final data cleaning script, ignoring the \ident{recode} patch.
+
+This example shows why AI assistants provide a more convenient developer interface. The \ident{recode}
+patch would, in some cases, be a reasonable data cleaning recommendation, but a human analyst can
+look at the values and see that this is not the case here. (We could imagine an assistant that
+integrates semantic information into the process and recognizes that the values represent cities.
+This solves the specific problem shown here, but leaves many similar problems unsolved.) When using
+datadiff as a cleaning function in R, we would get all three patches, apply them and then (hopefully)
+notice the incorrect recoding. We could fix the result by tweaking the parameters of the function,
+e.g.~by lowering the threshold for \ident{recode} patches. When using an AI assistant, the analyst
+can review all patches instantly and skip the unsuitable ones.
+
+As before, the data wrangling expressions can be evaluated using the \emph{eval} function (to get 
+previews when writing the code in a notebook environment), but they can also be translated to 
+simple R or Python scripts for production use.
 
 \subsection{Alternative design of datadiff AI assistant}
+\label{sec:examples-datadiff2}
+
+\tpnote{Again, we should have just one version, preferably this one. Can we somehow formulate
+this as a Bayesian model -- i.e. we have some prior that is updated based on the choices that
+a user makes? If so, that would be neat.}
 
@@
 optimization up-front and then merely allows the user to choose some of the patches. This is 
-problematic, because the global optimization can be slow and it does not use the user input to 
-improve the results.
+problematic, because the global optimization can be slow (so the initial step might take too
+much time) and the result of the optimization does not improve based on the user input.
 
-For example, if the user could specify that a certain column should not be
-recoded, datadiff could run the global optimization taking this constraint into account and, say,
-match and reorder columns differently (this is currently not possible, because the current version
-runs the global optimization once up-front and then merely offers a choice of produced patches).
+For example, if the user could specify that a certain column should not be recoded, datadiff 
+could run an optimization taking this constraint into account and, say, match columns differently 
+(this is currently not possible, because the current version runs the global optimization once 
+up-front and then merely offers a choice of produced patches).
 
-An alternative datadiff AI assistant would generate expressions in the same domain-specific 
+A better datadiff AI assistant would generate expressions in the same domain-specific 
 language as the current one (list of patches $p$). However, the tree would be generated and 
@@
 the best possible patch given certain constraints, together with a choice of most likely constraints
-the user might want to add. A constraint $c$ would be as follows:
+the user might want to add. A constraint $c$ is defined as:
 %
@@
 constraints and generates a cleaning script together with a list of constraints (sorted by their
-likelihood) that the user might want to add. (If we allowed nodes with open-ended options, the
+ranking) that the user might want to add. (If we allowed nodes with open-ended options, the
 user could write an arbitrary constraint, but for now, we just assume that the AI assistant can
@@
 the revised AI assistant for datadiff would generate for the above two files. (In addition to a DSL 
-expression $e$ and likelihood $p$, we also include a set of constriants $C$ associated with each 
+expression $e$ and rangking $r$, we also include a set of constriants $C$ associated with each 
 node):
@@
 \begin{array}{l}
-C=[], p=1,~ e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
-\qquad\longrightarrow C=[\ident{norecode}(2)], p=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);\\
+C=[], r=1,~ e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{norecode}(2)], r=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);\\
 \qquad\qquad\longrightarrow C=[\ident{norecode}(2); \ident{nodelete}(3)], p=0.2, e=(\ldots)\\
-\qquad\longrightarrow C=[\ident{nodelete}(3)], p=0.2, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
-\qquad\longrightarrow C=[\ident{match}(1,1)], p=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
-\qquad\longrightarrow C=[\ident{match}(1,2)], p=0.8, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\qquad\longrightarrow (\ldots)\\
+\qquad\longrightarrow C=[\ident{nodelete}(3)], r=0.2, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{match}(1,1)], r=0.4, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow C=[\ident{match}(1,2)], r=0.8, e=\ident{delete}(\ldots);~ \ident{permute}(\ldots);~ \ident{recode}(\ldots)\\
+\qquad\longrightarrow (\ldots)
 \end{array}
@@
 
+\newpage
 The first two sub-nodes prevent datadiff from generating two patches that it employed. The first one
@@
 columns. The listing only shows the first two (map column $1$ to column $1$; map column $1$ to column $2$),
-but the AI assistant would likely generate the full range, sorted by likelihood.
+but the AI assistant can generate the full list.
 
 \section{Evaluating AI assistants}
+\label{sec:evaluation}
 
-When building an AI assistant, we want to be able to evaluate how good it is. For an AI tool that 
-is implemented as a data cleaning function, we can run it on sample inputs and asses how good 
-the resulting outputs are. (Even this is not so simple, but we might know what the optimal results 
-look like for some inputs, or we could generate inputs by introducing errors into clean datasets 
-and see if those were eliminated.)
+\tpnote{This section has only notes on ``methods for evaluating AI assistants in general'' but it 
+does not actually do any concrete evaluation of specific AI assistants using those general methods.
+We'll obviously need to add that in a real paper (once we have concrete AI assistants we want to 
+talk about).}
 
-The situation with AI assistants is more complicated as they are interactive and aim to produce
-cleaning scripts that humans can understand. We could evaluate AI assistants using user studies 
-(and this would be, no doubt, worthwhile), but there are a couple of things that we can say about
-AI assistants in a formal way too.
+Evaluating AI assistants is hard, because they are not fully automatic. They are designed to 
+assist a human data analyst in solving a data wrangling problem. We could evaluate the experience 
+of using an AI assistant on a sample task using a user study. This would be no doubt worthwhile, 
+but it would assess the user experience rather than the quality of the artifical intelligence 
+methods employed by an AI assistant. In this section, we consider how to evaluate whether an 
+AI assistant provides good data wrangling suggestions to the data analyst.
 
-\subsection{Evaluating AI assistant in an automatic way}
+The fact that all AI assistants follow a common framework -- i.e. they can be viewed as trees with
+ranked branches and data wrangling scripts in each node -- means that we can discuss evaluation 
+methods that can be used to assess the quality of any AI assistant, rather than inventing a 
+suitable evaluation method individually for each assistant. The most important aspect of such
+evaluation methods is how they account for the interactivity of AI assistants. Before discussing
+this, we first consider how can a fully automatic data wrangling function be evaluated.
+
+\subsection{Evaluating fully automatic cleaning functions}
+\label{sec:eval-auto}
+
+\tpnote{This is somewhat long methodological comment. It may be useful for the AIDA team, but
+perhaps not so much for a real paper (perhaps if it was a bit shorter).}
+We first consider how to evaluate an AI data wrangling tool that is implemented as a simple 
+function -- the data analyst calls it on input data and it produces clean data as an output.
+There are several ways in which such function can be evaluated:
+%
+\begin{itemize}
+\item \textbf{Sample benchmarks.} If we have a collection of sample inputs with clean outputs,
+  we can assess how close the outputs produced by the tool are to the clean outputs. This method
+  relies on having an extensive benchmark for each particular problem, which is rarely the case.
+\item \textbf{Corruption patches.} We can generate input/output pairs by taking 
+  a number of clean datasets (as the outputs) and corrupting them by introducing
+  annomalies which the automatic tool should then detect and remove. This evaluation method requires 
+  less manual work, but it is limited by how realistict the corruption patches are (i.e.~we only
+  test the tool on problems that we can easily imagine).
+\item \textbf{Statistical analysis.} In some cases, there might be a way to generate inputs 
+  semi-randomly, run the cleaning tool and then use some statistical property of the input/output
+  pair to assess the tool. This can be done when there is a statistical property that is easy
+  to check, but finding the transformation is hard.
+\end{itemize}
+
+\subsection{Evaluating interactive AI assistant}
+
+Evaluating AI assistants is more complicated than evaluating automatic functions.
+AI assistants do not just produce one output -- they organise possible outputs in a tree with 
+ranked branches and we should evaluate how good the tree is overall. The following three 
+sections consider different aspects of the tree that could be evaluated.
+
+\subsubsection{Evaluating AI assistant as an automatic tool}
+
 First of all, we can treat an AI assistant as a fully automatic tool. To do this, we automatically
-choose the most likely option provided by the AI assistant until we reach a leaf node (or, perhaps,
-until we reach a case where the sub-node has a likelihood $p\leq 0.5$).
+choose the most likely option provided by the AI assistant until we reach a leaf node or
+until we reach a node where all the branches have ranking below a certain threshold (e.g.~$r\leq 0.5$).
 
-More formally, given a dataset $d$ and an AI assistant $a$ (as defined on page 2), we run the 
-AI assistant $a(d)$ to obtain a list of options $(n_1, p_1, e_k, t_1), \ldots (n_k, p_k, e_k, t_k)$.
-We choose the most likely node $(n_i, p_i, e_i, t_i)$ (i.e. $p_i \geq p_j$ for all $j$). If 
-$t_i$ is empty or the likelihoods of all options inside $t_i$ are less than $0.5$, we return
+More formally, given a dataset $d$ and an AI assistant $a$ (as defined in Section~\ref{sec:trees-def}), 
+we run the AI assistant $a(d)$ to obtain a list of options $(n_1, r_1, e_k, t_1), \ldots (n_k, r_k, e_k, t_k)$.
+We choose the most likely node $(n_i, r_i, e_i, t_i)$ (i.e. $r_i \geq r_j$ for all $j$). If 
+$t_i$ is empty or the ranking of all options inside $t_i$ are less than a given threshold, we return
 an expression $e_i$; otherwise, we recursively follow the sub-tree $t_i$. Finally, we evaluate
-the resulting data cleaning expression using $\mathit{eval}(d, e_i)$ and asses how good the 
-cleaned data set is using whatever methods we would use for an automatic cleaning function.
+the resulting data cleaning expression using $\mathit{eval}(d, e_i)$ to get the clean output for
+the input data $d$.
 
+Now that we have an input/output pair, we can evaluate how good the cleaned data set is using 
+methods for evaluating automatic cleaning functions discussed in the previous section (either 
+by running the assistant on a benchmark, or by generating a benchmark using corrpution patches).
 This way, we can guarantee that the interactive AI assistant is as good as an automatic cleaning 
@@
 
-\subsection{Evaluating interactivity for annomaly handling}
+\subsubsection{Evaluating interactivity of AI assistants}
 
@@
 \end{minipage}
-\vspace{1em}
+\vspace{0.5em}
 
 \noindent
-As humans, we can see that the column {\texttt ID} should be aligned with {\texttt Code}
-and {\texttt Count} should likely match {\texttt Value}. However, unless the AI assistant
-considers possible meanings of the file headers, it has no way of discovering this.
+As humans, we can look at the column names and guess that the column {\texttt ID} should be 
+aligned with {\texttt Code} and {\texttt Count} should likely match {\texttt Value}. However, 
+unless the AI assistant ``understands English'', it has no way of discovering this.
 
-We might call such annomalies \emph{adversary annomaly}. There is very likely no way of 
-characterising adversary annomalies in general for all AI assistants, but it should be possible
-to characterise them for concrete data wrangling problems. Examples might include rows that
-are not outliers in the statistical sense, textual values that have the right format, but wrong 
-meaning such as ``Refugee Olympic Athletes'' among the names of countries\footnote{This is a name
-for an actual team of 10 athletes competing in the 2016 Rio Olympic games.}, or a suspicious 
-numerical value such as $123.45678$ in a numerical column with values 
-$190.52136, 58.68297, 123.45678, 67.47067, 47.69846$.
+\paragraph{Hidden anomalies.}
+We say that a \emph{hidden anomaly} is a value in the data set that has statistical properties
+compatible with the rest of the values, but does not fit among the other values in some other
+sense -- which typically relies on human understanding of the value or on additional semantic
+information. The definition is likely always going to depend on a particular AI assistant -- we
+can continue improving AI assistant to detect more and more hidden anomalies, but there will always
+be cases that require human insigth. That said, there are some typical examples of hidden anomalies.
 
-An important quality of an AI assistant is tht it makes it possible to eliminate adversary
-annomalies with a small number of interactions, ideally just by choosing an appropriate 
-non-default offered option at one point when traversing the tree. For example, given the
-above example, the alternative datadiff AI assistant would allow us to choose a tree node that
-adds a constraint $\ident{match}(2,3)$ and we would get the correct result with just one manual
-intervention (the other column would be reordered automatically).
+Examples of hidden anomalies might include rows that are not outliers in the statistical sense
+(such as the row in a CSV file containing copyright information, but having correct number of commas), 
+textual values that have the right format, but wrong meaning (such as ``Refugee Olympic Athletes'' 
+among the names of countries\footnote{This is a name for an actual team of 10 athletes competing 
+in the 2016 Rio Olympic games.}), or a suspicious numerical value such as $123.45678$ in a numerical 
+column with values $190.52136, 58.68297, 123.45678, 67.47067, 47.69846$.
+
+Following the idea of ``corruption patches'' introduced when discussing the evaluation of automatic
+data wrangling tools (Section~\ref{sec:eval-auto}), it might be possible to produce a set of common
+hidden anomalies and write scripts that will corrupt a clean data set and also introduce hidden 
+anomalies during the process. However, it is not clear whether this can be done in general, or
+whether this needs to be done case-by-case for each particular AI assistant.
+
+\paragraph{Eliminating hidden anomalies interactively.}
+Hidden anomalies are problems in data that cannot be detected and removed by a fully automatic
+data wrangling tool (i.e.~a tool that is implemented as a function). An important quality of AI 
+assistants is that they should make it possible to eliminate hidden anomalies with a small 
+number of interactions, ideally just by choosing an appropriate non-default offered option at one 
+point when traversing the tree. For example, given the above example, the alternative datadiff AI 
+assistant would allow us to choose a tree node that adds a constraint $\ident{match}(2,3)$ and 
+we would get the correct result with just one manual intervention (the other column would be 
+reordered automatically).
 
@@
 corner cases that are not discovered by the machine learning algorithm behind it. To do this,
-we introduce adversary annomalies into the data -- those are plausible real-world data annomalies 
+we introduce hidden anomalies into the data -- those are plausible real-world data annomalies 
 that are not discovered by the algorithm emplyed by the AI assistant. A good AI assistant allows
@@
 
-\subsection{Easy discoverability of best results}
+\subsubsection{Easy discoverability of best results}
 
@@
 is available immediately in the root node (and users only need to continue exploring a
-sub-tree if they need to address an adversary annomaly).
+sub-tree if they need to address an hidden annomaly).
 
+\begin{figure}
+\includegraphics[scale=0.55,trim=1cm 8.5cm 11cmm 4cm,crop]{diagrams.pdf}
+\caption{Information flow in an adaptive AI assistant based on collecting constraints.}
+\label{fig:aiflow}
+\end{figure}
+
+\section{Further work and open questions}
+The formal model and examples presented in this document are very basic, but they capture the
+main idea of ``what is an AI assistant''. We focused on the simplest possible model of 
+interaction -- choosing one from several options, modelled as selecting one branch from a tree.
+This has obvious limitations. It does not allow the data analyst to give free form feedback 
+to the AI assistant. However, this model can be nicely integrated with developer tools (via 
+auto-complete) and still captures a wide range of useful AI assistants. Considering more general
+models would be an interesting follow-up work, but this paper focuses on taking the first step.
+
+\tpnote{For a proper paper, I think we need to write more about the constraints/Bayesian model,
+but we can leave composing of AI assistants as future work.}
+
+In this section, we discuss two other directions for future work. First, we consider one way in
+which many AI assistants could be built. Second, we consider ways of composing multiple AI 
+assistants.
+
+\subsection{Adaptive AI assistants}
+One pattern that seems particularly useful when implementing AI assistants is using constraints
+as discussed in the second version of datadiff assistant (Section~\ref{sec:examples-datadiff2}).
+In this case, we implement a function that takes a set of constraints and produces the best 
+possible data wrangling script for given data under given constraints. The function also needs
+to provide possible constraints that could reasonably be added to the set. Given such function,
+we can build an AI assistant that creates a tree -- nodes correspond to sets of constraints and
+branches correspond to adding constraints to the set.
+
+The idea of an AI assistant that is generated based on constraints is illustrated in 
+Figure~\ref{fig:aiflow}. The AI assistant is invoked on the input data with an initial set of
+constraints (typically empty). It produces a data wrangling script (best recommendation under
+the given constraints) together with a set of constraints that it suggests to the data anlyst.
+The analyst either accepts the script, or chooses one of the constraints and adds it to the
+list of constraints for the AI assistant. The assistant then makes another suggestion and the
+loop continues.
+
+Not all AI assistants can be implemented in this way, but it is a useful common pattern that is
+worth developing in more detail. There are a number of open questions.
+
+\paragraph{Bayesian interpretation.}
+The idea of having initial set of constraints and then adapting the result of a machine learning
+algorithm based on input from the user sounds closely related to Bayesian machine learning. In 
+that case, we start with some prior assumptions about data. We can then run a machine learning 
+algorithm to make predictions about the data based on those assumptions. However, we then adapt
+the model based on new observations.
+
+\tpnote{I'm writing this, but I don't know what I'm saying! I just hope that someone from the AIDA 
+team can correct this and write something more interesting :-)}
+
+In case of AI assistants, the prior would be either some automatically generated ``best guess''
+and the new observations used to adapt the model would be the inputs (or choices) specified by 
+the data analyst. The first question is, what is the relationship of this idea to the approach
+based on constraints -- are these two just two different ways of saying the same thing, is one
+of the methods more general than the other, or are they just two unrelated useful patterns?
+
+\tpnote{These are just random thoughts -- but I think we can only answer this if we actually 
+try building some example concrete AI assistant based on the Bayesian idea -- and then we'll
+see how it actually needs to work. It would be great if we did this!}
+The second question is what is needed in order to better support the Bayesian model of AI assistants.
+Would the data analyst need to specify some initial input to the AI assistant? And what are the
+ways in which the additional data might be specified? In our current model, the AI assistant has
+to provide a range of options to choose from. Is this flexible enough for an AI assistant based
+on a Bayesian model, or do we need a more flexible form of interaction, for example one where the
+data analyst can input some data manually as examples?
+
+\paragraph{Specifying initial parameters.} 
+In the case of datadiff, we started with an empty set of constraints. However, in some cases, it
+might be desirable to provide more input to the AI assistant before using it. Such input could 
+include an initial set of constraints (in a format akin to that used by the revised version of 
+datadiff) or information about a Bayesian prior. For example, the user might want to specify 
+assumptions about relationships between columns in a CSV file (in order to help an AI assistant
+that attempts to fill missing values based on values from related columns).
+\tpnote{The idea about column relationships might be useful for the work Alfredo is doing.}
+
+Yet another interesting example would be AI assistants that take a probabilistic model as an 
+input. This might be a simple description of a model or a more complex one such as a function
+written in a probabilistic programming language. Such AI assistant could then interactively
+work with the user on either refining the program, or inferring best parameters. Alternatively,
+the AI assistant could be used as an interactive mechanism for constructing a probabilistic 
+program -- it might start with an empty program and then provide a way for adding different 
+program logic to it.
+\tpnote{Again, this is just wild speculation -- we'd need a concrete example to say more about
+this, but I quite like the idea...}
+
+\subsection{Composing multiple assistants}
+
+As discussed earlier, AI assistants are designed to solve one particular data wrangling task and
+we expect that a data analyst will typically need to interact with several AI assistants during 
+the data wrangling process. In this document, we assume that this will be done manually and that
+the AI assistants will be invoked manually, one after another, to solve specific tasks. This 
+might be sufficient if the tasks are distinct (e.g.~scraping data from the web, merging data
+sets, identifying outliers, etc.), but it is also worth considering how to best integrate multiple 
+AI assistants.
+
+\tpnote{This is all fairly speculative, but I wanted to record all ideas that we discussed!}
+
+\paragraph{Combining assistants.}
+When the user loads a messy data set, they might not initially know how to get started cleaning
+it. They have a variety of AI assistants available, but it might still be difficult to know
+where to start. One way of integrating AI assistants that would be useful in this scenario is
+creating a ``meta-AI assistant'' that advises the user what AI assistants to apply and in which
+order. This could be, perhaps, done by looking at the rankings of choices reported by the AI
+assistants and finding the ones that are the most certain about their results.
+
+\paragraph{Sharing information between assistants.}
+
+It might be also desirable to have a more fine grained way of integrating AI assistants. For
+example, say we have two separate AI assistants; one for parsing CSV files and an one for inferring
+types from data. Ideally, we would be able to interleave the uses of these two, so that the CSV
+assistant finds data in the text file and the type inferrence assistant then infers the type
+of the data. The information about the quality of the inferred types could then be fed back to the
+CSV assistant (which might use it as a measure of how well it identified the data in the file).
+
+This level of interaction is not currently possible, because there is no way of exchanging 
+information between AI assistants. There are two possible directions we could investigate. 
+First, it might be possible for multiple adaptive assistants based on constraints to share the 
+wrangling domain specific langauge and constraints language. We could then invoke multiple
+assistants in the data cleaning loop as shown in Figure~\ref{fig:aiflow}.
+
+Alternatively, the AI assistants could communicate by attaching more semantic information to 
+the data (or, generating scripts that attach semantic information to the data). If the input 
+data was not just a data frame with raw values, but a table with additional semantic information,
+one assistant could recover some semantic information and another assistant could then use it.
+For example, we might run type inferrence assistant first (to annotate data with likely types).
+If we then invoked datadiff, it could use these inferred types to better match columns from 
+multiple data files.
 \end{document}
+
+%
